### 2024-08-08 manuscript/paper.qmd
Add manuscript files
Not currently part of the pipeline

---
title: Generating Correlated Data for Omics Simulation
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  cache: true
format:
  html:
    toc: true
    toc-expand: true
bibliography: references.bib
---

## Abstract

Simulation of realistic omics data is a key input for benchmarking studies that help users obtain optimal computational pipelines.
Omics data involves large numbers of measured features on each samples and these measures are generally correlated with each other.
However, simulation too often ignores these correlations, perhaps due to the inconvenience and computational hurdles of doing so.
We describe two approaches to solving this problem that are simple to implement and computationally very fast, both of which work by decomposing the covariance matrix into a diagonal part and a low-rank part.
We demonstrate the importance of including correlation in two benchmarking applications.
First, we show that variance of results from the popular DESeq2 method increases when dependence is included.
Second, we demonstrate that a method, CYCLOPS, for inferring circadian time of collection from transcriptomics actually improves its performance when dependence is included.
We provide an R package, dependentsimr, that has efficient implementations of these methods and can generate dependent data with arbitrary distributions, including discrete (binary, ordered categorical, Poisson, negative binomial), continuous (normal), or with an empirical distribution.

## Introduction

Omics data is in the “p \>\> n” regime where there are fewer samples than measurements per sample.
This creates dual challenges in generating realistic simulated data for the purposes of benchmarking.
First, there isn’t enough data to be able to compute a dependence structure (e.g., a full-rank correlation matrix).
Second, generating omics-scale data with a specified correlation matrix is slow due to the typical $O(p^3)$ nature of these algorithms.
These often mean that simulators assume independence of the measurements, which does not reflect reality.

Recently, we wrote guidelines for performing omics benchmarking [@Brooks2024] and encouraged authors to ensure that any simulated data includes realistic correlation and dependence of measurements within a sample.
However, this highlighted that one reason this is often neglected is that there is relatively little guidance for how to achieve it and that existing solutions are cumbersome.

Here, we give a simple solution this problem by using a correlation matrix that decomposes into a diagonal part and a low-rank part to both approximate realistic dependencies in a real data set and generate simulated data mimicking the real data.
Using a NORTA (Normal to Anything) approach, the marginal (univariate) distributions can have realistic forms.
We implement this strategy as an R package which supports normal, Poisson, DESeq2-based (negative binomial with sample-specific size factors), and empirical (for ordinal data) marginal distributions.
This makes it particularly suited to RNA-seq data but also widely applicable.

We show two applications which demonstrate the importance of including dependence of measurements in simulated data when benchmarking computational pipelines.
First, we simulate RNA-seq data with differential expression between two conditions.
Using DESeq2 to determine the differentially expressed genes, we found that dependence had little impact on the accuracy of reported p-values but increased the variance of those estimates, meaning that experiments simulated with dependence were more likely to be conservative and more likely to be anti-conservative compared to simulations of independent data.
Second, we simulated a time series of RNA-seq data points and used the CYCLOPS method to infer collection time from the RNA-seq data, without time labels.
Surprisingly, we found that including dependence in the data set resulted in improved performance with CYCLOPS.

## Results

### Comparison to Real Data

```{r}
#| include: false
source("compare_to_real.R")
```

We used mouse liver RNA-seq data set from accession GSE77221[@Weger2019] where we expected a significant amount of gene-gene correlation due to the samples having been collected throughout the course of a day.
We used this data to prime our simulation and compared real and simulated data sets, both with 12 samples, to each other.
We repeated the simulations a total of 8 times to estimate variance.
The simulations were then repeated without any dependence for comparison.

Simulated data captures the genes' mean and variance accurately (@fig-compare-to-real a-b).
Moreover, we computed the gene-gene correlation on a random subset of `r N_rows_gene_corr` high-expressed genes.
Since only 12 samples are in each real or simulated data set, correlation estimates are very noisy.
However, the distribution of correlations across all pairs of these genes is much closer to that of real data when simulated with dependence than without (@fig-compare-to-real c).

Next, we compared the simulations with and without dependence to the real data set when projected onto the top two principal components of the real data set (@fig-compare-to-real d).
The simulations with dependence are distributed around the entire space like the real data, but the independent simulations have unrealistically low variance in these components, clustering tightly around the origin.
Lastly, we compared the PCA of each simulated data set considered separately from the real data.
Simulations with independent data showed unrealistically low levels of variance in top principal components, but dependent simulations actually show too high elevations of variance in top components.
This is a consequence of our PCA strategy: since variance in the direction of the real data set's top principal component is exactly matched, the simulated data set's top principal component will be even larger than that.

```{r}
#| label: fig-compare-to-real
#| fig-cap: !expr glue::glue("Comparison to real data run on a mouse liver example from GSE77221. (a-b) Comparison of gene (a) mean expression and (b) variance, log-scaled. The line of equality is marked in black. Points are colored according to the density of points in their region. (c) Quantile-quantile plot comparing correlation values of gene pairs from real data and simulated data (both with and without dependence). A random sample of {N_rows_gene_corr} genes with at least {HIGH_EXPR_CUTOFF} reads was used. Values on the diagonal line indicate a match between the simulated and real data sets. (d) Projections onto the top two principal components of the real data set for both real and simulated data. All 8 simulations (96 samples total for each of dependent and independent simulations) shown. (e) Principal component analysis was performed on all data sets and the variance captured by the top components is shown. Unlike (d), these components were fit from each data set considered separately instead of reusing the weights from the real data.")
#| fig-width: 10
#| fig-height: 10
compare_to_real_plot
```

### DESeq2 Application

TODO Show performance of DESeq2 both with and without dependence and a figure

### CYCLOPS Application

TODO Show performance of CYCLOPS with and without dependence and a figure

## Methods

TODO: how it works

## Discussion

We present `dependentsimr`, an R package that generates omics-scale data with realistic correlation.
TODO: limitations TODO: alternatives

### 2024-08-08 manuscript/compare_to_real.R

@@ -1,0 +1,158 @@
+
+# Comparison of simulated data to a real dataset
+
+library(dependentsimr)
+library(ggplot2)
+library(ggpointdensity)
+library(tidyr)
+library(dplyr)
+library(patchwork)
+
+set.seed(55)
+
+# CONFIGURATION
+N_rows_gene_corr <- 3000
+HIGH_EXPR_CUTOFF <- 300
+
+## Generate the simulated data
+# Load the real data ---------------------------------
+# from GEO: GSE77221
+head(Weger18)
+weger18_library_sizes <- apply(Weger18, 2, sum)
+
+# Run dependentsimr on this data --------------------
+N_SAMPLES <- 96
+library_sizes <- (weger18_library_sizes + rep(0, N_SAMPLES)) / mean(weger18_library_sizes) # Recycle the existing library sizes
+# Simulate with the PCA method
+rs <- get_random_structure(list(counts=Weger18), method="pca", rank=2, type="DESeq2")
+draws <- draw_from_multivariate_corr(rs, n_samples=N_SAMPLES, size_factors=library_sizes)$counts
+
+# Simulate with the corpcor method
+rs_corpcor <- get_random_structure(list(counts=Weger18), method="corpcor", type="DESeq2")
+draws_corpcor <- draw_from_multivariate_corr(rs_corpcor, n_samples=N_SAMPLES, size_factors=library_sizes)$counts
+
+# Simulate without any dependence
+rs_indep <- remove_dependence(rs)
+indep_draws <- draw_from_multivariate_corr(rs_indep, n_samples=N_SAMPLES, size_factors=library_sizes)$counts
+
+
+# Scale to Counts Per Million ------------------------
+cpm <- function(x) { # counts per million
+  return(t(t(x) / apply(x, 2, sum) * 1e6))
+}
+scaled_read_data <- cpm(Weger18)
+scaled_draws <- cpm(draws)
+scaled_draws_corpcor <- cpm(draws_corpcor)
+scaled_indep_draws <- cpm(indep_draws)
+
+## Generate the figures
+
+simulated_mean <- apply(scaled_draws[,1:ncol(Weger18)], 1, mean)
+simulated_variance <- apply(scaled_draws[,1:ncol(Weger18)], 1, var)
+real_mean <- apply(scaled_read_data, 1, mean)
+real_variance <- apply(scaled_read_data, 1, var)
+marg_dist <- data.frame(
+  real_mean = real_mean,
[…]

### 2024-08-09 run_pipeline.sh
Add manuscript to the pipeline

@@ -2,6 +2,7 @@
 module load /project/itmatlab/sharedmodules/use.shared
 module load python/3.10
 module load R/4.3
+module load curl-7.49.0
 source venv/bin/activate
 
 

### 2024-08-09 scripts/process_data.fly.DMSO.py
Add fly data processing script (not incorporated into pipeline)

@@ -1,0 +1,35 @@
+import pathlib
+import re
+import pandas
+
+dataset = {}
+for filepath in pathlib.Path("data/GSE81142_RAW").glob("*.txt.gz"):
+    m = re.match(r"GSM[0-9]*_([a-zA-Z0-9]*).reverse.htseq.counts.txt.gz", filepath.name)
+    if m:
+        sample_name = m.groups()[0]
+    else:
+        print(f"WARNING: Skipping unexpected file name {filepath}")
+        continue
+
+    d = pandas.read_csv(filepath, sep="\t", names=["GENE_ID", "counts"])
+    d = d[~d.GENE_ID.str.startswith("__")]
+    dataset[sample_name] = d.set_index("GENE_ID").counts
+
+pandas.DataFrame(dataset).to_csv("processed/GSE81142.counts.txt.gz", sep="\t")
+
+metadata = []
+for sample in dataset.keys():
+    m = re.match(f"M([mf])([0-9])h([0-9])cR([0-9])", sample)
+    if not m:
+        print(f"ERROR: sample name not of the expected format: {sample}")
+        break
+    sex,hour,concentration,rep = m.groups()
+    metadata.append({
+        "sample_id": sample,
+        "sex": {'m': 'male', 'f': 'female'}[sex],
+        "hour": int(hour),
+        "concentration": int(concentration),
+        "rep": int(rep),
+    })
+df = pandas.DataFrame(metadata)
+df.to_csv("processed/GSE81142.metadata.txt", sep="\t", index=False)

### 2024-08-20 manuscript/paper.qmd
Update paper.qmd, move plotting scripts to manuscript, and remove unnecessary input files in simulate_time_data.R and Snakefile.

@@
 
-TODO Show performance of DESeq2 both with and without dependence and a figure
+We benchmark DESeq2, a popular differential expression analysis tool, using data sets simulated with dependence and ones simulated without dependence to compare its performances on both.
+We use a mouse cortex data set GSE151923 and select samples from male mice to simulate 20 control samples with rank $k=2$ dependence structure, and 20 control samples assuming independence of the genes.
+We then randomly select 5% of the genes to be differential expressed, and simulate 20 "experimental" samples each with and and without dependence.
+
+Finally, we run DESeq2 on the two simulated experiments and compare the output FDR with the true percentages of genes that are differential expressed (@fig-DESeq2 a-b).
+We observe that DESeq2 is slightly anti-conservative on both data sets, with similar mean true FDRs for each estimated FDR cutoff.
+However, there is a greater variance in the performance of DESeq2 on the data sets simulated with dependence, indicating that it potentially performs less reliably on data sets with gene-gene dependence, as in real data sets.
+We speculate that this is due to DESeq2's assumption that the gene expressions are independent.
+
+To demonstrate the application of our simulation method for another organism, we also simulate data sets using a fly whole body RNA-Seq data set GSE81142.
+We select samples of male flies without treatment and after at least 2 hours of feeding to infer the dependence structure of the quantified genes.
+We then simulate differential expression experiments with and without dependence as above, with 20 samples for each condition.
+We observe a similar result (@fig-DESeq2 c-d) as for the mouse cortex data sets.
+
+```{r}
+#| include: false
+source("DE_plots.R")
+```
+
+```{r}
+#| label: fig-DESeq2
+#| fig-cap: "Performance of DESeq2 on simulated datasets.
+(a-b) Comparison of true FDR and DESeq2 reported FDR for data sets simulated from the mouse cortex data set (GSE151923, Wang et al.
+2022), (a) without dependence and (b) with dependence.
+(c-d) Comparison of true FDR and DESeq2 reported FDR for data sets simulated from the fly whole body data set (GSE81142), (c) without dependence and (d) with dependence."
+#| fig-width: 7
+#| fig-height: 6
+DESeq2_plot
+```
 
@@
 
-TODO Show performance of CYCLOPS with and without dependence and a figure
+We further use our simulation method to benchmark CYCLOPS (Anafi et al.
+2017), which uses an autoencoder to identify circular structures in the data and infer relative times for a set of unlabeled samples.
+We use a mouse liver time series data set (GSE151565, Aaronson et al.
+2020), which contains 12 samples for ZT0, and 6 samples each for ZT3, ZT6, ZT9, ZT12, ZT15, ZT18 and ZT21.
+We compute the dependence structure of the genes as well as the variances of marginal distributions using the ZT0 samples, and compute the means of gene expressions at each time point.
+We then simulate 20 time series data sets with rank $2$ dependence structure, each with 8 time points and 6 samples per time point, and 20 such data sets without dependence.
+
+We run CYCLOPS on each data set with a list of cyclic mouse liver genes (JTK p-value $<0.05$), which yields an estimated relative phase for each sample.
+For the evaluation metric for CYCLOPS' performance, we use the circular correlation, defined as follows:
+$$\rho = \frac{\sum_{1\leq i<j\leq n}sin(X_i-X_j)sin(Y_i-Y_j)}{(\sum_{1\leq i<j\leq n}sin(X_i-X_j)^2)^{1/2}(\sum_{1\leq i<j\leq n}sin(Y_i-Y_j)^2)^{1/2}},$$
+where $n$ is the number of samples, $X_i$ and $Y_i$ are the true phase and CYCLOPS-estimated phase respectively for the $i$-th sample.
+$\rho$ has value between $-1$ and $1$, and a $|\rho|$ close to $1$ indicates accurate predictions by CYCLOPS.
+
+We find that CYCLOPS performs better overall on simulated time series data sets with dependence than on those without dependence (@fig-CYCLOPS), indicating that the presence of gene-gene dependence is essential for CYCLOPS' performance.
+
+```{r}
+#| include: false
+source("circular_correlation.R")
+```
+
+```{r}
+#| label: fig-CYCLOPS
+#| fig-cap: "Performance of CYCLOPS on simulated time series data sets based on mouse liver data set (GSE151565, Aaronson et al.
+2020).
+(a) Violin plot comparing absolute circular correlations between true phases and CYCLOPS estimated phases on the simulated data sets with and without dependence.
+(b-c) Examples of CYCLOPS estimated phases on a simulated data set (b) without dependence and (c) with dependence."
+cyclops_plot
+```
 

### 2024-08-20 manuscript/DE_plots.R

@@ -4,14 +4,14 @@ library(DESeq2)
 library(ggplot2)
 library(dplyr)
 library(tidyr)
+library(patchwork)
 
-fdr_df0 <- read_csv("processed/DE/Mouse.Cortex.Male.k=0.fdr.csv")
+fdr_df0 <- read_csv("~/dependent_sim/dependent_sim_paper/processed/DE/Mouse.Cortex.Male.k=0.fdr.csv")
 
 mouse_cortex_fdr_0 <- ggplot(fdr_df0, aes(cutoff, mean, ymin=lower, ymax=upper)) +
     geom_ribbon(fill="grey") +
     geom_point() +
     geom_abline(slope=1, intercept = 0) +
-    labs(title = "k=0") +
     xlab("reported FDR") +
     ylab("true FDR") +
     scale_x_log10() +
@@ -27,13 +27,12 @@ mouse_cortex_fdr_0 <- ggplot(fdr_df0, aes(cutoff, mean, ymin=lower, ymax=upper))
 #     geom_smooth(color="black", method="lm") +
 #     geom_abline(slope=1, intercept = 0)
 
-fdr_df2 <- read_csv("processed/DE/Mouse.Cortex.Male.k=2.fdr.csv")
+fdr_df2 <- read_csv("~/dependent_sim/dependent_sim_paper/processed/DE/Mouse.Cortex.Male.k=2.fdr.csv")
 
 mouse_cortex_fdr_2 <- ggplot(fdr_df2, aes(cutoff, mean, ymin=lower, ymax=upper)) +
     geom_ribbon(fill="grey") +
     geom_point() +
     geom_abline(slope=1, intercept = 0) +
-    labs(title = "k=0") +
     xlab("reported FDR") +
     ylab("true FDR") +
     scale_x_log10() +
@@ -41,13 +40,12 @@ mouse_cortex_fdr_2 <- ggplot(fdr_df2, aes(cutoff, mean, ymin=lower, ymax=upper))
     coord_cartesian(xlim = c(0.001, 1), ylim= c(0.001,1))
 
 
-fdr_df0_fly <- read_csv("processed/DE/Fly.WholeBody.Male.k=0.fdr.csv")
+fdr_df0_fly <- read_csv("~/dependent_sim/dependent_sim_paper/processed/DE/Fly.WholeBody.Male.k=0.fdr.csv")
 
[…]

### 2024-08-20 manuscript/circular_correlation.R

@@ -3,6 +3,7 @@ library(tibble)
 library(ggplot2)
 library(dplyr)
 library(tidyr)
+library(patchwork)
 
 # formula for the circular correlation
 get_cir_corr <- function(X, Y) {
@@ -34,7 +35,7 @@ get_cir_corr <- function(X, Y) {
 corrs <- tibble(k0 = rep(0, 20), k2 = rep(0, 20))
 for (j in (0:19)) {
     #k=0
-    phaselist <- read_csv(paste0("processed/cyclops/k=0/batch=", j, "/cyclops_estimated_phaselist.csv"))
+    phaselist <- read_csv(paste0("~/dependent_sim/dependent_sim_paper/processed/cyclops/k=0/batch=", j, "/cyclops_estimated_phaselist.csv"))
     for (i in (1:48)) {
         phaselist[i, 'ID'] <- strsplit(as.character(phaselist[i, 'ID']), "_")[[1]][1] |> 
             substring(3)
@@ -44,7 +45,7 @@ for (j in (0:19)) {
     corrs$k0[j+1] <- get_cir_corr(phaselist$new_ID, phaselist$phase)
     
     #k=2
-    phaselist <- read_csv(paste0("processed/cyclops/k=2/batch=", j, "/cyclops_estimated_phaselist.csv"))
+    phaselist <- read_csv(paste0("~/dependent_sim/dependent_sim_paper/processed/cyclops/k=2/batch=", j, "/cyclops_estimated_phaselist.csv"))
     for (i in (1:48)) {
         phaselist[i, 'ID'] <- strsplit(as.character(phaselist[i, 'ID']), "_")[[1]][1] |> 
             substring(3)
@@ -62,7 +63,7 @@ circular_correlation_violin <- ggplot(corrs_new, aes(k, abs(value))) +
 circular_correlation_violin
 
 # rainbow plots visualizing true and estimated phases
-phaselist <- read_csv(paste0("processed/cyclops/k=0/batch=", 5, "/cyclops_estimated_phaselist.csv"))
+phaselist <- read_csv(paste0("~/dependent_sim/dependent_sim_paper/processed/cyclops/k=0/batch=", 5, "/cyclops_estimated_phaselist.csv"))
 for (i in (1:48)) {
     phaselist[i, 'ID'] <- strsplit(as.character(phaselist[i, 'ID']), "_")[[1]][1] |> 
         substring(3)
[…]

### 2024-08-20 scripts/simulate_time_data.R

@@ -4,29 +4,6 @@ library(tibble)
 library(DESeq2)
 library(tidyverse)
 
-metadata <- read.delim("processed/GSE151923_sample_metadata.txt", sep="\t")
-read_counts <- read.delim("data/GSE151923_metaReadCount_ensembl.txt.gz", sep="\t")
-
-row.names(read_counts) <- read_counts[,1]
-read_counts <- read_counts[,6:ncol(read_counts)]
-
-# Drop NAs - the data file is missing part of the very last row, giving NAs
-read_counts <- read_counts[!apply(is.na(read_counts), 1, any),]
-
-# Drop a sample that has very low read counts
-read_depth <- apply(read_counts, 2, sum)
-READ_DEPTH_CUTOFF = 10e6
-read_counts <- read_counts[,read_depth > READ_DEPTH_CUTOFF]
-
-# Separate by sex
-sex <- vapply(colnames(read_counts),
-    function(id) {
-        return(metadata[metadata["sample_id"] == id, "sex"])
-    }, "")
-
-male_read_counts <- read_counts[,sex == 'male']
-female_read_counts <- read_counts[,sex == 'female']
-
 # use ZT0 of raw time series data to get dependence structure
 male_read_counts <- read.csv("processed/Liver_ZT0-counts.csv") |> 
     column_to_rownames(var="raw_data.EnsemblID")

### 2024-08-20 manuscript/paper.qmd
revise the paper

--- 2024-08-08 manuscript/paper.qmd
@@
 
-TODO Show performance of DESeq2 both with and without dependence and a figure
+We benchmark DESeq2, a popular differential expression analysis tool, using data sets simulated with dependence and ones simulated without dependence to compare its performances on both.
+We use a fly whole body RNA-Seq data set GSE81142 and select samples of male flies without treatment and after at least 2 hours of feeding to simulate 20 control samples with rank $k=2$ dependence structure, and 20 control samples assuming independence of the genes.
+We then randomly select 5% of the genes to be up or down regulated, with $\log_2$ fold change uniformly distributed between $0.2$ and $2.0$, and simulate 20 "experimental" samples each with and and without dependence.
+
+Finally, we run DESeq2 on the two simulated experiments and compare the output false discovery rate (FDR) with the true percentages of genes that are differential expressed (@fig-DESeq2 a-b).
+We observe that DESeq2 is slightly anti-conservative on both data sets, with similar mean true FDRs for each estimated FDR cutoff.
+However, there is a greater variance in the performance of DESeq2 on the data sets simulated with dependence, indicating that it potentially performs less reliably on data sets with gene-gene dependence, as in real data sets.
+We speculate that this is due to DESeq2's assumption that the gene expressions are independent.
+
+To demonstrate the application of our simulation method for another organism, we also simulate data sets using a mouse cortex data set GSE151923 and select samples from male mice to infer the dependence structure of the quantified genes.
+We then simulate differential expression experiments with and without dependence as above, with 20 samples for each condition.
+We observe a similar result (@fig-DESeq2 c-d) as for the fly whole body data sets.
+
+```{r}
+#| include: false
+source("DE_plots.R")
+```
+
+```{r}
+#| label: fig-DESeq2
+#| fig-cap: "Performance of DESeq2 on simulated datasets.
+(a-b) Comparison of true FDR and DESeq2 reported FDR for data sets simulated from the fly whole body data set (GSE81142), (a) without dependence and (b) with dependence.
+(c-d) Comparison of true FDR and DESeq2 reported FDR for data sets simulated from the mouse cortex data set (GSE151923, Wang et al.
+2022), (c) without dependence and (d) with dependence."
+#| fig-width: 7
+#| fig-height: 6
+DESeq2_fdr_plot
+```
 
@@
 
-TODO Show performance of CYCLOPS with and without dependence and a figure
+We further use our simulation method to benchmark CYCLOPS (Anafi et al.
+2017), which infer relative times for a set of unlabeled samples using an autoencoder to identify circular structures.
+We use a mouse liver time series data set (GSE151565, Aaronson et al.
+2020), which contains a total of 77 samples every 3 hours, for 36 hours.
+We compute the dependence structure of the genes as well as the variances of marginal distributions using the 12 time point 0 samples, and compute the means of gene expressions at each time point.
+We then use these to simulate 20 time series data sets with rank $2$ dependence structure, each with 8 time points and 6 samples per time point, and 20 such data sets without dependence.
+
+We run CYCLOPS on each data set with a list of cyclic mouse liver genes (JTK p-value $<0.05$), which yields an estimated relative time for each sample.
+For the evaluation metric for CYCLOPS' performance, we use the circular correlation, defined as follows:
+$$\rho = \frac{\sum_{1\leq i<j\leq n}sin(X_i-X_j)sin(Y_i-Y_j)}{(\sum_{1\leq i<j\leq n}sin(X_i-X_j)^2)^{1/2}(\sum_{1\leq i<j\leq n}sin(Y_i-Y_j)^2)^{1/2}},$$
+where $n$ is the number of samples, $X_i$ and $Y_i$ are the true time and CYCLOPS-estimated time respectively for the $i$-th sample.
+$\rho$ has value between $-1$ and $1$, and a $|\rho|$ close to $1$ indicates accurate predictions by CYCLOPS.
+
+We find that CYCLOPS performs better overall on simulated time series data sets with dependence than on those without dependence (@fig-CYCLOPS), indicating that the presence of gene-gene dependence is essential for CYCLOPS' performance.
+
+```{r}
+#| include: false
+source("circular_correlation.R")
+```
+
+```{r}
+#| label: fig-CYCLOPS
+#| fig-cap: "Performance of CYCLOPS on simulated time series data sets based on mouse liver data set (GSE151565, Aaronson et al.
+2020).
+(a) Violin plot comparing absolute circular correlations between true phases and CYCLOPS estimated phases on the simulated data sets with and without dependence.
+(b-c) Examples of CYCLOPS estimated phases on a simulated data set (b) without dependence and (c) with dependence."
+cyclops_plot
+```
 

### 2024-08-20 manuscript/DE_plots.R

@@ -65,5 +65,5 @@ fly_fdr_2 <- ggplot(fdr_df2_fly, aes(cutoff, mean, ymin=lower, ymax=upper)) +
     scale_y_log10() +
     coord_cartesian(xlim = c(0.001, 1), ylim= c(0.001,1))
 
-DESeq2_plot <- (mouse_cortex_fdr_0 | mouse_cortex_fdr_2) / (fly_fdr_0 | fly_fdr_2) + plot_annotation(tag_levels="a")
-DESeq2_plot
+DESeq2_fdr_plot <- (fly_fdr_0 | fly_fdr_2) / (mouse_cortex_fdr_0 | mouse_cortex_fdr_2) + plot_annotation(tag_levels="a")
+DESeq2_fdr_plot

### 2024-09-19 manuscript/paper.qmd
Update for wishart method

@@
 Simulated data captures the genes' mean and variance accurately (@fig-compare-to-real a-b).
-Moreover, we computed the gene-gene correlation on a random subset of `r N_rows_gene_corr` high-expressed genes.
+Moreover, we computed the gene-gene correlation on pairs of high-expressed genes.
 Since only 12 samples are in each real or simulated data set, correlation estimates are very noisy.
@@
 (c) Quantile-quantile plot comparing correlation values of gene pairs from real data and simulated data (both with and without dependence).
-A random sample of {N_rows_gene_corr} genes with at least {HIGH_EXPR_CUTOFF} reads was used.
+Genes with at least {HIGH_EXPR_CUTOFF} reads were used.
 Values on the diagonal line indicate a match between the simulated and real data sets.

### 2024-09-19 manuscript/compare_to_real.R

@@ -1,4 +1,3 @@
-
 # Comparison of simulated data to a real dataset
 
 library(dependentsimr)
@@ -6,30 +5,42 @@ library(ggplot2)
 library(ggpointdensity)
 library(tidyr)
 library(dplyr)
+library(readr)
 library(patchwork)
+library(rlang)
 
 set.seed(55)
 
 # CONFIGURATION
-N_rows_gene_corr <- 3000
 HIGH_EXPR_CUTOFF <- 300
+TYPE_ORDER <- c("real", "indep", "pca", "wishart", "corpcor")
+TYPE_COLORS <- RColorBrewer::brewer.pal(length(TYPE_ORDER), "Set2")
 
 ## Generate the simulated data
 # Load the real data ---------------------------------
 # from GEO: GSE77221
-head(Weger18)
-weger18_library_sizes <- apply(Weger18, 2, sum)
+raw <- read_tsv("../data/GSE151923_metaReadCount_ensembl.txt.gz") |>
+  select(-GeneName, -Description, -Chromosome, -Strand)
+read_counts <- as.matrix(raw[,2:13])
+rownames(read_counts) <- raw$GeneID
+
+#head(Weger18)
+actual_library_sizes <- apply(read_counts, 2, sum)
 
 # Run dependentsimr on this data --------------------
 N_SAMPLES <- 96
-library_sizes <- (weger18_library_sizes + rep(0, N_SAMPLES)) / mean(weger18_library_sizes) # Recycle the existing library sizes
+library_sizes <- (actual_library_sizes + rep(0, N_SAMPLES)) / mean(actual_library_sizes) # Recycle the existing library sizes
 # Simulate with the PCA method
-rs <- get_random_structure(list(counts=Weger18), method="pca", rank=2, type="DESeq2")
+rs <- get_random_structure(list(counts=read_counts), method="pca", rank=2, type="DESeq2")
 draws <- draw_from_multivariate_corr(rs, n_samples=N_SAMPLES, size_factors=library_sizes)$counts
 
 # Simulate with the corpcor method
-rs_corpcor <- get_random_structure(list(counts=Weger18), method="corpcor", type="DESeq2")
+rs_corpcor <- get_random_structure(list(counts=read_counts), method="corpcor", type="DESeq2")
 draws_corpcor <- draw_from_multivariate_corr(rs_corpcor, n_samples=N_SAMPLES, size_factors=library_sizes)$counts
+
[…]

### 2024-09-19 scripts/simDE.R

@@ -4,6 +4,7 @@ library(DESeq2)
 library(ggplot2)
 library(dplyr)
 library(tidyr)
+library(rlang)
 
 simDE_analysis <- function(filename, samples) {
 

### 2024-09-19 scripts/simulate_data.R

@@ -2,6 +2,7 @@ library(dependentsimr)
 library(dplyr)
 library(tibble)
 library(DESeq2)
+library(rlang)
 
 metadata <- read.delim("processed/GSE151923_sample_metadata.txt", sep="\t")
 read_counts <- read.delim("data/GSE151923_metaReadCount_ensembl.txt.gz", sep="\t")

### 2024-09-19 scripts/simulate_data.fly.R

@@ -3,6 +3,7 @@ library(tidyr)
 library(dplyr)
 library(tibble)
 library(DESeq2)
+library(rlang)
 
 read_counts <- read.delim("data/GSE81142.counts.txt.gz", row.names=1)
 metadata <- read.delim("processed/GSE81142_sample_metadata.txt")

### 2024-09-19 scripts/simulate_time_data.R

@@ -3,6 +3,7 @@ library(dplyr)
 library(tibble)
 library(DESeq2)
 library(tidyverse)
+library(rlang)
 
 # use ZT0 of raw time series data to get dependence structure
 male_read_counts <- read.csv("processed/Liver_ZT0-counts.csv") |> 

### 2024-09-23 manuscript/paper.qmd
Add methods and discussion

@@
 However, simulation too often ignores these correlations, perhaps due to the inconvenience and computational hurdles of doing so.
-We describe two approaches to solving this problem that are simple to implement and computationally very fast, both of which work by decomposing the covariance matrix into a diagonal part and a low-rank part.
+We describe and implement three approaches based off decomposing the covariance matrix into a diagonal part and a low-rank part and using a Gaussian copula approach.
+These approaches are efficient at omics-scale problems.
 We demonstrate the importance of including correlation in two benchmarking applications.
@@
 
-Omics data is in the “p \>\> n” regime where there are fewer samples than measurements per sample.
+Omics data is in the $p \gg n$ regime where there are fewer samples than measurements per sample.
 This creates dual challenges in generating realistic simulated data for the purposes of benchmarking.
@@
 
-Here, we give a simple solution this problem by using a correlation matrix that decomposes into a diagonal part and a low-rank part to both approximate realistic dependencies in a real data set and generate simulated data mimicking the real data.
-Using a NORTA (Normal to Anything) approach, the marginal (univariate) distributions can have realistic forms.
+Here, we give solutions to this problem by using a correlation matrix that decomposes into a diagonal part and a low-rank part to both approximate realistic dependencies in a real data set and generate simulated data mimicking the real data.
+Using a Gaussian copula approach (also referred to as NORTA, for "normal to anything"), the marginal (univariate) distributions can have realistic forms.
+These solutions operate by taking a real dataset and mimicking it.
 We implement this strategy as an R package which supports normal, Poisson, DESeq2-based (negative binomial with sample-specific size factors), and empirical (for ordinal data) marginal distributions.
 This makes it particularly suited to RNA-seq data but also widely applicable.
+
+We implement three different strategies for determining the diagonal and low-rank parts of the covariance matrix.
+First, the 'PCA' method uses principal component analysis (PCA) and pick the low-rank part such that the simulated data has the same variance in the top $k$ PCA components of the reference dataset.
+Second, the 'spiked Wishart' method fits $k$ components such that simulations with the same number of samples as the reference dataset will have, on average, the same PCA component variances as the reference
+Unlike 'PCA', these variances are computed with resepect to the simulated data's own PCA and not using the PCA weights of the reference dataset.
+Third, the 'corpcor' method uses the popular R library `corpcor`, which implements a James-Stein type shrinkage estimator for the covariance matrix as a linear interpolation of the sample covariance matrix and a diagonal matrix.
+No method exactly captures the input data, indicating room for future research, but all improve upon a the common approach of assuming independence.
 
@@
 
+<!-- TODO which dataset do we want to use here? -->
 We used mouse liver RNA-seq data set from accession GSE77221[@Weger2019] where we expected a significant amount of gene-gene correlation due to the samples having been collected throughout the course of a day.
@@
 
-We benchmark DESeq2, a popular differential expression analysis tool, using data sets simulated with dependence and ones simulated without dependence to compare its performances on both.
+We benchmark DESeq2 <!-- TODO cite -->, a popular differential expression analysis tool, using data sets simulated with dependence and ones simulated without dependence to compare its performances on both.
 We use a fly whole body RNA-Seq data set GSE81142 and select samples of male flies without treatment and after at least 2 hours of feeding to simulate 20 control samples with rank $k=2$ dependence structure, and 20 control samples assuming independence of the genes.
@@
 
-We further use our simulation method to benchmark CYCLOPS (Anafi et al.
-2017), which infer relative times for a set of unlabeled samples using an autoencoder to identify circular structures.
+We further use our simulation method to benchmark CYCLOPS <!-- TODO cite -->, which infer relative times for a set of unlabeled samples using an autoencoder to identify circular structures.
 We use a mouse liver time series data set (GSE151565, Aaronson et al.
@@
 
-TODO: how it works
+Assume that we have a reference dataset $X$ given by an $p \times n$ data matrix of $p$ features measured in each of $n$ independent samples.
+We want to capture correlations between the $p$ features, which could represent gene expressions, protein abundances, or other measured values.
+We refer to these features as genes for simplicity.
+Our goal is to generate simulated data with the same $p$ genes, the same marginal distributions of each gene as in $X$ and realistic gene-gene dependence.
+
+### Gaussian Copula
+
+All three methods are based off a well-known Gaussian copula approach.
+This operates first by normalizing each gene by fitting a distribution (such as a normal distribution, Poisson, negative binomial, or other form), and then applying the fit cumulative distribution function (CDF) to the observed values.
+Finally, those are fed to a standard normal distribution's inverse CDF to obtain values that are approximately normally distributed.
+These values are then used to compute a covariance matrix $\Sigma$ and the data is assumed to follow a multivariate normal distribution in $p$ dimensions with that covariance matrix.
+Note that the sample covariance matrix is a poor choice for this $\Sigma$.
+In general, the sample covariance matrix is at most rank $n$, and will be approximately rank $n-1$ due to the normalization procedure.
+This means that generated data would lie on an $n-1$ dimensional hyperplane, which we know is unrealistic.
+
+Instead, we want to estimate some other matrix of the form $\Sigma = D + PP^T$ where $D$ is $p \times p$ diagonal and $P$ is $p \times k$ for $k \ll p$.
+This means that $P P^T$ is low-rank.
+Conveniently, this form allows efficient simulation of multivariate normal data $N(0, \Sigma)$.
+Once data is obtained $Z ~ N(0, \Sigma)$, then undo the normalization process to obtain data with the same marginal distributions as the fit marginal distributions but with dependence determined by $\Sigma$.
+We describe this in detail below:
+
+1.  Fit marginal distributions to each feature in $X$ to determine CDFs $F_{i}$ for each feature.
+2.  Transform $X$ to normalized values by $Z_{ij} = \Phi^{-1}(F_{i}(X_{ij}))$ where $\Phi$ is the CDF of the standard normal distribution.
+3.  Compute $D$, $U$, $W$ matrices from $X$ by one of three methods (below).
+4.  Generate $k$ i.i.d. standard normally distributed values $u$ and $p$ i.i.d standard normally distributed values $v$.
+5.  Set $Z' = UWu + D v$.
+6.  Output the vector $X'$ where $X'_i = F_i^{-1}(\Phi(Z'))$.
+
+The generated data $Z'$ has covaraince matrix $\Sigma = D + U W U^T$.
+Moreover, we require that $\Sigma$ satisfies that $e_i \Sigma e_i^T$ is approximately 1.
+That guarantees that the output $X'$ has each entry with the same mariginal distributions $F_i$ as was originally fit and inherits gene-gene dependence from $Z'$.
+
+This is a standard Gaussian copula approach tailored to our specific format of $\Sigma$.
+Most implementations of this approach instead rely upon performing a Cholesky decomposition of $\Sigma$, which requires significant time and memory resources for omics-scale problems with over tens of thousands of genes.
+
+###  PCA Method
+
+The first of our three methods attempts to match the top $k$ PCA components of the normalized reference dataset $Z$.
+Specifically, let $u_1, \ldots, u_k$ be the left singular vectors of $Z$ with $\lambda_1, \ldots, \lambda_k$ the corresponding top $k$ signular values.
+This method computes $\Sigma$ such that $u_i^T \Sigma u_i = \lambda_i^2$, i.e. that the variance in the direction of $u_i$ exactly matches of the reference dataset's variance in that same direction.
+One solution is to use the sample covariance matrix, but that is not full rank and would match for all $i \leq n$ instead of just $i \leq k$.
+Instead, we use the following:
+
+1.
+Compute $A_{ij} = \delta_{ij} - \sum_\ell U_{\ell,i}^2 U_{\ell,j}^2$ and $B_{i} = \lambda_i^2/(n-1)^2 - \sum_\ell U_{\ell,i}^2 V_{\ell\ell}$ where $\delta_{ij}$ is the Kronecker delta and $V = Z^T Z / (n-1)$ is the covariance matrix of $Z$.
+2. Solve $A w = B$ and set $W$ to be the diagonal matrix with $w$ along its diagonal.
+3. Set $U$ to be the $p \times k$ matrix with columns $u_i$.
+4. Set $D$ to be the diagonal matrix with $D_{ii} = \sqrt{V_{ii} - (UW^2U^T)_{ii}}$, which is the remaining variance.
+
+Step 4 ensures that $e_i \Sigma e_i^T = 1$.
+
+### Spiked Wishart Method
+
+The second method also makes use of PCA but has a different objective.
+If $n$ samples are drawn from $N(0, \Sigma)$ then we want the variances of their PCA components to match those of the reference dataset.
+Specifically, let $\lambda_1, \ldots, \lambda_{n-1}$ are the ${n-1}$ non-zero singular values of $Z$ ($\lambda_n$ is always approximately zero due to the normalization procedure), and let $\lambda'_1, \ldots, \lambda'_{n-1}$ be the singular values of $Z'$ where $Z'$ has $n-1$ columns each iid $N(0, \Sigma)$.
+Then we want to choose $\Sigma$ such that $E[\lambda'_i] = \lambda_i$ for each $i, where $E[Y]$ denotes the expectation of the random variable $Y$.
+
+Since the distribution of the $\lambda'_i$ does not have a known analytic solution, we approximate this situation with the spiked Wishart distribution.
+The rank $k-1$ Wishart distribution is that of the sample covariance matrix of $Y$ where the $n$ columns of $Y$ are iid $N(0, \Sigma)$.
+The spiked Wishart is the special case where $\Sigma$ has $k$ arbitrary eigenvalues and the remaining are all equal to a constant.
+Note that the singular values of $Y$ are the square roots of the eigenvalues of its sample covariance matrix.
+While our case has $\Sigma$ non-diagonal, $\Sigma$ may be diagonalized by orthogonal rotations due the sepctral theorem, and orthogonal rotations do not change the singular values of $Y$.
+Therefore, the distribution of singular values is not affected by the assumption that $\Sigma$ is diagonal.
+Moreover, for the form $\Sigma = D + U W U^T$ where $p$ is very large, each column of $U$ is typically very close to orthogonal to any $e_i$, a standard basis vector.
+Therefore, when $D = cI$ for some constant $c$, we can approximate $\Sigma$ as having $k$ arbitrary eigenvalues corresponding to $U$ and $n$ remaining eigenvalues all equal to $c$ corresponding to $D$.
+This is a spiked Wishart distribution.
+
+However, the spiked Wishart distribution also has no known analytic solution for the distribution of its eigenvalues either.
+Therefore, we use a stochastic gradient descent method that we recently described. <!-- TODO cite -->
+Since the dataset has been normalized, $c$ will be close to one and $e_i \Sigma e_i^T \approx 1$.
+
+Specifically, we do:
+1. Set $U$ to be the $p \times k$ matrix with columns $u_i$, the left singular vectors of $Z$.
+2. Compute $w_1, \ldots, w_k$ and $c$ by stochastic gradient descent minimizing $\sum_{i} (E[\lambda'_i] - \lambda_i)^2$ for $\Sigma$ diagonal with entries $w_1, \ldots, w_k, c, \ldots, c$.
+4. Set $W$ diagonal with the entries $w_1, \ldots, w_k$.
+3. Set $D = cI$
+
+### Corpcor Method
+
+The `corpcor` package <!-- TODO cite --> computes a James-Stein type shrinkage estimator for the covariance matrix.
+For large $p$, this greatly improves the estimate of the covariance matrix by introducing a little bias towards zero correlations and equal variances of genes.
+It computes optimal values of $\lambda_1$, and $\lambda_2$, its two regularization coefficients.
+It then uses $\lambda_1$ to linearly interpolate the sample covariance matrix towards the identity matrix $I$ and $\lambda_2$ to interpolate the vector of variances towards the median variance value.
+Since the sample covariance matrix is rank at most $n$, we again obtain a matrix of the form $\Sigma = D + UWU^T$.
+
+This algorithm is:
+1. Compute the $\lambda_1$ and $\lambda_2$ values from `corpcor::estimate.lambda` and `corpcor::estimate.lambda.var` functions on $Z$, respectively.
+2.
+Set $D$ to be diagonal with $D_{ii} = \sqrt{\lambda_1} (\lambda_2 \sigma_{med} + (1 - \lambda_2) \sigma_i)$ where $\sigma_i$ is the standard deviation of the $Z_{i\cdot}$ and $\sigma_{med}$ is the median of the $\sigma_i$.
+3. Set $U$ to be $\sqrt{1-\lambda} S Z / \sqrt{n-1}$ where $S$ is the diagonal matrix with $S_{ii} = \lambda_2 \sigma_{med} / \sigma_i + (1 - \lambda_2)$.
+4. Set $W$ to the identity.
 
@@
 We present `dependentsimr`, an R package that generates omics-scale data with realistic correlation.
-TODO: limitations TODO: alternatives
+We demonstrate this package on RNA-seq data, using the DESeq2 method to fit negative binomial marginal distributions.
+However, this package is actually quite general and supports normal, Poisson, negative binomial, and arbitrary ordered discrete distributions using the empirical CDF.
+Moreover, it can support multi-modal data such as is increasingly common in multi-omics.
+
+We demonstrate the importance of including gene-gene dependence in simulated data by two application benchmarks.
+In the first, we demonstrate that DESeq2 results are substantially more variable when simulating with gene-gene dependence.
+In the second, we demonstrate that the performance of CYCLOPS in estimating circadian phases depends upon gene-gene dependence, with an unexpected result of better performance in the dependent case.
+
+Our comparisons to a real dataset show that none of our three methods are able to exactly capture some aspects of the real dataset.
+In particular, the gene-gene correlations were too high in the PCA method and too low in the spiked Wishart and corpcor methods.
+Surprisingly, spiked Wishart and corpcor methods improved in this metric only slightly compared to the simulations with independent genes.
+These observations demonstrate that there is room for future improvements in these techniques and possibly demonstrate that the datasets are too far from a multivariate normal distribution even after normalization.
+Nonetheless, these methods represent significant improvements by other metrics and we recommend the inclusion of some dependence in nearly every simulated dataset.
+We highlight some alternatives that have been taken below.
+
+### Alternatives
+
+The R package [SPsimSeq](https://github.com/CenterForStatistics-UGent/SPsimSeq) <!-- TODO cite --> provides a dedicated RNA-seq and scRNA-seq simulator using a NORTA approach to simulate gene dependence.
+In contrast to this package, it uses WGCNA to determine the correlation matrix, which is a gene network approach.
+However, this method takes significant computational resources.
+Indeed, the `SPsimSeq` paper generated data for just 5000 genes based on a randomly sampled 5000 gene subset of the RNA-seq data and our attempts to use `SPsimSeq` to generate a full sample exhausted the memory of a 24GB computer.
+In contrast, our method runs in seconds to generate a 40,000 gene samples on the same computer.
+`SPsimSeq` is more specialized and full-featured for RNA-seq simulation, providing, for example, native differential expression (DE) options.
+In contrast, our `dependentsimr` package requires manually setting marginal expression values to inject DE, but also supports other marginal distributions for situations outside of RNA-seq.
+
+The [scDesign2](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-021-02367-2) simulator <!-- TODO cite --> for scRNA-seq uses NORTA and, like our method, uses the approach of estimating the correlation matrix from the normalized dataset.
+However, it limits this correlation matrix to top-expressed genes.
+
+Other Gaussian copula-based R packages that may be applicable, at least for datasets with smaller numbers of features, include `bindata`, `GenOrd`, and `SimMultiCorrData`, the last of these being the most comprehensive.
+The `bigsimr` package provides faster implementations of these methods to scale up to omics-level data.
+However, even this is computationally demanding; their paper references generating 20,000-dimensional vectors in “under an hour” using 16 threads.
+The `copula` package provides even more flexible dependence options through use of copulas (the NORTA approach is equivalent to using Gaussian copulas).
+All of these packages provide more flexibility in specifying dependence than our package, which can only mimic existing datasets, and therefore the longer run-times may be unavoidable for use cases where researchers need to parameterize the dependence structure.

### 2024-09-24 manuscript/paper.qmd
Add some references

@@
 <!-- TODO which dataset do we want to use here? -->
-We used mouse liver RNA-seq data set from accession GSE77221[@Weger2019] where we expected a significant amount of gene-gene correlation due to the samples having been collected throughout the course of a day.
+We used mouse liver RNA-seq data set from accession GSE77221 [@Weger2019] where we expected a significant amount of gene-gene correlation due to the samples having been collected throughout the course of a day.
 We used this data to prime our simulation and compared real and simulated data sets, both with 12 samples, to each other.
@@
 
-We benchmark DESeq2 <!-- TODO cite -->, a popular differential expression analysis tool, using data sets simulated with dependence and ones simulated without dependence to compare its performances on both.
+We benchmark DESeq2 [@Love2014], a popular differential expression analysis tool, using data sets simulated with dependence and ones simulated without dependence to compare its performances on both.
 We use a fly whole body RNA-Seq data set GSE81142 and select samples of male flies without treatment and after at least 2 hours of feeding to simulate 20 control samples with rank $k=2$ dependence structure, and 20 control samples assuming independence of the genes.
@@
 
-We further use our simulation method to benchmark CYCLOPS <!-- TODO cite -->, which infer relative times for a set of unlabeled samples using an autoencoder to identify circular structures.
+We further use our simulation method to benchmark CYCLOPS [@Anafi2017], which infer relative times for a set of unlabeled samples using an autoencoder to identify circular structures.
 We use a mouse liver time series data set (GSE151565, Aaronson et al.
@@
 Specifically, we do:
+
 1. Set $U$ to be the $p \times k$ matrix with columns $u_i$, the left singular vectors of $Z$.
@@
 
-The `corpcor` package <!-- TODO cite --> computes a James-Stein type shrinkage estimator for the covariance matrix.
+The `corpcor` package [@SchaferStrimmer2005; @OpgenRheinStrimmer2007] computes a James-Stein type shrinkage estimator for the covariance matrix.
 For large $p$, this greatly improves the estimate of the covariance matrix by introducing a little bias towards zero correlations and equal variances of genes.
@@
 This algorithm is:
+
 1. Compute the $\lambda_1$ and $\lambda_2$ values from `corpcor::estimate.lambda` and `corpcor::estimate.lambda.var` functions on $Z$, respectively.
@@
 
-The R package [SPsimSeq](https://github.com/CenterForStatistics-UGent/SPsimSeq) <!-- TODO cite --> provides a dedicated RNA-seq and scRNA-seq simulator using a NORTA approach to simulate gene dependence.
+The R package [SPsimSeq](https://github.com/CenterForStatistics-UGent/SPsimSeq) [@Assefa2020] provides a dedicated RNA-seq and scRNA-seq simulator using a NORTA approach to simulate gene dependence.
 In contrast to this package, it uses WGCNA to determine the correlation matrix, which is a gene network approach.
@@
 
-The [scDesign2](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-021-02367-2) simulator <!-- TODO cite --> for scRNA-seq uses NORTA and, like our method, uses the approach of estimating the correlation matrix from the normalized dataset.
+The [scDesign2](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-021-02367-2) simulator [@Sun2021] for scRNA-seq uses NORTA and, like our method, uses the approach of estimating the correlation matrix from the normalized dataset.
 However, it limits this correlation matrix to top-expressed genes.

### 2024-09-24 manuscript/paper.qmd
More references and tweaks

@@
 Third, the 'corpcor' method uses the popular R library `corpcor`, which implements a James-Stein type shrinkage estimator for the covariance matrix as a linear interpolation of the sample covariance matrix and a diagonal matrix.
-No method exactly captures the input data, indicating room for future research, but all improve upon a the common approach of assuming independence.
+No method exactly captures the input data, indicating room for future research, but all improve upon the common approach of assuming independence.
 
@@
 
-<!-- TODO which dataset do we want to use here? -->
-We used mouse liver RNA-seq data set from accession GSE77221 [@Weger2019] where we expected a significant amount of gene-gene correlation due to the samples having been collected throughout the course of a day.
-We used this data to prime our simulation and compared real and simulated data sets, both with 12 samples, to each other.
-We repeated the simulations a total of 8 times to estimate variance.
-The simulations were then repeated without any dependence for comparison.
+We chose as a references dataset, 12 mouse liver RNA-seq samples from accession GSE151923 [@Wang2022].
+We then simulated data mimicking this reference using all three simulation methods (PCA, spiked Wishart, and corpcor) as well as a simulation with independent genes.
+We repeated the simulations, each of 12 samples, a total of 8 times to estimate variance.
+For the PCA method, we used $k=2$ and for the spiked Wishart, $k=11$.
+The coprcor method always uses the full data matrix, analogous to $k=11$.
+Note that PCA method must use a rank $k<11$ in order to generate full-rank data, see Methods, so these parametesr are not directly comparable across methods.
 
 Simulated data captures the genes' mean and variance accurately (@fig-compare-to-real a-b).
-Moreover, we computed the gene-gene correlation on pairs of high-expressed genes.
-Since only 12 samples are in each real or simulated data set, correlation estimates are very noisy.
-However, the distribution of correlations across all pairs of these genes is much closer to that of real data when simulated with dependence than without (@fig-compare-to-real c).
+Next, we compared to the real data set when projected onto the top two principal components of the real data set (@fig-compare-to-real c).
+The simulations with dependence are distributed around the entire space like the real data, but the independent simulations have unrealistically low variance in these components, clustering tightly around the origin.
 
-Next, we compared the simulations with and without dependence to the real data set when projected onto the top two principal components of the real data set (@fig-compare-to-real d).
-The simulations with dependence are distributed around the entire space like the real data, but the independent simulations have unrealistically low variance in these components, clustering tightly around the origin.
-Lastly, we compared the PCA of each simulated data set considered separately from the real data.
-Simulations with independent data showed unrealistically low levels of variance in top principal components, but dependent simulations actually show too high elevations of variance in top components.
-This is a consequence of our PCA strategy: since variance in the direction of the real data set's top principal component is exactly matched, the simulated data set's top principal component will be even larger than that.
+Then, we computed the gene-gene correlation on pairs of high-expressed genes (at least 300 mean reads).
+The simulation with independence showed the least levels of gene-gene correlations (@fig-compare-to-real d).
+However, the PCA method overshot the reference data set and the spiked Wishart and corpcor methods only slightly improved upon the independent simulation.
+
+Lastly, we compared the variances of principal components on each data set (@fig-compare-to-real e).
+These were computed separately for each data set, unlike (@fig-compare-to-real c) which used the reference data set's PCA weights for all data sets.
+The independent data has much lower variance than the real data set in the top four principal components.
+The spiked Wishart method comes closest to the real data set, as it optimizes for fitting these values.
+Surprisingly, the corpcor method performs only somewhat better than the independent method.
+The PCA method puts a large amount of variance into the first two components (due to using $k=2$) and then undershoots the other components.
 
@@
 We benchmark DESeq2 [@Love2014], a popular differential expression analysis tool, using data sets simulated with dependence and ones simulated without dependence to compare its performances on both.
+DESeq2 presents an interesting case because several aspects of it assume independence of genes and so may be adversely affected by gene-gene dependence.
+First, the independent filtering step [@Bourgon2010-uv] assumes independence but has generally be regarded to be robust to typical gene-gene dependence.
+Relatedly, the false discovery rate (FDR) [@BH_FDR] allows only certain forms of dependence.
+Lastly, DESeq2's empirical Bayes steps could possibly be affected by gene dependence.
+
 We use a fly whole body RNA-Seq data set GSE81142 and select samples of male flies without treatment and after at least 2 hours of feeding to simulate 20 control samples with rank $k=2$ dependence structure, and 20 control samples assuming independence of the genes.
@@
 However, there is a greater variance in the performance of DESeq2 on the data sets simulated with dependence, indicating that it potentially performs less reliably on data sets with gene-gene dependence, as in real data sets.
-We speculate that this is due to DESeq2's assumption that the gene expressions are independent.
 
-To demonstrate the application of our simulation method for another organism, we also simulate data sets using a mouse cortex data set GSE151923 and select samples from male mice to infer the dependence structure of the quantified genes.
+To demonstrate the application of our simulation method for another organism, we also simulate data sets using a mouse cortex data set GSE151923 [@Wang2022] and select samples from male mice to infer the dependence structure of the quantified genes.
 We then simulate differential expression experiments with and without dependence as above, with 20 samples for each condition.
@@
 #| label: fig-DESeq2
-#| fig-cap: "Performance of DESeq2 on simulated datasets.
-(a-b) Comparison of true FDR and DESeq2 reported FDR for data sets simulated from the fly whole body data set (GSE81142), (a) without dependence and (b) with dependence.
+#| fig-cap: "Performance of DESeq2 on simulated datasets.(a-b) Comparison of true FDR and DESeq2 reported FDR for data sets simulated from the fly whole body data set (GSE81142), (a) without dependence and (b) with dependence.
+Diagonal line represents perfect estimation of FDR.
 (c-d) Comparison of true FDR and DESeq2 reported FDR for data sets simulated from the mouse cortex data set (GSE151923, Wang et al.
@@
 For the evaluation metric for CYCLOPS' performance, we use the circular correlation, defined as follows:
-$$\rho = \frac{\sum_{1\leq i<j\leq n}sin(X_i-X_j)sin(Y_i-Y_j)}{(\sum_{1\leq i<j\leq n}sin(X_i-X_j)^2)^{1/2}(\sum_{1\leq i<j\leq n}sin(Y_i-Y_j)^2)^{1/2}},$$
+$$\rho = \frac{\sum_{1\leq i<j\leq n}\sin(X_i-X_j) \sin(Y_i-Y_j)}{(\sum_{1\leq i<j\leq n} \sin(X_i-X_j)^2)^{1/2}(\sum_{1\leq i<j\leq n} \sin(Y_i-Y_j)^2)^{1/2}},$$
 where $n$ is the number of samples, $X_i$ and $Y_i$ are the true time and CYCLOPS-estimated time respectively for the $i$-th sample.
@@
 (a) Violin plot comparing absolute circular correlations between true phases and CYCLOPS estimated phases on the simulated data sets with and without dependence.
-(b-c) Examples of CYCLOPS estimated phases on a simulated data set (b) without dependence and (c) with dependence."
+(b-c) Examples of CYCLOPS estimated phases on a simulated data set (b) without dependence and (c) with dependence.
+CYCLOPS shows good performance when it separates out points by color (true circadian time)."
 cyclops_plot

### 2024-09-24 manuscript/circular_correlation.R

@@ -74,7 +74,7 @@ phaselist$true_counts_counts_counts_time <- as.numeric(phaselist$ID)
 phaselist$sin_estimated_phase <- sin(phaselist$phase)
 phaselist$cos_estimated_phase <- cos(phaselist$phase)
 phaseplot_0 <- ggplot(phaselist,aes(cos_estimated_phase,sin_estimated_phase))+
-    scale_color_gradientn(colors = rainbow(8), limits=c(0, 24)) +
+    scale_color_gradientn(colors = rainbow(8), limits=c(0, 24), name="true time") +
     geom_point(aes(colour=true_counts_counts_counts_time)) +
     xlim(-1, 1) +
     ylim(-1, 1) +
@@ -95,7 +95,7 @@ phaselist$true_counts_counts_counts_time <- as.numeric(phaselist$ID)
 phaselist$sin_estimated_phase <- sin(phaselist$phase)
 phaselist$cos_estimated_phase <- cos(phaselist$phase)
 phaseplot_2 <- ggplot(phaselist,aes(cos_estimated_phase,sin_estimated_phase))+
-    scale_color_gradientn(colors = rainbow(8), limits=c(0, 24)) +
+    scale_color_gradientn(colors = rainbow(8), limits=c(0, 24), name="true time") +
     geom_point(aes(colour=true_counts_counts_counts_time)) +
     xlim(-1, 1) +
     ylim(-1, 1) +
@@ -104,4 +104,4 @@ phaseplot_2
 #ggsave(paste0("plots/k=2_batch=", j, "_scatter.png"),
 #       scale = 1, width = 5, height = 4, units = "in")
 
-cyclops_plot <- circular_correlation_violin / (phaseplot_0 | phaseplot_2) + plot_annotation(tag_levels="a")
+cyclops_plot <- circular_correlation_violin / (phaseplot_0 | phaseplot_2) + plot_annotation(tag_levels="a") + plot_layout(guides = 'collect')

### 2024-09-24 manuscript/compare_to_real.R

@@ -16,15 +16,12 @@ HIGH_EXPR_CUTOFF <- 300
 TYPE_ORDER <- c("real", "indep", "pca", "wishart", "corpcor")
 TYPE_COLORS <- RColorBrewer::brewer.pal(length(TYPE_ORDER), "Set2")
 
-## Generate the simulated data
 # Load the real data ---------------------------------
-# from GEO: GSE77221
 raw <- read_tsv("../data/GSE151923_metaReadCount_ensembl.txt.gz") |>
   select(-GeneName, -Description, -Chromosome, -Strand)
 read_counts <- as.matrix(raw[,2:13])
 rownames(read_counts) <- raw$GeneID
 
-#head(Weger18)
 actual_library_sizes <- apply(read_counts, 2, sum)
 
 # Run dependentsimr on this data --------------------
@@ -78,6 +75,31 @@ g2<-ggplot(marg_dist|> filter(real_mean > 0.1))+
   geom_abline(slope=1, intercept=0) +
   scale_color_viridis_c(option="inferno")
 
+# Plot the top principal components --------------------
+constant_rows <- apply(read_counts, 1, function(x) all(x - mean(x) == 0))
+pca <- prcomp(t(scaled_read_data[!constant_rows,]), rank.=2, scale.=TRUE)
+projected_read_data <- predict(pca, t(scaled_read_data[!constant_rows, ]))
+projected_draws <- predict(pca, t(scaled_draws[!constant_rows, ]))
+projected_corpcor <- predict(pca, t(scaled_draws_corpcor[!constant_rows, ]))
+projected_wishart <- predict(pca, t(scaled_draws_wishart[!constant_rows, ]))
+projected_indep_draws <- predict(pca, t(scaled_indep_draws[!constant_rows, ]))
+both_data <- data.frame(list(
+  PC1 = c(projected_read_data[,"PC1"], projected_draws[,"PC1"], projected_corpcor[,"PC1"], projected_wishart[,"PC1"], projected_indep_draws[,"PC1"]),
+  PC2 = c(projected_read_data[,"PC2"], projected_draws[,"PC2"], projected_corpcor[,"PC2"], projected_wishart[,"PC2"], projected_indep_draws[,"PC2"]),
+  type = c(
+    rep("real", nrow(projected_read_data)),
+    rep("pca", nrow(projected_draws)),
+    rep("corpcor", nrow(projected_draws)),
+    rep("wishart", nrow(projected_draws)),
[…]

### 2024-09-26 scripts/simulate_data.R
work in progress for snakefile and simulate_data.R

@@ -4,6 +4,8 @@ library(tibble)
 library(DESeq2)
 library(rlang)
 
+method <- snakemake@output$wildcards.method
+
 metadata <- read.delim("processed/GSE151923_sample_metadata.txt", sep="\t")
 read_counts <- read.delim("data/GSE151923_metaReadCount_ensembl.txt.gz", sep="\t")
 
@@ -35,7 +37,21 @@ female_read_counts <- read_counts[,sex == 'female']
 ###lfc <- results(dds, c("sex", "female", "male"))$log2FoldChange
 
 # Produce simulated data
-male_rs <- get_random_structure(list(data=male_read_counts), method="pca", rank=2, types="DESeq2")
+if (method == "pca") {
+    # Simulate with the PCA method
+    rs <- get_random_structure(list(counts=read_counts), method="pca", rank=2, type="DESeq2")
+} else if (method == "corpcor") {
+    # Simulate with the corpcor method
+    rs <- get_random_structure(list(counts=read_counts), method="corpcor", type="DESeq2")
+} else if (method == "wishart") {
+    # Simulate with the spiked Wishart method
+    rs <- get_random_structure(list(counts=read_counts), rank=11, method="spiked Wishart", type="DESeq2")
+} else if (method == "indep") {
+    # Simulate without any dependence
+    rs <- get_random_structure(list(counts=read_counts), method="pca", rank=2, type="DESeq2")
+    rs <- remove_dependence(rs)
+}
+
 
 actual_library_sizes <- male_read_counts |> apply(2, sum)
 
@@ -67,11 +83,7 @@ generate <- function(rs, seed, name) {
 
 ## Generate Controls
 # With dependence
-male_sim <- generate(male_rs, seed=1, name="Mouse.Cortex.Male.k=2.Control")
-
-# Without dependence (k=0)
-male_indep_rs <- remove_dependence(male_rs)
-male_indep_sim <- generate(male_indep_rs, seed=2, name="Mouse.Cortex.Male.k=0.Control")
+male_sim <- generate(rs, seed=1, name=paste0("Mouse.Cortex.Male.",method,".Control"))
 
 # Randomly select some genes to be DE
 set.seed(0)

### 2024-10-11 manuscript/paper.qmd
Add references and edit manuscript
Use same mouse liver data for comparison to real

@@
 title: Generating Correlated Data for Omics Simulation
+author:
+    - name: "Jianing Yang"
+      affiliations:
+        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania"
+        - name: "Chronobiology and Sleep Institute, University of Pennsylvania"
+    - name: "Gregory R. Grant"
+      orcid: 0000-0002-0139-7658
+      affiliations:
+        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania"
+    - name: "Thomas G. Brooks"
+      orcid: 0000-0002-6980-0079
+      affiliations:
+        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania"
 editor: 
@@
 This makes it particularly suited to RNA-seq data but also widely applicable.
+Moreover, it supports multi-modal data such as for multi-omics.
 
 We implement three different strategies for determining the diagonal and low-rank parts of the covariance matrix.
-First, the 'PCA' method uses principal component analysis (PCA) and pick the low-rank part such that the simulated data has the same variance in the top $k$ PCA components of the reference dataset.
-Second, the 'spiked Wishart' method fits $k$ components such that simulations with the same number of samples as the reference dataset will have, on average, the same PCA component variances as the reference
+First, the 'PCA' method uses principal component analysis (PCA) and picks the low-rank part such that the simulated data has the same variance in the top $k$ PCA components of the reference dataset.
+Second, the 'spiked Wishart' method fits $k$ components such that simulations with the same number of samples as the reference dataset will have, on average, the same PCA component variances as the reference.
 Unlike 'PCA', these variances are computed with resepect to the simulated data's own PCA and not using the PCA weights of the reference dataset.
@@
 First, we simulate RNA-seq data with differential expression between two conditions.
-Using DESeq2 to determine the differentially expressed genes, we found that dependence had little impact on the accuracy of reported p-values but increased the variance of those estimates, meaning that experiments simulated with dependence were more likely to be conservative and more likely to be anti-conservative compared to simulations of independent data.
+Using DESeq2 to determine the differentially expressed genes, we found that dependence had little impact on the accuracy of reported p-values but increased the variance of those estimates.
 Second, we simulated a time series of RNA-seq data points and used the CYCLOPS method to infer collection time from the RNA-seq data, without time labels.
-Surprisingly, we found that including dependence in the data set resulted in improved performance with CYCLOPS.
+Surprisingly, we found that including dependence in the data set improved the performance of CYCLOPS.
 
@@
 
-We chose as a references dataset, 12 mouse liver RNA-seq samples from accession GSE151923 [@Wang2022].
+We chose as a references dataset, 12 mouse cortex RNA-seq samples from accession GSE151923 [@Wang2022].
 We then simulated data mimicking this reference using all three simulation methods (PCA, spiked Wishart, and corpcor) as well as a simulation with independent genes.
@@
 The coprcor method always uses the full data matrix, analogous to $k=11$.
-Note that PCA method must use a rank $k<11$ in order to generate full-rank data, see Methods, so these parametesr are not directly comparable across methods.
+Note that PCA method must use a rank $k<11$ in order to generate full-rank data, see Methods, so these parameters are not directly comparable across methods.
 
@@
 #| label: fig-compare-to-real
-#| fig-cap: !expr glue::glue("Comparison to real data run on a mouse liver example from GSE77221.
-(a-b) Comparison of gene (a) mean expression and (b) variance, log-scaled.
+#| fig-cap: !expr glue::glue("Comparison to real data run on a mouse cortex data set from GSE151923.
+(a-b) Comparison of gene (a) mean expression and (b) variance, log-scaled in real and PCA simulated data.
 The line of equality is marked in black.
 Points are colored according to the density of points in their region.
+Wishart and corpcor methods give similar results (not shown).
 (c) Quantile-quantile plot comparing correlation values of gene pairs from real data and simulated data (both with and without dependence).
@@
 
-We benchmark DESeq2 [@Love2014], a popular differential expression analysis tool, using data sets simulated with dependence and ones simulated without dependence to compare its performances on both.
+We benchmarked DESeq2 [@Love2014], a popular differential expression analysis tool, using data sets simulated with dependence and ones simulated without dependence to compare its performances on both.
 DESeq2 presents an interesting case because several aspects of it assume independence of genes and so may be adversely affected by gene-gene dependence.
@@
 
-We use a fly whole body RNA-Seq data set GSE81142 and select samples of male flies without treatment and after at least 2 hours of feeding to simulate 20 control samples with rank $k=2$ dependence structure, and 20 control samples assuming independence of the genes.
-We then randomly select 5% of the genes to be up or down regulated, with $\log_2$ fold change uniformly distributed between $0.2$ and $2.0$, and simulate 20 "experimental" samples each with and and without dependence.
+We used a fly whole body RNA-Seq data set GSE81142 and selected samples of male flies without treatment and after at least 2 hours of feeding to simulate 5 "control" samples.
+We then randomly select 5% of the genes to be up or down regulated, with $\log_2$ fold change uniformly distributed between $0.2$ and $2.0$, and simulate 5 "experimental" samples.
+This was repeated 20 times for each of four dependence conditions (independent, PCA $k=2$, Wishart $k=11$, and corpcor).
 
-Finally, we run DESeq2 on the two simulated experiments and compare the output false discovery rate (FDR) with the true percentages of genes that are differential expressed (@fig-DESeq2 a-b).
-We observe that DESeq2 is slightly anti-conservative on both data sets, with similar mean true FDRs for each estimated FDR cutoff.
-However, there is a greater variance in the performance of DESeq2 on the data sets simulated with dependence, indicating that it potentially performs less reliably on data sets with gene-gene dependence, as in real data sets.
+Finally, we ran DESeq2 on each simulated 5 vs 5 experiment and compared the output false discovery rate (FDR) with the true percentages of genes that are differential expressed (@fig-DESeq2 a-b).
+We observed that DESeq2 is anti-conservative on both data sets, with similar mean true FDRs for each estimated FDR cutoff.
+However, there was a greater variance in the performance of DESeq2 on the data sets simulated with dependence, indicating that it performs less consistently on data sets with gene-gene dependence, as in real data sets.
 
-To demonstrate the application of our simulation method for another organism, we also simulate data sets using a mouse cortex data set GSE151923 [@Wang2022] and select samples from male mice to infer the dependence structure of the quantified genes.
-We then simulate differential expression experiments with and without dependence as above, with 20 samples for each condition.
-We observe a similar result (@fig-DESeq2 c-d) as for the fly whole body data sets.
+To demonstrate the application of our simulation method for another organism, we also simulated data sets using mouse cortex data set GSE151923 [@Wang2022] and selected samples from male mice.
+We then simulated differential expression experiments as above and observed a similar result (@fig-DESeq2 c-d) as for the fly whole body data sets.
 
@@
 
-We further use our simulation method to benchmark CYCLOPS [@Anafi2017], which infer relative times for a set of unlabeled samples using an autoencoder to identify circular structures.
+We further used our simulation method to benchmark CYCLOPS [@Anafi2017], which infers relative times for a set of unlabeled samples using an autoencoder to identify circular structures.
 We use a mouse liver time series data set (GSE151565, Aaronson et al.
@@
 
-We run CYCLOPS on each data set with a list of cyclic mouse liver genes (JTK p-value $<0.05$), which yields an estimated relative time for each sample.
-For the evaluation metric for CYCLOPS' performance, we use the circular correlation, defined as follows:
+We ran CYCLOPS on each data set with a list of cyclic mouse liver genes (JTK p-value $<0.05$), which yielded an estimated relative time for each sample.
+We evaluated CYCLOPS' performance compared to true circadian time using the circular correlation, defined as follows:
 $$\rho = \frac{\sum_{1\leq i<j\leq n}\sin(X_i-X_j) \sin(Y_i-Y_j)}{(\sum_{1\leq i<j\leq n} \sin(X_i-X_j)^2)^{1/2}(\sum_{1\leq i<j\leq n} \sin(Y_i-Y_j)^2)^{1/2}},$$
-where $n$ is the number of samples, $X_i$ and $Y_i$ are the true time and CYCLOPS-estimated time respectively for the $i$-th sample.
+where $n$ is the number of samples, $X_i$ and $Y_i$ are the true time and CYCLOPS-estimated time, respectively, for the $i$-th sample.
 $\rho$ has value between $-1$ and $1$, and a $|\rho|$ close to $1$ indicates accurate predictions by CYCLOPS.
 
-We find that CYCLOPS performs better overall on simulated time series data sets with dependence than on those without dependence (@fig-CYCLOPS), indicating that the presence of gene-gene dependence is essential for CYCLOPS' performance.
+We find that CYCLOPS performance depends significantly on the dependence structure of the simulated time series data (@fig-CYCLOPS), indicating that the presence of gene-gene dependence is highly impactful on CYCLOPS' performance.
+Overall, CYCLOPS performance is poor in the independent data and improved in both PCA and corpcor data, but surprisingly the Wishart data also shows poor performance.
+Since by default CYCLOPS performs dimension reduction to the number of dimensions needed to capture 99% of the variance, we hypothesize that data with dependence allows fewer dimensions to be used which improved the performance.
 
@@
 2020).
-(a) Violin plot comparing absolute circular correlations between true phases and CYCLOPS estimated phases on the simulated data sets with and without dependence.
-(b-c) Examples of CYCLOPS estimated phases on a simulated data set (b) without dependence and (c) with dependence.
+(a) Violin plot comparing absolute circular correlations between true phases and CYCLOPS estimated phases on the simulated data sets under the four dependence methods.
+(b-e) Examples of CYCLOPS estimated phases.
 CYCLOPS shows good performance when it separates out points by color (true circadian time)."
@@
 Conveniently, this form allows efficient simulation of multivariate normal data $N(0, \Sigma)$.
-Once data is obtained $Z ~ N(0, \Sigma)$, then undo the normalization process to obtain data with the same marginal distributions as the fit marginal distributions but with dependence determined by $\Sigma$.
+Once data is obtained $Z \sim N(0, \Sigma)$, then one can undo the normalization process to obtain data with the same marginal distributions as the fit marginal distributions but with dependence determined by $\Sigma$.
 We describe this in detail below:
@@
 
-The generated data $Z'$ has covaraince matrix $\Sigma = D + U W U^T$.
+The generated data $Z'$ has covariance matrix $\Sigma = D^2 + U W U^T$.
 Moreover, we require that $\Sigma$ satisfies that $e_i \Sigma e_i^T$ is approximately 1.
-That guarantees that the output $X'$ has each entry with the same mariginal distributions $F_i$ as was originally fit and inherits gene-gene dependence from $Z'$.
+That guarantees that the output $X'$ has each entry with the same marginal distributions $F_i$ as was originally fit and inherits gene-gene dependence from $Z'$.
 
@@
 
-Step 4 ensures that $e_i \Sigma e_i^T = 1$.
+Steps 1 and 2 give that $u_i \Sigma u_i^T = \lambda_i$ for $i = 1, \ldots, k$.
+Step 4 ensures that $e_j \Sigma e_j^T = 1$ for $j = 1, \ldots, p$.
 
@@
 If $n$ samples are drawn from $N(0, \Sigma)$ then we want the variances of their PCA components to match those of the reference dataset.
-Specifically, let $\lambda_1, \ldots, \lambda_{n-1}$ are the ${n-1}$ non-zero singular values of $Z$ ($\lambda_n$ is always approximately zero due to the normalization procedure), and let $\lambda'_1, \ldots, \lambda'_{n-1}$ be the singular values of $Z'$ where $Z'$ has $n-1$ columns each iid $N(0, \Sigma)$.
-Then we want to choose $\Sigma$ such that $E[\lambda'_i] = \lambda_i$ for each $i, where $E[Y]$ denotes the expectation of the random variable $Y$.
+Specifically, let $\lambda_1, \ldots, \lambda_{n-1}$ be the ${n-1}$ non-zero singular values of $Z$ ($\lambda_n$ is always approximately zero due to the normalization procedure), and let $\lambda'_1, \ldots, \lambda'_{n-1}$ be the singular values of $Z'$ where $Z'$ has $n-1$ columns each iid $N(0, \Sigma)$.
+Then we want to choose $\Sigma$ such that $E[\lambda'_i] = \lambda_i$ for each $i$, where $E[Y]$ denotes the expectation of the random variable $Y$.
 
@@
 Moreover, for the form $\Sigma = D + U W U^T$ where $p$ is very large, each column of $U$ is typically very close to orthogonal to any $e_i$, a standard basis vector.
-Therefore, when $D = cI$ for some constant $c$, we can approximate $\Sigma$ as having $k$ arbitrary eigenvalues corresponding to $U$ and $n$ remaining eigenvalues all equal to $c$ corresponding to $D$.
+Therefore, when $D = cI$ for some constant $c$, we can approximate $\Sigma$ as having $k$ arbitrary eigenvalues from $W$ and $n$ remaining eigenvalues all equal to $c$ corresponding to $D$.
 This is a spiked Wishart distribution.
@@
 However, the spiked Wishart distribution also has no known analytic solution for the distribution of its eigenvalues either.
-Therefore, we use a stochastic gradient descent method that we recently described. <!-- TODO cite -->
+Therefore, we use an efficient sampling and stochastic gradient descent method that we recently described [@wishart].
 Since the dataset has been normalized, $c$ will be close to one and $e_i \Sigma e_i^T \approx 1$.
@@
 Surprisingly, spiked Wishart and corpcor methods improved in this metric only slightly compared to the simulations with independent genes.
-These observations demonstrate that there is room for future improvements in these techniques and possibly demonstrate that the datasets are too far from a multivariate normal distribution even after normalization.
+These observations demonstrate that there is room for future improvements over independent data in these techniques and possibly demonstrate that the datasets are too far from a multivariate normal distribution even after normalization.
 Nonetheless, these methods represent significant improvements by other metrics and we recommend the inclusion of some dependence in nearly every simulated dataset.
@@
 
-The R package [SPsimSeq](https://github.com/CenterForStatistics-UGent/SPsimSeq) [@Assefa2020] provides a dedicated RNA-seq and scRNA-seq simulator using a NORTA approach to simulate gene dependence.
+The R package [SPsimSeq](https://github.com/CenterForStatistics-UGent/SPsimSeq) [@Assefa2020] provides a dedicated RNA-seq and single-cell RNA-seq simulator using a Gaussian copula approach to simulate gene dependence.
 In contrast to this package, it uses WGCNA to determine the correlation matrix, which is a gene network approach.
@@
 `SPsimSeq` is more specialized and full-featured for RNA-seq simulation, providing, for example, native differential expression (DE) options.
-In contrast, our `dependentsimr` package requires manually setting marginal expression values to inject DE, but also supports other marginal distributions for situations outside of RNA-seq.
+In comparison, our `dependentsimr` package requires manually setting marginal expression values to inject DE, but also supports other marginal distributions for situations outside of RNA-seq.
 
-The [scDesign2](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-021-02367-2) simulator [@Sun2021] for scRNA-seq uses NORTA and, like our method, uses the approach of estimating the correlation matrix from the normalized dataset.
+The [scDesign2](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-021-02367-2) simulator [@Sun2021] for single-cell RNA-seq also uses Gaussian copula and, like our method, uses the approach of estimating the correlation matrix from the normalized dataset.
 However, it limits this correlation matrix to top-expressed genes.
+Since correlation is most discernible in high-expressed genes, this approach is reasonable but requires making certain arbitrary cutoffs that our methods avoid.
 

### 2024-10-11 manuscript/compare_to_real.R

@@ -20,7 +20,7 @@ TYPE_COLORS <- RColorBrewer::brewer.pal(length(TYPE_ORDER), "Set2")
 raw <- read_tsv("../data/GSE151923_metaReadCount_ensembl.txt.gz") |>
   select(-GeneName, -Description, -Chromosome, -Strand)
 read_counts <- as.matrix(raw[,2:13])
-rownames(read_counts) <- raw$GeneID
+rownames(read_counts) <- raw[[1]]
 
 actual_library_sizes <- apply(read_counts, 2, sum)
 

### 2024-10-11 manuscript/paper.qmd
added violin plot of number of eigengenes used by CYCLOPS.

--- 2024-09-24 manuscript/paper.qmd
@@
 
-We use a fly whole body RNA-Seq data set GSE81142 and select samples of male flies without treatment and after at least 2 hours of feeding to simulate 20 control samples with rank $k=2$ dependence structure, and 20 control samples assuming independence of the genes.
-We then randomly select 5% of the genes to be up or down regulated, with $\log_2$ fold change uniformly distributed between $0.2$ and $2.0$, and simulate 20 "experimental" samples each with and and without dependence.
+We use a fly whole body RNA-Seq data set GSE81142 and select samples of male flies without treatment and after at least 2 hours of feeding to simulate 20 control samples with each of the three simulation methods with dependence structure-PCA, Wishart and corpcor, as well as assuming independence of the genes.
+We then randomly select 5% of the genes to be up or down regulated, with $\log_2$ fold change uniformly distributed between $0.2$ and $2.0$, and simulate 20 "experimental" samples with each of the four methods.
 
-Finally, we run DESeq2 on the two simulated experiments and compare the output false discovery rate (FDR) with the true percentages of genes that are differential expressed (@fig-DESeq2 a-b).
+Finally, we run DESeq2 on the four simulated experiments and compare the output false discovery rate (FDR) with the true percentages of genes that are differential expressed (@fig-DESeq2 a-d).
 We observe that DESeq2 is slightly anti-conservative on both data sets, with similar mean true FDRs for each estimated FDR cutoff.
@@
 To demonstrate the application of our simulation method for another organism, we also simulate data sets using a mouse cortex data set GSE151923 [@Wang2022] and select samples from male mice to infer the dependence structure of the quantified genes.
-We then simulate differential expression experiments with and without dependence as above, with 20 samples for each condition.
-We observe a similar result (@fig-DESeq2 c-d) as for the fly whole body data sets.
+We then simulate differential expression experiments with the four simulation methods as above, with 20 samples for each condition.
+We observe a similar result (@fig-DESeq2 e-h) as for the fly whole body data sets.
 
@@
 #| label: fig-DESeq2
-#| fig-cap: "Performance of DESeq2 on simulated datasets.(a-b) Comparison of true FDR and DESeq2 reported FDR for data sets simulated from the fly whole body data set (GSE81142), (a) without dependence and (b) with dependence.
+#| fig-cap: "Performance of DESeq2 on simulated datasets.(a-d) Comparison of true FDR and DESeq2 reported FDR for data sets simulated from the fly whole body data set (GSE81142), (a) without dependence, (b) using PCA, (c) using Wishart and (d) using corpcor.
 Diagonal line represents perfect estimation of FDR.
-(c-d) Comparison of true FDR and DESeq2 reported FDR for data sets simulated from the mouse cortex data set (GSE151923, Wang et al.
-2022), (c) without dependence and (d) with dependence."
-#| fig-width: 7
+(e-h) Comparison of true FDR and DESeq2 reported FDR for data sets simulated from the mouse cortex data set (GSE151923, Wang et al.
+2022), (e) without dependence, (f) using PCA, (g) using Wishart and (h) using corpcor."
+#| fig-width: 12
 #| fig-height: 6
@@
 
-We find that CYCLOPS performs better overall on simulated time series data sets with dependence than on those without dependence (@fig-CYCLOPS), indicating that the presence of gene-gene dependence is essential for CYCLOPS' performance.
+We find that CYCLOPS performs better overall on time series data sets simulated with PCA and corpcor than on those without dependence and with Wishart (@fig-CYCLOPS), indicating that the presence of gene-gene dependence, as well as the method used to simulate dependence is essential for CYCLOPS' performance.
+This is likely due to the number of eigengenes needed by CYCLOPS to meet the variance threshold (@fig_CYCLOPS b).
 
@@
 2020).
-(a) Violin plot comparing absolute circular correlations between true phases and CYCLOPS estimated phases on the simulated data sets with and without dependence.
-(b-c) Examples of CYCLOPS estimated phases on a simulated data set (b) without dependence and (c) with dependence.
+(a) Violin plot comparing absolute circular correlations between true phases and CYCLOPS estimated phases on the simulated data sets with the four methods.
+(b) Violin plot comparing the number of eigengenes used by CYCLOPS between the four methods; the red dotted line indicates the number of eigengenes used by CYCLOPS on the real data (13).
+(c-f) Examples of CYCLOPS estimated phases on a simulated data set (b) without dependence, (c) with PCA, (d) with Wishart and (d) with corpcor.
 CYCLOPS shows good performance when it separates out points by color (true circadian time)."
+#| fig-width: 12
+#| fig-height: 8
 cyclops_plot

### 2024-10-11 manuscript/circular_correlation.R

@@ -74,7 +74,19 @@ for (method in c("indep", "pca", "wishart", "corpcor")) {
         ggtitle(method)
 }
 
+# plot the number of eigengenes used by CYCLOPS
+eigen <- read.delim("num_eigengene.txt", header = FALSE, sep=" ") |> 
+    select(V5, V6) |> 
+    rename(num = V5, method = V6) |> 
+    mutate(method = sapply(strsplit(method,"/|\\\\"), '[', 4))
+eigen <- eigen[1:80,]
+eigen_violin <- ggplot(eigen, aes(method, num)) +
+    geom_violin() +
+    ylab("number of eigengenes") +
+    geom_hline(yintercept = 13, linetype = "dashed", color = "red")
+
 cyclops_plot <- circular_correlation_violin / 
+    eigen_violin /
     (phaseplots$indep | phaseplots$pca | phaseplots$wishart | phaseplots$corpcor) + 
     plot_annotation(tag_levels="a")# + plot_layout(guides = 'collect')
 cyclops_plot

### 2024-10-18 manuscript/paper.qmd
Add references and edit manuscript
Use same mouse liver data for comparison to real

--- 2024-10-11 manuscript/paper.qmd
@@
 Overall, CYCLOPS performance is poor in the independent data and improved in both PCA and corpcor data, but surprisingly the Wishart data also shows poor performance.
-Since by default CYCLOPS performs dimension reduction to the number of dimensions needed to capture 99% of the variance, we hypothesize that data with dependence allows fewer dimensions to be used which improved the performance.
+Since by default CYCLOPS performs dimension reduction to the number of dimensions needed to capture 99% of the variance, this performance difference is likely due to the number of eigengenes needed by CYCLOPS to meet the variance threshold (@fig_CYCLOPS b).
 
@@
 #| label: fig-CYCLOPS
-#| fig-cap: "Performance of CYCLOPS on simulated time series data sets based on mouse liver data set (GSE151565, Aaronson et al.
-2020).
-(a) Violin plot comparing absolute circular correlations between true phases and CYCLOPS estimated phases on the simulated data sets under the four dependence methods.
-(b-e) Examples of CYCLOPS estimated phases.
+#| fig-cap: "Performance of CYCLOPS on simulated time series data sets based on mouse liver data set (GSE151565).
+(a) Violin plot comparing absolute circular correlations between true phases and CYCLOPS estimated phases on the simulated data sets with the four methods.
+(b) Violin plot comparing the number of eigengenes used by CYCLOPS between the four methods; the red dotted line indicates the number of eigengenes used by CYCLOPS on the real data (13).
+(c-f) Examples of CYCLOPS estimated phases on the simulated data sets.
 CYCLOPS shows good performance when it separates out points by color (true circadian time)."
+#| fig-width: 12
+#| fig-height: 8
 cyclops_plot

### 2024-10-18 manuscript/paper.qmd
Edits

@@
     - name: "Jianing Yang"
+      orcid: 0000-0002-2048-9398
       affiliations:
@@
 First, we show that variance of results from the popular DESeq2 method increases when dependence is included.
-Second, we demonstrate that a method, CYCLOPS, for inferring circadian time of collection from transcriptomics actually improves its performance when dependence is included.
+Second, we demonstrate that a method, CYCLOPS, for inferring circadian time of collection from transcriptomics performs substantially worse or better under different dependence situations.
 We provide an R package, dependentsimr, that has efficient implementations of these methods and can generate dependent data with arbitrary distributions, including discrete (binary, ordered categorical, Poisson, negative binomial), continuous (normal), or with an empirical distribution.
@@
 Recently, we wrote guidelines for performing omics benchmarking [@Brooks2024] and encouraged authors to ensure that any simulated data includes realistic correlation and dependence of measurements within a sample.
-However, this highlighted that one reason this is often neglected is that there is relatively little guidance for how to achieve it and that existing solutions are cumbersome.
+However, this highlighted that one reason dependence is often neglected is the relatively lack of guidance for how to achieve it and that existing solutions are cumbersome.
 
 Here, we give solutions to this problem by using a correlation matrix that decomposes into a diagonal part and a low-rank part to both approximate realistic dependencies in a real data set and generate simulated data mimicking the real data.
-Using a Gaussian copula approach (also referred to as NORTA, for "normal to anything"), the marginal (univariate) distributions can have realistic forms.
+Using a Gaussian copula approach (also referred to as NORTA, for "normal to anything" [@norta]), the marginal (univariate) distributions can have realistic forms.
 These solutions operate by taking a real dataset and mimicking it.
@@
 First, we simulate RNA-seq data with differential expression between two conditions.
-Using DESeq2 to determine the differentially expressed genes, we found that dependence had little impact on the accuracy of reported p-values but increased the variance of those estimates.
+Using DESeq2 to determine the differentially expressed genes, we found that dependence had little impact on the accuracy of reported $p$-values but increased the variance of those estimates.
 Second, we simulated a time series of RNA-seq data points and used the CYCLOPS method to infer collection time from the RNA-seq data, without time labels.
@@
 
-We chose as a references dataset, 12 mouse cortex RNA-seq samples from accession GSE151923 [@Wang2022].
+To compare the three simulation methods with a real data set, we chose as a references data set 12 mouse cortex RNA-seq samples from accession GSE151923 [@Wang2022].
 We then simulated data mimicking this reference using all three simulation methods (PCA, spiked Wishart, and corpcor) as well as a simulation with independent genes.
 We repeated the simulations, each of 12 samples, a total of 8 times to estimate variance.
-For the PCA method, we used $k=2$ and for the spiked Wishart, $k=11$.
+For the PCA method, we used $k=2$ dimensions and for the spiked Wishart, $k=11$.
 The coprcor method always uses the full data matrix, analogous to $k=11$.
@@
 (d) Projections onto the top two principal components of the real data set for both real and simulated data.
-All 8 simulations (96 samples total for each of dependent and independent simulations) shown.
+All 8 simulations (96 samples for each simulation) shown.
 (e) Principal component analysis was performed on all data sets and the variance captured by the top components is shown.
@@
 DESeq2 presents an interesting case because several aspects of it assume independence of genes and so may be adversely affected by gene-gene dependence.
-First, the independent filtering step [@Bourgon2010-uv] assumes independence but has generally be regarded to be robust to typical gene-gene dependence.
+First, the independent filtering step [@Bourgon2010-uv] assumes independence but has been reported to be robust to typical gene-gene dependence.
 Relatedly, the false discovery rate (FDR) [@BH_FDR] allows only certain forms of dependence.
@@
 We used a fly whole body RNA-Seq data set GSE81142 and selected samples of male flies without treatment and after at least 2 hours of feeding to simulate 5 "control" samples.
-We then randomly select 5% of the genes to be up or down regulated, with $\log_2$ fold change uniformly distributed between $0.2$ and $2.0$, and simulate 5 "experimental" samples.
-This was repeated 20 times for each of four dependence conditions (independent, PCA $k=2$, Wishart $k=11$, and corpcor).
+We then randomly select 5% of the genes to be up or down regulated, with $\log_2$ fold change uniformly distributed between $0.2$ and $2.0$, either up or down chosen randomly, and simulated 5 "experimental" samples.
+This was repeated 20 times for each of four dependence conditions (independent, PCA, Wishart, and corpcor).
 
-Finally, we ran DESeq2 on each simulated 5 vs 5 experiment and compared the output false discovery rate (FDR) with the true percentages of genes that are differential expressed (@fig-DESeq2 a-b).
+Finally, we ran DESeq2 on each simulated 5 vs 5 experiment and compared the output FDR with the true percentages of genes that are differential expressed (@fig-DESeq2 a-d).
 We observed that DESeq2 is anti-conservative on both data sets, with similar mean true FDRs for each estimated FDR cutoff.
@@
 To demonstrate the application of our simulation method for another organism, we also simulated data sets using mouse cortex data set GSE151923 [@Wang2022] and selected samples from male mice.
-We then simulated differential expression experiments as above and observed a similar result (@fig-DESeq2 c-d) as for the fly whole body data sets.
+We then simulated differential expression experiments as above and observed a similar result (@fig-DESeq2 e-f) as for the fly whole body data sets.
 
@@
 #| label: fig-DESeq2
-#| fig-cap: "Performance of DESeq2 on simulated datasets.(a-d) Comparison of true FDR and DESeq2 reported FDR for data sets simulated from the fly whole body data set (GSE81142), (a) without dependence, (b) using PCA, (c) using Wishart and (d) using corpcor.
+#| fig-cap: "Performance of DESeq2 on simulated datasets.
+(a-d) Comparison of true FDR and DESeq2 reported FDR for data sets simulated from the fly whole body data set (GSE81142), (a) without dependence, (b) using PCA, (c) using Wishart and (d) using corpcor.
 Diagonal line represents perfect estimation of FDR.
-(e-h) Comparison of true FDR and DESeq2 reported FDR for data sets simulated from the mouse cortex data set (GSE151923, Wang et al.
-2022), (e) without dependence, (f) using PCA, (g) using Wishart and (h) using corpcor."
+(e-h) Comparison of true FDR and DESeq2 reported FDR for data sets simulated from the mouse cortex data set (GSE151923), (e) without dependence, (f) using PCA, (g) using Wishart and (h) using corpcor."
 #| fig-width: 12
@@
 
-We further used our simulation method to benchmark CYCLOPS [@Anafi2017], which infers relative times for a set of unlabeled samples using an autoencoder to identify circular structures.
-We use a mouse liver time series data set (GSE151565, Aaronson et al.
-2020), which contains a total of 77 samples every 3 hours, for 36 hours.
+We next used our simulation method to benchmark CYCLOPS [@Anafi2017], which infers relative times for a set of unlabeled samples using an autoencoder to identify circular structures.
+We use a mouse liver time series data set GSE151565, which contains a total of 77 samples every 3 hours, for 36 hours.
 We compute the dependence structure of the genes as well as the variances of marginal distributions using the 12 time point 0 samples, and compute the means of gene expressions at each time point.
-We then use these to simulate 20 time series data sets with rank $2$ dependence structure, each with 8 time points and 6 samples per time point, and 20 such data sets without dependence.
+We then use these to simulate 20 time series data sets for each of the independent, PCA, Wishart, and corpcor simulation methods.
 
-We ran CYCLOPS on each data set with a list of cyclic mouse liver genes (JTK p-value $<0.05$), which yielded an estimated relative time for each sample.
-We evaluated CYCLOPS' performance compared to true circadian time using the circular correlation, defined as follows:
+We ran CYCLOPS on each data set with a list of cyclic mouse liver genes (from [@Zhang2014], JTK p-value $<0.05$), which yielded an estimated relative time for each sample.
+We evaluated CYCLOPS' performance compared to true circadian time using the circular correlation [@correlation_coeff], defined as follows:
 $$\rho = \frac{\sum_{1\leq i<j\leq n}\sin(X_i-X_j) \sin(Y_i-Y_j)}{(\sum_{1\leq i<j\leq n} \sin(X_i-X_j)^2)^{1/2}(\sum_{1\leq i<j\leq n} \sin(Y_i-Y_j)^2)^{1/2}},$$
@@
 Overall, CYCLOPS performance is poor in the independent data and improved in both PCA and corpcor data, but surprisingly the Wishart data also shows poor performance.
-Since by default CYCLOPS performs dimension reduction to the number of dimensions needed to capture 99% of the variance, this performance difference is likely due to the number of eigengenes needed by CYCLOPS to meet the variance threshold (@fig_CYCLOPS b).
+Since by default CYCLOPS performs dimension reduction to the number of dimensions needed to capture 99% of the variance, this performance difference is likely due to the number of eigengenes needed by CYCLOPS to meet the variance threshold (@fig-CYCLOPS b).
 
@@
 However, even this is computationally demanding; their paper references generating 20,000-dimensional vectors in “under an hour” using 16 threads.
-The `copula` package provides even more flexible dependence options through use of copulas (the NORTA approach is equivalent to using Gaussian copulas).
+The `copula` package provides even more flexible dependence options through use of copulas.
 All of these packages provide more flexibility in specifying dependence than our package, which can only mimic existing datasets, and therefore the longer run-times may be unavoidable for use cases where researchers need to parameterize the dependence structure.
+
+## Data availability
+
+Source code for all simulations and figures in this plot is available at [github.com/itmat/dependent_sim_paper/](https://github.com/itmat/dependent_sim_paper/).
+All data used is available from GSE151923, GSE81142, GSE151565.

### 2024-10-18 manuscript/circular_correlation.R

@@ -3,6 +3,7 @@ library(tibble)
 library(ggplot2)
 library(dplyr)
 library(tidyr)
+library(stringr)
 library(patchwork)
 
 # formula for the circular correlation
@@ -75,12 +76,11 @@ for (method in c("indep", "pca", "wishart", "corpcor")) {
 }
 
 # plot the number of eigengenes used by CYCLOPS
-eigen <- read.delim("num_eigengene.txt", header = FALSE, sep=" ") |> 
-    select(V5, V6) |> 
-    rename(num = V5, method = V6) |> 
-    mutate(method = sapply(strsplit(method,"/|\\\\"), '[', 4))
-eigen <- eigen[1:80,]
-eigen_violin <- ggplot(eigen, aes(method, num)) +
+eigen <- read_delim("num_eigengene.txt", col_names=c("num", "file")) |> 
+    mutate(method = str_extract(file, "./processed/cyclops/([a-z]+)", group=1))
+sim_eigen <- eigen |> filter(method != "real")
+real_eigens <- (eigen |> filter(method == "real"))$num
+eigen_violin <- ggplot(sim_eigen, aes(method, num)) +
     geom_violin() +
     ylab("number of eigengenes") +
     geom_hline(yintercept = 13, linetype = "dashed", color = "red")

### 2024-10-22 manuscript/compare_to_real.R
change colors

@@ -14,7 +14,7 @@ set.seed(55)
 # CONFIGURATION
 HIGH_EXPR_CUTOFF <- 300
 TYPE_ORDER <- c("real", "indep", "pca", "wishart", "corpcor")
-TYPE_COLORS <- RColorBrewer::brewer.pal(length(TYPE_ORDER), "Set2")
+TYPE_COLORS <- c("black", RColorBrewer::brewer.pal(length(TYPE_ORDER)-1, "Dark2"))
 
 # Load the real data ---------------------------------
 raw <- read_tsv("../data/GSE151923_metaReadCount_ensembl.txt.gz") |>
@@ -84,14 +84,14 @@ projected_corpcor <- predict(pca, t(scaled_draws_corpcor[!constant_rows, ]))
 projected_wishart <- predict(pca, t(scaled_draws_wishart[!constant_rows, ]))
 projected_indep_draws <- predict(pca, t(scaled_indep_draws[!constant_rows, ]))
 both_data <- data.frame(list(
-  PC1 = c(projected_read_data[,"PC1"], projected_draws[,"PC1"], projected_corpcor[,"PC1"], projected_wishart[,"PC1"], projected_indep_draws[,"PC1"]),
-  PC2 = c(projected_read_data[,"PC2"], projected_draws[,"PC2"], projected_corpcor[,"PC2"], projected_wishart[,"PC2"], projected_indep_draws[,"PC2"]),
+  PC1 = c(projected_read_data[1:12,"PC1"], projected_draws[1:12,"PC1"], projected_corpcor[1:12,"PC1"], projected_wishart[1:12,"PC1"], projected_indep_draws[1:12,"PC1"]),
+  PC2 = c(projected_read_data[1:12,"PC2"], projected_draws[1:12,"PC2"], projected_corpcor[1:12,"PC2"], projected_wishart[1:12,"PC2"], projected_indep_draws[1:12,"PC2"]),
   type = c(
-    rep("real", nrow(projected_read_data)),
-    rep("pca", nrow(projected_draws)),
-    rep("corpcor", nrow(projected_draws)),
-    rep("wishart", nrow(projected_draws)),
-    rep("indep", nrow(projected_indep_draws))
+    rep("real", 12),
+    rep("pca", 12),
+    rep("corpcor", 12),
+    rep("wishart", 12),
+    rep("indep", 12)
   )
 )) %>%
   mutate(type = factor(type, levels=TYPE_ORDER, ordered=TRUE))

### 2024-10-22 manuscript/compare_to_real.R
change colors

@@ -14,7 +14,7 @@ set.seed(55)
 # CONFIGURATION
 HIGH_EXPR_CUTOFF <- 300
 TYPE_ORDER <- c("real", "indep", "pca", "wishart", "corpcor")
-TYPE_COLORS <- RColorBrewer::brewer.pal(length(TYPE_ORDER), "Set2")
+TYPE_COLORS <- c("black", RColorBrewer::brewer.pal(length(TYPE_ORDER)-1, "Dark2"))
 
 # Load the real data ---------------------------------
 raw <- read_tsv("../data/GSE151923_metaReadCount_ensembl.txt.gz") |>
@@ -84,15 +84,15 @@ projected_corpcor <- predict(pca, t(scaled_draws_corpcor[!constant_rows, ]))
 projected_wishart <- predict(pca, t(scaled_draws_wishart[!constant_rows, ]))
 projected_indep_draws <- predict(pca, t(scaled_indep_draws[!constant_rows, ]))
 both_data <- data.frame(list(
-  PC1 = c(projected_read_data[,"PC1"], projected_draws[,"PC1"], projected_corpcor[,"PC1"], projected_wishart[,"PC1"], projected_indep_draws[,"PC1"]),
-  PC2 = c(projected_read_data[,"PC2"], projected_draws[,"PC2"], projected_corpcor[,"PC2"], projected_wishart[,"PC2"], projected_indep_draws[,"PC2"]),
-  type = c(
-    rep("real", nrow(projected_read_data)),
-    rep("pca", nrow(projected_draws)),
-    rep("corpcor", nrow(projected_draws)),
-    rep("wishart", nrow(projected_draws)),
-    rep("indep", nrow(projected_indep_draws))
-  )
+    PC1 = c(projected_read_data[1:12,"PC1"], projected_draws[1:12,"PC1"], projected_corpcor[1:12,"PC1"], projected_wishart[1:12,"PC1"], projected_indep_draws[1:12,"PC1"]),
+    PC2 = c(projected_read_data[1:12,"PC2"], projected_draws[1:12,"PC2"], projected_corpcor[1:12,"PC2"], projected_wishart[1:12,"PC2"], projected_indep_draws[1:12,"PC2"]),
+    type = c(
+        rep("real", 12),
+        rep("pca", 12),
+        rep("corpcor", 12),
+        rep("wishart", 12),
+        rep("indep", 12)
+    )
 )) %>%
[…]

### 2024-10-22 manuscript/paper.qmd
add colors to plots and minor wording edits to manuscript.

@@
 We used a fly whole body RNA-Seq data set GSE81142 and selected samples of male flies without treatment and after at least 2 hours of feeding to simulate 5 "control" samples.
-We then randomly select 5% of the genes to be up or down regulated, with $\log_2$ fold change uniformly distributed between $0.2$ and $2.0$, either up or down chosen randomly, and simulated 5 "experimental" samples.
+We then randomly selected 5% of the genes to be differentially expressed, with absolute $\log_2$ fold change uniformly distributed between $0.2$ and $2.0$, either up or down regulated chosen randomly, and simulated 5 "experimental" samples.
 This was repeated 20 times for each of four dependence conditions (independent, PCA, Wishart, and corpcor).
@@
 Finally, we ran DESeq2 on each simulated 5 vs 5 experiment and compared the output FDR with the true percentages of genes that are differential expressed (@fig-DESeq2 a-d).
-We observed that DESeq2 is anti-conservative on both data sets, with similar mean true FDRs for each estimated FDR cutoff.
+We observed that DESeq2 is anti-conservative on all data sets, with similar mean true FDRs for each estimated FDR cutoff.
 However, there was a greater variance in the performance of DESeq2 on the data sets simulated with dependence, indicating that it performs less consistently on data sets with gene-gene dependence, as in real data sets.
@@
 #| fig-cap: "Performance of DESeq2 on simulated datasets.
-(a-d) Comparison of true FDR and DESeq2 reported FDR for data sets simulated from the fly whole body data set (GSE81142), (a) without dependence, (b) using PCA, (c) using Wishart and (d) using corpcor.
+(a-d) Comparison of true false discovery proportions and DESeq2 reported False Discovery Rates, plotted on a log scale, for data sets simulated from the fly whole body data set (GSE81142), (a) without dependence, (b) using PCA, (c) using Wishart and (d) using corpcor.
 Diagonal line represents perfect estimation of FDR.
-(e-h) Comparison of true FDR and DESeq2 reported FDR for data sets simulated from the mouse cortex data set (GSE151923), (e) without dependence, (f) using PCA, (g) using Wishart and (h) using corpcor."
+(e-h) Comparison of true false discovery proportions and DESeq2 reported FDR for data sets simulated from the mouse cortex data set (GSE151923), (e) without dependence, (f) using PCA, (g) using Wishart and (h) using corpcor."
 #| fig-width: 12

### 2024-10-22 manuscript/DE_plots.R

@@ -6,13 +6,16 @@ library(dplyr)
 library(tidyr)
 library(patchwork)
 
+TYPE_ORDER <- c("indep", "pca", "wishart", "corpcor")
+TYPE_COLORS <- RColorBrewer::brewer.pal(4, "Dark2")
 mouse_plots = list()
-for (method in c("indep", "pca", "wishart", "corpcor")) {
+for (i in 1:4) {
+    method <- TYPE_ORDER[i]
     fdr_df <- read_csv(paste0("../processed/DE/Mouse.Cortex.Male.",method,".fdr.csv"))
     
     mouse_plots[[method]] <- ggplot(fdr_df, aes(cutoff, mean, ymin=lower, ymax=upper)) +
-        geom_ribbon(fill="grey") +
-        geom_point() +
+        geom_ribbon(fill=TYPE_COLORS[i], alpha=0.3) +
+        geom_point(color=TYPE_COLORS[i]) +
         geom_abline(slope=1, intercept = 0) +
         ggtitle(method) +
         xlab("reported FDR") +
@@ -32,12 +35,13 @@ for (method in c("indep", "pca", "wishart", "corpcor")) {
 #     geom_abline(slope=1, intercept = 0)
 
 fly_plots = list()
-for (method in c("indep", "pca", "wishart", "corpcor")) {
+for (i in 1:4) {
+    method <- TYPE_ORDER[i]
     fdr_df_fly <- read_csv(paste0("../processed/DE/Fly.WholeBody.Male.",method,".fdr.csv"))
     
     fly_plots[[method]] <- ggplot(fdr_df_fly, aes(cutoff, mean, ymin=lower, ymax=upper)) +
-        geom_ribbon(fill="grey") +
-        geom_point() +
+        geom_ribbon(fill=TYPE_COLORS[i], alpha=0.3) +
+        geom_point(color=TYPE_COLORS[i]) +
         geom_abline(slope=1, intercept = 0) +
         ggtitle(method) +
         xlab("reported FDR") +

### 2024-10-22 manuscript/circular_correlation.R

@@ -6,6 +6,9 @@ library(tidyr)
 library(stringr)
 library(patchwork)
 
+TYPE_ORDER <- c("indep", "pca", "wishart", "corpcor")
+TYPE_COLORS <- RColorBrewer::brewer.pal(4, "Dark2")
+
 # formula for the circular correlation
 get_cir_corr <- function(X, Y) {
     sum1 <- 0
@@ -49,9 +52,11 @@ for (method in c("indep", "pca", "wishart", "corpcor")) {
 
 # violin plot comparing circular correlation for simulated datasets with and without dependence
 corrs_new <- pivot_longer(corrs, cols = 1:4, names_to = "method")
-circular_correlation_violin <- ggplot(corrs_new, aes(method, abs(value))) +
+corrs_new$method <- factor(corrs_new$method, levels = TYPE_ORDER, ordered = TRUE)
+circular_correlation_violin <- ggplot(corrs_new, aes(method, abs(value), color=method)) +
     geom_violin() +
-    ylab("absolute circular correlation")
+    ylab("absolute circular correlation") +
+    scale_color_manual(values = TYPE_COLORS, breaks = TYPE_ORDER, name="method")
 circular_correlation_violin
 
 # rainbow plots visualizing true_counts_counts_counts and estimated phases
@@ -80,10 +85,12 @@ eigen <- read_delim("num_eigengene.txt", col_names=c("num", "file")) |>
     mutate(method = str_extract(file, "./processed/cyclops/([a-z]+)", group=1))
 sim_eigen <- eigen |> filter(method != "real")
 real_eigens <- (eigen |> filter(method == "real"))$num
-eigen_violin <- ggplot(sim_eigen, aes(method, num)) +
+sim_eigen$method <- factor(sim_eigen$method, levels = TYPE_ORDER, ordered = TRUE)
+eigen_violin <- ggplot(sim_eigen, aes(method, num, color=method)) +
     geom_violin() +
     ylab("number of eigengenes") +
-    geom_hline(yintercept = 13, linetype = "dashed", color = "red")
+    scale_color_manual(values = TYPE_COLORS, breaks = TYPE_ORDER, name="method") +
+    geom_hline(yintercept = 13, linetype = "dashed", color = "black")
 
 cyclops_plot <- circular_correlation_violin / 
     eigen_violin /

### 2024-10-25 manuscript/circular_correlation.R
Update cyclops plots

@@ -76,14 +76,28 @@ for (method in c("indep", "pca", "wishart", "corpcor")) {
 }
 
 # plot the number of eigengenes used by CYCLOPS
-eigen <- read_delim("num_eigengene.txt", col_names=c("num", "file")) |> 
-    mutate(method = str_extract(file, "./processed/cyclops/([a-z]+)", group=1))
-sim_eigen <- eigen |> filter(method != "real")
-real_eigens <- (eigen |> filter(method == "real"))$num
+eigengenes <- list()
+for (method in c("indep", "pca", "wishart", "corpcor")) {
+    m <- list()
+    for (batch in 0:19) {
+        n <- read_csv(paste0("../processed/cyclops/",method,"/batch=",batch,"/cyclops_EigengeneExp.csv"), show_col_types = FALSE) |>
+            nrow()
+        m[[length(m)+1]] <- n
+    }
+    eigengenes[[length(eigengenes)+1]] <- tibble(
+        method = method,
+        num = unlist(m),
+    )
+}
+sim_eigen <- bind_rows(eigengenes)
+print(sim_eigen)
+real_eigens <- read_csv("../processed/cyclops/real_data/cyclops_EigengeneExp.csv", show_col_types=FALSE) |>
+    nrow()
+
 eigen_violin <- ggplot(sim_eigen, aes(method, num)) +
     geom_violin() +
     ylab("number of eigengenes") +
-    geom_hline(yintercept = 13, linetype = "dashed", color = "red")
+    geom_hline(yintercept = real_eigens, linetype = "dashed", color = "black")
 
 cyclops_plot <- circular_correlation_violin / 
     eigen_violin /

### 2024-12-03 manuscript/paper.qmd
Rewrite to be more instructional

@@
 However, simulation too often ignores these correlations, perhaps due to the inconvenience and computational hurdles of doing so.
-We describe and implement three approaches based off decomposing the covariance matrix into a diagonal part and a low-rank part and using a Gaussian copula approach.
-These approaches are efficient at omics-scale problems.
-We demonstrate the importance of including correlation in two benchmarking applications.
+To alleviate this, we describe in detail three approaches for quickly generating omics-scale data with correlated measures which mimic real data sets.
+These approaches all are based on a Gaussian copula approach with a covariance matrix that decomposes into a diagonal part and a low-rank part.
+We use these approaches to demonstrate the importance of including correlation in two benchmarking applications.
 First, we show that variance of results from the popular DESeq2 method increases when dependence is included.
@@
 
-Omics data is in the $p \gg n$ regime where there are fewer samples than measurements per sample.
+Omics data typically has far fewer samples than measurements per sample.
 This creates dual challenges in generating realistic simulated data for the purposes of benchmarking.
 First, there isn’t enough data to be able to compute a dependence structure (e.g., a full-rank correlation matrix).
-Second, generating omics-scale data with a specified correlation matrix is slow due to the typical $O(p^3)$ nature of these algorithms.
+Second, generating omics-scale data with a specified correlation matrix is slow due to the typical $O(n^3)$ nature of these algorithms, where $n$ is the number of measurements per sample.
+Moreover, there is a lack of practical guidance on how to generate simulated data with realistic dependence.
 These often mean that simulators assume independence of the measurements, which does not reflect reality.
 
-Recently, we wrote guidelines for performing omics benchmarking [@Brooks2024] and encouraged authors to ensure that any simulated data includes realistic correlation and dependence of measurements within a sample.
-However, this highlighted that one reason dependence is often neglected is the relatively lack of guidance for how to achieve it and that existing solutions are cumbersome.
+Here, we give an introduction to the theory and practice of generating dependent data and describe three related solutions which all offer good performance and ease-of-use even for large omics-scale problems.
+This expands off a discussion we originally wrote as part of a larger discussion on best practices in omics benchmarking [@Brooks2024].
+Our goal here is to produce guidelines that show that generating correlated data does not have to be onerous and instead should be considered a baseline requirement when simulating data.
 
-Here, we give solutions to this problem by using a correlation matrix that decomposes into a diagonal part and a low-rank part to both approximate realistic dependencies in a real data set and generate simulated data mimicking the real data.
-Using a Gaussian copula approach (also referred to as NORTA, for "normal to anything" [@norta]), the marginal (univariate) distributions can have realistic forms.
+We present three methods that operate by inferring a covariance matrix that decomposes into a diagonal part and a low-rank part.
+Using a Gaussian copula [@Nelsen1998-nq] approach (also referred to as NORTA, for "normal to anything" [@norta]), the marginal (univariate) distributions can have realistic forms.
 These solutions operate by taking a real dataset and mimicking it.
-We implement this strategy as an R package which supports normal, Poisson, DESeq2-based (negative binomial with sample-specific size factors), and empirical (for ordinal data) marginal distributions.
-This makes it particularly suited to RNA-seq data but also widely applicable.
-Moreover, it supports multi-modal data such as for multi-omics.
+For ease of use, we implement this in an R package which supports normal, Poisson, DESeq2-based (negative binomial with sample-specific size factors), and empirical (for ordinal data) marginal distributions.
 
-We implement three different strategies for determining the diagonal and low-rank parts of the covariance matrix.
+We implemented three different strategies for determining the diagonal and low-rank parts of the covariance matrix.
 First, the 'PCA' method uses principal component analysis (PCA) and picks the low-rank part such that the simulated data has the same variance in the top $k$ PCA components of the reference dataset.
@@
 Unlike 'PCA', these variances are computed with resepect to the simulated data's own PCA and not using the PCA weights of the reference dataset.
-Third, the 'corpcor' method uses the popular R library `corpcor`, which implements a James-Stein type shrinkage estimator for the covariance matrix as a linear interpolation of the sample covariance matrix and a diagonal matrix.
+Third, the 'corpcor' method uses the popular R library `corpcor` [@SchaferStrimmer2005; @OpgenRheinStrimmer2007], which implements a James-Stein type shrinkage estimator for the covariance matrix as a linear interpolation of the sample covariance matrix and a diagonal matrix.
 No method exactly captures the input data, indicating room for future research, but all improve upon the common approach of assuming independence.
 
-We show two applications which demonstrate the importance of including dependence of measurements in simulated data when benchmarking computational pipelines.
+We show two applications which demonstrate the effects of including dependence of measurements in simulated data when benchmarking computational pipelines.
 First, we simulate RNA-seq data with differential expression between two conditions.
 Using DESeq2 to determine the differentially expressed genes, we found that dependence had little impact on the accuracy of reported $p$-values but increased the variance of those estimates.
-Second, we simulated a time series of RNA-seq data points and used the CYCLOPS method to infer collection time from the RNA-seq data, without time labels.
-Surprisingly, we found that including dependence in the data set improved the performance of CYCLOPS.
+Second, we simulated a time series of RNA-seq data points and used the CYCLOPS method [@Anafi2017] to infer collection time from the RNA-seq data, without time labels.
+In some cases, inclusion of dependence significantly improved performance of CYCLOPS.
 
@@
 
-### Comparison to Real Data
+Assume that we have a reference dataset $X$ given by an $p \times n$ data matrix of $p$ features measured in each of $n$ independent samples.
+We want to capture correlations between the $p$ features, which could represent gene expressions, protein abundances, or other measured values.
+We refer to these features as genes for simplicity.
+Our goal is to generate simulated data with the same $p$ genes, the same marginal distributions of each gene as in $X$ and realistic gene-gene dependence.
+
+### Multivariate normal distribution
+
+We first discuss the simplest case, where our data set is multivariate normally distributed.
+The distribution $N(\mu, \Sigma)$ is the multivariate normal distribution with mean vector $\mu$ and covariance matrix $\Sigma$.
+Here $\mu$ is a $p \times 1$ column vector and $\Sigma$ is a $p \times p$ matrix.
+In order for this to work, $\Sigma$ must be symmetric and positive semi-definite, meaning that all of its eigenvalues are non-negative 
+This matrix is conceptually simple, since $\Sigma_{ij}$ gives the the covariance of $x_i$ and $x_j$ when $x \sym N(\mu, \Sigma)$.
+More specifically, this is the population covariance matrix of $N(\mu, \Sigma)$, which does not generally equal the sample covariance matrix.
+Indeed, if $X$ is $n$ samples from $N(\mu, \Sigma)$, then $\hat \Sigma := (X - \bar X) (X - \bar X)^T / (n - 1)$ is the sample covariance matrix where $\bar X$ is the average of the $n$ samples.
+For large $n$, $\hat \Sigma$ will closely approximate $\Sigma$, but we care primarily about the situation where $n$ is small.
+In particular, $\hat \Sigma$ is at most a rank $n - 1$ matrix while $\Sigma$ could be up to rank $p$.
+This means that $N(\mu, \hat \Sigma)$ and $N(\mu, \Sigma)$ are quite different distributions: every sample from $N(\mu, \hat Sigma)$ is contained in an $n-1$ dimensional plane.
+If $n$ is small, then this is very unlike real data, which typically is close to $p$ dimensional.
+
+The difficulty then is that we only know $\hat \Sigma$ from our reference data set, but we need to choose a $\Sigma$ with which we can simulate data and the obvious choice of $\hat \Sigma$ is inadequate.
+The most common choice is to assume $\Sigma$ is a diagonal matrix.
+This is the situation we want to avoid where the generated data is independent: $x_i$ and $x_j$ have zero covariance unless $i = j$.
+However, this has some nice properties, such as being simple and fast to simulate (just generate univariate normal data for each variable).
+
+We describe three alternative approaches in Methods of how to choose a $\Sigma$ that will produce data similar to the input data.
+All three of these rely on a specific form of $\Sigma$, namely that
+$$ \Sigma = D^2 + PP^T $$
+where $D$ is a $p \times p$ diagonal matrix and $P$ is $p \times k$ for some $k \ll p$.
+This is a combination of an independent part (the diagonal matrix) and a low-rank part ($P P^T$).
+The low-rank part is also simple to generate data for.
+If $x$ is a vector of $k$ independent univariate standard normal values, then $P x \sim N(0, P P^T)$.
+The choice of specific $D$ and $P$ matrices is more in-depth and we leave the details for the Methods section.
+We have three alternative means for doing so, which we refer to as the PCA, spiked Wishart, and corpcor methods.
+
+Lastly, we emphasize that multivariate normal distributions do not capture all, or even most, types of possible dependence.
+Indeed, we see this even in the 2-dimensional case where it is well known that correlation describes only a linear relationship between two variables while in reality they may have much more complex relations.
+In higher dimensions, the problem is only worse.
+So any method based off multivariate normal distributions are making large assumptions about distribution.
+However, it is necessary to make some assumption like this.
+In the next section, though, we see that "normal" part is actually not a large obstacle.
+
+### Gaussian copula
+
+Building on the multivariate normal distribution, a popular approach to describe dependence in a high-dimensional settings is called the Gaussian copula approach.
+The idea of this approach is that by normalizing and later reversing the normalization, data that does not fit a normal distribution can still have its dependence structure described using a multivariate normal distribution.
+This allows the marginal (i.e., univariate) distributions of each genes to be specified separately from the dependence between genes.
+This operates first by normalizing each gene by fitting a distribution (such as a normal distribution, Poisson, negative binomial, or other form), and then applying the fit cumulative distribution function (CDF) to the observed values.
+Finally, those are fed to a standard normal distribution's inverse CDF to obtain values that are approximately normally distributed.
+These values are then used to compute a covariance matrix $\Sigma$ and the data is assumed to follow a multivariate normal distribution in $p$ dimensions with that covariance matrix.
+
+Here, we describe the approach using the form of covariance matrix $\Sigma = D^2 + PP^T$ as above.
+Once data is obtained $Z \sim N(0, \Sigma)$, then one can undo the normalization process to obtain data with the same marginal distributions as the fit marginal distributions but with dependence determined by $\Sigma$.
+We describe this in detail:
+
+1.  Fit marginal distributions to each feature in $X$ to determine CDFs $F_{i}$ for each feature.
+2.  Transform $X$ to normalized values by $Z_{ij} = \Phi^{-1}(F_{i}(X_{ij}))$ where $\Phi$ is the CDF of the standard normal distribution.
+3.  Compute $D$, $U$, $W$ matrices from $X$ by one of three methods (see Methods).
+4.  Generate $k$ i.i.d. standard normally distributed values $u$ and $p$ i.i.d standard normally distributed values $v$.
+5.  Set $Z' = UWu + D v$.
+6.  Output the vector $X'$ where $X'_i = F_i^{-1}(\Phi(Z'))$.
+
+The generated data $Z'$ has covariance matrix $\Sigma = D^2 + U W U^T = D^2 + P P^T$, where $P := U \sqrt{W}$.
+Moreover, we require that $\Sigma$ satisfies that $e_i \Sigma e_i^T$ is approximately 1.
+That guarantees that the output $X'$ has each entry with the same marginal distributions $F_i$ as was originally fit and inherits gene-gene dependence from $Z'$.
+This method is computationally efficient, taking hardly any more time or memory than simulations without dependence.
+
+### Comparison to real data
 
@@
 
-### DESeq2 Application
+### DESeq2 application
 
@@
 
-### CYCLOPS Application
+### CYCLOPS application
 
@@
 
-Assume that we have a reference dataset $X$ given by an $p \times n$ data matrix of $p$ features measured in each of $n$ independent samples.
-We want to capture correlations between the $p$ features, which could represent gene expressions, protein abundances, or other measured values.
-We refer to these features as genes for simplicity.
-Our goal is to generate simulated data with the same $p$ genes, the same marginal distributions of each gene as in $X$ and realistic gene-gene dependence.
+Below, we describe the three methods for selecting the components of the covariance matrix $\Sigma := D + P P^T$.
 
-### Gaussian Copula
-
-All three methods are based off a well-known Gaussian copula approach.
-This operates first by normalizing each gene by fitting a distribution (such as a normal distribution, Poisson, negative binomial, or other form), and then applying the fit cumulative distribution function (CDF) to the observed values.
-Finally, those are fed to a standard normal distribution's inverse CDF to obtain values that are approximately normally distributed.
-These values are then used to compute a covariance matrix $\Sigma$ and the data is assumed to follow a multivariate normal distribution in $p$ dimensions with that covariance matrix.
-Note that the sample covariance matrix is a poor choice for this $\Sigma$.
-In general, the sample covariance matrix is at most rank $n$, and will be approximately rank $n-1$ due to the normalization procedure.
-This means that generated data would lie on an $n-1$ dimensional hyperplane, which we know is unrealistic.
-
-Instead, we want to estimate some other matrix of the form $\Sigma = D + PP^T$ where $D$ is $p \times p$ diagonal and $P$ is $p \times k$ for $k \ll p$.
-This means that $P P^T$ is low-rank.
-Conveniently, this form allows efficient simulation of multivariate normal data $N(0, \Sigma)$.
-Once data is obtained $Z \sim N(0, \Sigma)$, then one can undo the normalization process to obtain data with the same marginal distributions as the fit marginal distributions but with dependence determined by $\Sigma$.
-We describe this in detail below:
-
-1.  Fit marginal distributions to each feature in $X$ to determine CDFs $F_{i}$ for each feature.
-2.  Transform $X$ to normalized values by $Z_{ij} = \Phi^{-1}(F_{i}(X_{ij}))$ where $\Phi$ is the CDF of the standard normal distribution.
-3.  Compute $D$, $U$, $W$ matrices from $X$ by one of three methods (below).
-4.  Generate $k$ i.i.d. standard normally distributed values $u$ and $p$ i.i.d standard normally distributed values $v$.
-5.  Set $Z' = UWu + D v$.
-6.  Output the vector $X'$ where $X'_i = F_i^{-1}(\Phi(Z'))$.
-
-The generated data $Z'$ has covariance matrix $\Sigma = D^2 + U W U^T$.
-Moreover, we require that $\Sigma$ satisfies that $e_i \Sigma e_i^T$ is approximately 1.
-That guarantees that the output $X'$ has each entry with the same marginal distributions $F_i$ as was originally fit and inherits gene-gene dependence from $Z'$.
-
-This is a standard Gaussian copula approach tailored to our specific format of $\Sigma$.
-Most implementations of this approach instead rely upon performing a Cholesky decomposition of $\Sigma$, which requires significant time and memory resources for omics-scale problems with over tens of thousands of genes.
-
-###  PCA Method
+###  PCA method
 
@@
 
-### Spiked Wishart Method
+### Spiked Wishart method
 
@@
 
-### Corpcor Method
+### Corpcor method
 
@@
 
-We present `dependentsimr`, an R package that generates omics-scale data with realistic correlation.
-We demonstrate this package on RNA-seq data, using the DESeq2 method to fit negative binomial marginal distributions.
+We described the well-known Gaussian copula approach and recommended a specific form of covariance matrix which is well tailored to omics data simulation.
+We developed three methods using this form of covariance matrix which can be used to mimic a reference data set for simulation.
+All of these methods use a multivariate normal distribution as an intermediate step and therefore substantially restrict the kinds of dependence that can be simulated.
+However, when operating in a high-dimensional space some simplification is likely required.
+
+To encourage adoption of dependence in simulated omics data, we developed `dependentsimr`, an R package that generates omics-scale data with realistic correlation.
+This implementation is efficient and simple, requiring just two lines of code to fit a model to a reference data set and then simulate data from it.
+We demonstrated this package on RNA-seq data, using the DESeq2 method to fit negative binomial marginal distributions.
 However, this package is actually quite general and supports normal, Poisson, negative binomial, and arbitrary ordered discrete distributions using the empirical CDF.
@@
 
-We demonstrate the importance of including gene-gene dependence in simulated data by two application benchmarks.
-In the first, we demonstrate that DESeq2 results are substantially more variable when simulating with gene-gene dependence.
-In the second, we demonstrate that the performance of CYCLOPS in estimating circadian phases depends upon gene-gene dependence, with an unexpected result of better performance in the dependent case.
+We demonstrated the importance of including gene-gene dependence in simulated data by two application benchmarks.
+In the first, DESeq2 results were substantially more variable when simulating with gene-gene dependence.
+In the second, CYCLOPS performance in estimating circadian phases depends upon gene-gene dependence, was sensitive to dependence structure of the data.
 
-Our comparisons to a real dataset show that none of our three methods are able to exactly capture some aspects of the real dataset.
+Our comparisons to a real dataset show that none of our three methods are able to exactly capture all aspects of the real dataset.
 In particular, the gene-gene correlations were too high in the PCA method and too low in the spiked Wishart and corpcor methods.
-Surprisingly, spiked Wishart and corpcor methods improved in this metric only slightly compared to the simulations with independent genes.
-These observations demonstrate that there is room for future improvements over independent data in these techniques and possibly demonstrate that the datasets are too far from a multivariate normal distribution even after normalization.
-Nonetheless, these methods represent significant improvements by other metrics and we recommend the inclusion of some dependence in nearly every simulated dataset.
-We highlight some alternatives that have been taken below.
+Surprisingly, the spiked Wishart and corpcor methods improved in this metric only slightly compared to the simulations with independent genes.
+These observations demonstrate that there is room for future improvements over independent data in these techniques, possibly incorporating more recent developments in copulae [@copulae@].
+Possibly, this could demonstrate the limitations of methods based on the multivariate normal distribution or of the low-rank approximation used by all three of our methods. 
+Nonetheless, these methods represent significant improvements by other metrics and we recommend the inclusion of some dependence in nearly every simulated omics dataset.
 
 ### Alternatives
+
+We highlight some alternative approaches and software packages that have been taken below.
 
@@
 Source code for all simulations and figures in this plot is available at [github.com/itmat/dependent_sim_paper/](https://github.com/itmat/dependent_sim_paper/).
-All data used is available from GSE151923, GSE81142, GSE151565.
+All data used is available with accession numbers GSE151923, GSE81142, GSE151565.

### 2024-12-13 manuscript/paper.qmd
Add dependentsimr package reference
Changed CYCLOPS data to mouse cortex, batch size 4 and added simulate_data notebook.

@@
 Here $\mu$ is a $p \times 1$ column vector and $\Sigma$ is a $p \times p$ matrix.
-In order for this to work, $\Sigma$ must be symmetric and positive semi-definite, meaning that all of its eigenvalues are non-negative 
-This matrix is conceptually simple, since $\Sigma_{ij}$ gives the the covariance of $x_i$ and $x_j$ when $x \sym N(\mu, \Sigma)$.
+In order for this to work, $\Sigma$ must be symmetric and positive semi-definite, meaning that all of its eigenvalues are non-negative. 
+This matrix is conceptually simple, since $\Sigma_{ij}$ gives the the covariance of $x_i$ and $x_j$ when $x \sim N(\mu, \Sigma)$.
 More specifically, this is the population covariance matrix of $N(\mu, \Sigma)$, which does not generally equal the sample covariance matrix.
@@
 In particular, $\hat \Sigma$ is at most a rank $n - 1$ matrix while $\Sigma$ could be up to rank $p$.
-This means that $N(\mu, \hat \Sigma)$ and $N(\mu, \Sigma)$ are quite different distributions: every sample from $N(\mu, \hat Sigma)$ is contained in an $n-1$ dimensional plane.
+This means that $N(\mu, \hat \Sigma)$ and $N(\mu, \Sigma)$ are quite different distributions: every sample from $N(\mu, \hat \Sigma)$ is contained in an $n-1$ dimensional plane.
 If $n$ is small, then this is very unlike real data, which typically is close to $p$ dimensional.
@@
 We next used our simulation method to benchmark CYCLOPS [@Anafi2017], which infers relative times for a set of unlabeled samples using an autoencoder to identify circular structures.
-We use a mouse liver time series data set GSE151565, which contains a total of 77 samples every 3 hours, for 36 hours.
+We use a mouse cortex time series data set GSE151565, which contains a total of 77 samples every 3 hours, for 36 hours.
 We compute the dependence structure of the genes as well as the variances of marginal distributions using the 12 time point 0 samples, and compute the means of gene expressions at each time point.
@@
 
-We ran CYCLOPS on each data set with a list of cyclic mouse liver genes (from [@Zhang2014], JTK p-value $<0.05$), which yielded an estimated relative time for each sample.
+We ran CYCLOPS on each data set with a list of cyclic mouse genes (from [@Zhang2014], JTK p-value $<0.05$), which yielded an estimated relative time for each sample.
 We evaluated CYCLOPS' performance compared to true circadian time using the circular correlation [@correlation_coeff], defined as follows:
@@
 We find that CYCLOPS performance depends significantly on the dependence structure of the simulated time series data (@fig-CYCLOPS), indicating that the presence of gene-gene dependence is highly impactful on CYCLOPS' performance.
-Overall, CYCLOPS performance is poor in the independent data and improved in both PCA and corpcor data, but surprisingly the Wishart data also shows poor performance.
-Since by default CYCLOPS performs dimension reduction to the number of dimensions needed to capture 99% of the variance, this performance difference is likely due to the number of eigengenes needed by CYCLOPS to meet the variance threshold (@fig-CYCLOPS b).
+Overall, CYCLOPS performance is worse in the corpcor and Wishart data than in the independent data, but surprisingly the pca data shows better performance.
+Since by default CYCLOPS performs dimension reduction to the number of dimensions needed to capture 80% of the variance, this performance difference is likely partially due to the number of eigengenes needed by CYCLOPS to meet the variance threshold (@fig-CYCLOPS b).
 
@@
 #| label: fig-CYCLOPS
-#| fig-cap: "Performance of CYCLOPS on simulated time series data sets based on mouse liver data set (GSE151565).
+#| fig-cap: "Performance of CYCLOPS on simulated time series data sets based on mouse cortex data set (GSE151565).
 (a) Violin plot comparing absolute circular correlations between true phases and CYCLOPS estimated phases on the simulated data sets with the four methods.
-(b) Violin plot comparing the number of eigengenes used by CYCLOPS between the four methods; the red dotted line indicates the number of eigengenes used by CYCLOPS on the real data (13).
+(b) Violin plot comparing the number of eigengenes used by CYCLOPS between the four methods; the red dotted line indicates the number of eigengenes used by CYCLOPS on the real data (5).
 (c-f) Examples of CYCLOPS estimated phases on the simulated data sets.
@@
 Source code for all simulations and figures in this plot is available at [github.com/itmat/dependent_sim_paper/](https://github.com/itmat/dependent_sim_paper/).
+Source code for the `dependentsimr` package is available at [github.com/tgbrooks/dependent_sim](https://github.com/tgbrooks/dependent_sim/).
 All data used is available with accession numbers GSE151923, GSE81142, GSE151565.

### 2024-12-13 manuscript/circular_correlation.R

@@ -40,7 +40,7 @@ corrs <- tibble(indep = rep(0, 20), pca = rep(0, 20), wishart = rep(0, 20), corp
 for (method in c("indep", "pca", "wishart", "corpcor")) {
     for (j in (0:19)) {
         phaselist <- read_csv(paste0("../processed/cyclops/",method,"/batch=", j, "/cyclops_estimated_phaselist.csv"))
-        for (i in (1:48)) {
+        for (i in (1:32)) {
             phaselist[i, 'ID'] <- strsplit(as.character(phaselist[i, 'ID']), "_")[[1]][1] |> 
                 substring(3)
         }
@@ -62,8 +62,8 @@ circular_correlation_violin
 # rainbow plots visualizing true_counts_counts_counts and estimated phases
 phaseplots <- list()
 for (method in c("indep", "pca", "wishart", "corpcor")) {
-    phaselist <- read_csv(paste0("../processed/cyclops/",method,"/batch=", 8, "/cyclops_estimated_phaselist.csv"))
-    for (i in (1:48)) {
+    phaselist <- read_csv(paste0("../processed/cyclops/",method,"/batch=", 5, "/cyclops_estimated_phaselist.csv"))
+    for (i in (1:32)) {
         phaselist[i, 'ID'] <- strsplit(as.character(phaselist[i, 'ID']), "_")[[1]][1] |> 
             substring(3)
     }

### 2024-12-13 scripts/process_time_series.R

@@ -4,7 +4,7 @@ library(dplyr)
 library(tidyr)
 library(tidyverse)
 
-raw_data <- read_csv("data/GSE151565_Liver-counts.csv.gz")
+raw_data <- read_csv("data/GSE151565_Cortex-counts.csv.gz")
 
 normalized_data <- raw_data
 
@@ -46,12 +46,12 @@ mean_df <- mean_per_time |>
     purrr::reduce(full_join, by="normalized_data$EnsemblID")
 colnames(mean_df) <- c("EnsemblID", 'ZT0', 'ZT3', 'ZT6', 'ZT9',
                               'ZT12', 'ZT15', 'ZT18', 'ZT21')
-write_csv(mean_df, "processed/mean_per_time_Liver.csv")
+write_csv(mean_df, "processed/mean_per_time_Cortex.csv")
 
 processed_data <- data_per_time |> 
     reduce(full_join, by="normalized_data$EnsemblID")
-write_csv(processed_data, "processed/Liver_normalized_time_series_data.csv")
+write_csv(processed_data, "processed/Cortex_normalized_time_series_data.csv")
 
 data0 <- cbind(raw_data$EnsemblID, raw_data[, time_points==paste0('ZT',0)])
 data0 <- data0[-1, ]
-write_csv(data0, "processed/Liver_ZT0-counts.csv")
+write_csv(data0, "processed/Cortex_ZT0-counts.csv")

### 2024-12-13 scripts/simDE.R

@@ -68,6 +68,7 @@ simDE_analysis <- function(filename, samples) {
         rename(ENSEMBL_ID = row, qvalue = padj) |>
         mutate(Comparison = "case_vs_control") |> 
         arrange(pvalue, qvalue)
+    
         
     # create list of differentially expressed genes
     DEgenes <- 
@@ -84,6 +85,7 @@ simDE_analysis <- function(filename, samples) {
     filtered_results <- 
         deseq2_results |> 
         filter(qvalue < 0.1)
+    write_csv(filtered_results, paste0("processed/DE/", filename, "_DEG.csv"))
     
     # read true values for the case and control datasets
     case_true_vals <-
@@ -131,6 +133,9 @@ simDE_analysis <- function(filename, samples) {
 
 }
 
+## test GO analysis
+pca_DEG <- read_csv("processed/DE/Mouse.Cortex.Male.pca_DEG.csv")
+pca_DEG$ENSEMBL_ID
 
 # run DE analysis on the mouse data set for all four methods and get fdr values
 for (method in c("indep", "pca", "wishart", "corpcor")) {

### 2024-12-13 scripts/simulate_time_data.R

@@ -8,7 +8,7 @@ library(rlang)
 method <- snakemake@wildcards$method
 
 # use ZT0 of raw time series data to get dependence structure
-male_read_counts <- read.csv("processed/Liver_ZT0-counts.csv") |> 
+male_read_counts <- read.csv("processed/Cortex_ZT0-counts.csv") |> 
     column_to_rownames(var="raw_data.EnsemblID")
 
 # Get dependence structure
@@ -54,7 +54,7 @@ generate <- function(rs, seed) {
 male_sim <- generate(male_rs, .Random.seed)
 
 # read time point means
-time_point_means <- read.csv("processed/mean_per_time_Liver.csv") |> 
+time_point_means <- read.csv("processed/mean_per_time_Cortex.csv") |> 
     inner_join(tibble(EnsemblID = male_rs$rownames$data), by='EnsemblID')
 time_point_means$mean <- time_point_means[, -1] |> 
     mutate_all(as.numeric) |> 
@@ -82,11 +82,11 @@ for (i in (2 : (ncol(time_point_means)-1))) {
 
 time_data <- data_per_time |> 
     purrr::reduce(full_join, by="ENSEMBL_ID")
-write.csv(time_data, paste0("simulated_data/Liver_120_simulated_time_series_",method,".csv"), row.names = FALSE)
+write.csv(time_data, paste0("simulated_data/Cortex_120_simulated_time_series_",method,".csv"), row.names = FALSE)
 norm_time_data <- time_data
 read_depth <- apply(time_data[, -1], 2, sum)
 norm_time_data[, -1] <- as_tibble(as.matrix(time_data[, -1]) %*% diag(1000000/read_depth))
-write.csv(norm_time_data, paste0("simulated_data/Liver_120_normalized_simulated_time_series_",method,".csv"), row.names = FALSE)
+write.csv(norm_time_data, paste0("simulated_data/Cortex_120_normalized_simulated_time_series_",method,".csv"), row.names = FALSE)
 
 
 

### 2024-12-13 scripts/simulate_data.Rmd
Added notebook.

## Setup

Install the required packages, and choose a simulation method (one of "indep", "pca", "corpcor" and "wishart").

```{r}
library(dependentsimr)
library(dplyr)
library(tibble)
library(DESeq2)

method <- 'pca'
```

Read in metadata containing sample names, sample IDs and sex.
```{r}
metadata <- read.delim("/home/jianingy/dependent_sim/dependent_sim_paper/processed/GSE151923_sample_metadata.txt", sep="\t")
```

Read in the data containing read counts of genes for each sample.
```{r}
read_counts <- read.delim("/home/jianingy/dependent_sim/dependent_sim_paper/data/GSE151923_metaReadCount_ensembl.txt.gz", sep="\t")
```

Set rownames to be the gene IDs.
```{r}
row.names(read_counts) <- read_counts[,1]

read_counts <- read_counts[,6:ncol(read_counts)]
```

Drop NAs - the data file is missing part of the very last row, giving NAs.
```{r}
read_counts <- read_counts[!apply(is.na(read_counts), 1, any),]
```

Drop a sample that has very low read counts.
```{r}
read_depth <- apply(read_counts, 2, sum)

READ_DEPTH_CUTOFF = 10e6
read_counts <- read_counts[,read_depth > READ_DEPTH_CUTOFF]
```

Separate by sex. We will only use data from male mice in this example.
```{r}
sex <- vapply(colnames(read_counts),
    function(id) {
        return(metadata[metadata["sample_id"] == id, "sex"])
    }, "")

male_read_counts <- read_counts[,sex == 'male']
```

## Get Dependence Structure

For the given simulation method, get the dependence structure from the real data set.

```{r}

# Produce simulated data
if (method == "pca") {
    set.seed(1)
    # Simulate with the PCA method
    rs <- get_random_structure(list(data=male_read_counts), method="pca", rank=2, types="DESeq2")
} else if (method == "corpcor") {
    set.seed(2)
    # Simulate with the corpcor method
    rs <- get_random_structure(list(data=male_read_counts), method="corpcor", types="DESeq2")
} else if (method == "wishart") {
    set.seed(3)
    # Simulate with the spiked Wishart method
    rs <- get_random_structure(list(data=male_read_counts), rank=11, method="spiked Wishart", types="DESeq2")
} else if (method == "indep") {
    set.seed(4)
    # Simulate without any dependence
    dep_rs <- get_random_structure(list(data=male_read_counts), method="pca", rank=2, type="DESeq2")
    rs <- remove_dependence(dep_rs)
}
```

Get the read depth of samples in the real data set, and generate library sizes of the simulated data sets (5 data sets each with 20 samples).
```{r}
actual_library_sizes <- male_read_counts |> apply(2, sum)

N_SAMPLES <- 100
library_sizes <- sample(actual_library_sizes / mean(actual_library_sizes), size=N_SAMPLES, replace=TRUE)
```

## Simulate Control Data

Generate and save data+metadata (true values) from a specific random structure
```{r}
generate <- function(rs, name) {
  sim <- draw_from_multivariate_corr(rs, n_samples=N_SAMPLES, size_factors=library_sizes)$data
  colnames(sim) <- paste("sample", 1:ncol(sim), sep='')

sim |>
    data.frame() |>
    rownames_to_column("ENSEMBL_ID") |>
    write.table(file=paste0("/home/jianingy/dependent_sim/dependent_sim_paper/notebook_data/", name, ".txt"), sep="\t", row.names=FALSE)

  # metadata
  truth <- data.frame(
    gene_id = rownames(sim),
    mean = rs$marginals$data$q,
    disperion = rs$marginals$data$dispersion
  )
  truth[is.na(truth$mean), 2] <- 0 # NA mean is 0

  return(sim)
}

male_sim <- generate(rs, name=paste0("Mouse.Cortex.Male.",method,".Control"))
```

## Simulate Differentially Expressed Data

Randomly select some genes to be DE.
```{r}
FRAC_DE <- 0.05
LFC_MIN <- 0.2
LFC_MAX <- 2.0
N_DE <- floor(FRAC_DE * nrow(male_sim))

de_genes <- sample(nrow(male_sim), N_DE)
```

Simulate log fold changes for the case data sets.
```{r}
de_lfc <- runif(N_DE, LFC_MIN, LFC_MAX) * sample(c(-1, 1), size=N_DE, replace=TRUE)
```

Generate Cases data sets.
```{r}
if (method == "indep") {
    male_de_rs <- dep_rs
    male_de_rs$marginals$data$q[de_genes] <- 2^(de_lfc) * male_de_rs$marginals$data$q[de_genes]

    male_indep_de_rs <- remove_dependence(male_de_rs)
    male_indep_de_sim <- generate(male_indep_de_rs, name="Mouse.Cortex.Male.indep.Case")
} else {
    male_de_rs <- rs
    male_de_rs$marginals$data$q[de_genes] <- 2^(de_lfc) * male_de_rs$marginals$data$q[de_genes]

    male_de_sim <- generate(male_de_rs, name=paste0("Mouse.Cortex.Male.",method,".Case"))
}
```

### 2025-01-22 scripts/simDE.R
Cleanup unused GO analysis

@@ -133,10 +133,6 @@ simDE_analysis <- function(filename, samples) {
 
 }
 
-## test GO analysis
-pca_DEG <- read_csv("processed/DE/Mouse.Cortex.Male.pca_DEG.csv")
-pca_DEG$ENSEMBL_ID
-
 # run DE analysis on the mouse data set for all four methods and get fdr values
 for (method in c("indep", "pca", "wishart", "corpcor")) {
     fdr_lists = list()

### 2025-01-22 manuscript/paper.qmd
Make D^2 consistent, add funding sources

@@
 ---
-
 ## Abstract
@@
 
-Below, we describe the three methods for selecting the components of the covariance matrix $\Sigma := D + P P^T$.
+Below, we describe the three methods for selecting the components of the covariance matrix $\Sigma := D^2 + P P^T$.
 
@@
 Therefore, the distribution of singular values is not affected by the assumption that $\Sigma$ is diagonal.
-Moreover, for the form $\Sigma = D + U W U^T$ where $p$ is very large, each column of $U$ is typically very close to orthogonal to any $e_i$, a standard basis vector.
-Therefore, when $D = cI$ for some constant $c$, we can approximate $\Sigma$ as having $k$ arbitrary eigenvalues from $W$ and $n$ remaining eigenvalues all equal to $c$ corresponding to $D$.
+Moreover, for the form $\Sigma = D^2 + U W U^T$ where $p$ is very large, each column of $U$ is typically very close to orthogonal to any $e_i$, a standard basis vector.
+Therefore, when $D = c^2I$ for some constant $c$, we can approximate $\Sigma$ as having $k$ arbitrary eigenvalues from $W$ and $n$ remaining eigenvalues all equal to $c$ corresponding to $D$.
 This is a spiked Wishart distribution.
@@
 1. Set $U$ to be the $p \times k$ matrix with columns $u_i$, the left singular vectors of $Z$.
-2. Compute $w_1, \ldots, w_k$ and $c$ by stochastic gradient descent minimizing $\sum_{i} (E[\lambda'_i] - \lambda_i)^2$ for $\Sigma$ diagonal with entries $w_1, \ldots, w_k, c, \ldots, c$.
+2.
+Compute $w_1, \ldots, w_k$ and $c$ by stochastic gradient descent minimizing $\sum_{i} (E[\lambda'_i] - \lambda_i)^2$ for $\Sigma$ diagonal with entries $w_1, \ldots, w_k, c^2, \ldots, c^2$ [@wishart].
 4. Set $W$ diagonal with the entries $w_1, \ldots, w_k$.
@@
 It then uses $\lambda_1$ to linearly interpolate the sample covariance matrix towards the identity matrix $I$ and $\lambda_2$ to interpolate the vector of variances towards the median variance value.
-Since the sample covariance matrix is rank at most $n$, we again obtain a matrix of the form $\Sigma = D + UWU^T$.
+Since the sample covariance matrix is rank at most $n$, we again obtain a matrix of the form $\Sigma = D^2 + UWU^T$.
 
@@
 All data used is available with accession numbers GSE151923, GSE81142, GSE151565.
+
+## Funding statement and competing interests
+
+TB and GG received funding support from the National Center for Advancing Translational Sciences Grant (5UL1TR000003).
+The funders had no role in this research, the decision to publish, or the preparation of this manuscript.
+
+The authors declare no competing interests.

### 2025-01-23 manuscript/paper.qmd
Update plots, change CYCLOPS, edit manuscript

@@
 First, we show that variance of results from the popular DESeq2 method increases when dependence is included.
-Second, we demonstrate that a method, CYCLOPS, for inferring circadian time of collection from transcriptomics performs substantially worse or better under different dependence situations.
+Second, we demonstrate that CYCLOPS, a method for inferring circadian time of collection from transcriptomics, improves in performance when given gene-gene dependencies in some circumstances.
 We provide an R package, dependentsimr, that has efficient implementations of these methods and can generate dependent data with arbitrary distributions, including discrete (binary, ordered categorical, Poisson, negative binomial), continuous (normal), or with an empirical distribution.
@@
 Second, we simulated a time series of RNA-seq data points and used the CYCLOPS method [@Anafi2017] to infer collection time from the RNA-seq data, without time labels.
-In some cases, inclusion of dependence significantly improved performance of CYCLOPS.
+Depending upon settings used, performance of CYCLOPS dependent substantially on the dependence structure of the data, and surprisingly showed worst performance when given data with independent genes.
 
@@
 We next used our simulation method to benchmark CYCLOPS [@Anafi2017], which infers relative times for a set of unlabeled samples using an autoencoder to identify circular structures.
-We use a mouse cortex time series data set GSE151565, which contains a total of 77 samples every 3 hours, for 36 hours.
-We compute the dependence structure of the genes as well as the variances of marginal distributions using the 12 time point 0 samples, and compute the means of gene expressions at each time point.
-We then use these to simulate 20 time series data sets for each of the independent, PCA, Wishart, and corpcor simulation methods.
+We chose a mouse cortex time series data set GSE151565, which contains a total of 77 samples every 3 hours, for 36 hours.
+We computed the dependence structure of the genes as well as the variances of marginal distributions using the 12 time point 0 samples and computed the means of gene expressions at each time point.
+We then used these to simulate 20 time series data sets for each of the independent, PCA, Wishart, and corpcor simulation methods.
 
@@
 
-We find that CYCLOPS performance depends significantly on the dependence structure of the simulated time series data (@fig-CYCLOPS), indicating that the presence of gene-gene dependence is highly impactful on CYCLOPS' performance.
-Overall, CYCLOPS performance is worse in the corpcor and Wishart data than in the independent data, but surprisingly the pca data shows better performance.
-Since by default CYCLOPS performs dimension reduction to the number of dimensions needed to capture 80% of the variance, this performance difference is likely partially due to the number of eigengenes needed by CYCLOPS to meet the variance threshold (@fig-CYCLOPS b).
+By default, CYCLOPS performs dimension reduction so that each dimension (called an "eigengene") contains at least 3% of the total variance.
+We found that CYCLOPS performance depended significantly on this parameter, with the default producing good performance across all simulations.
+However, when dropping CYCLOPS to require just 2% variance in each eigengene, we found that its performance depends significantly on the dependence structure of the simulated time series data (@fig-CYCLOPS a).
+At that setting, CYCLOPS performance is much higher in the PCA method and moderately improved in Wishart method, compared to the independent method. 
+This difference is likely driven by the difference in the number of eigengenes used (@fig-CYCLOPS b), which is a measure of how much dependence is present in the data set.
+This demonstrates that the correlational structure of the transcriptome can have a major impact on performance.
 
@@
 #| label: fig-CYCLOPS
-#| fig-cap: "Performance of CYCLOPS on simulated time series data sets based on mouse cortex data set (GSE151565).
-(a) Violin plot comparing absolute circular correlations between true phases and CYCLOPS estimated phases on the simulated data sets with the four methods.
-(b) Violin plot comparing the number of eigengenes used by CYCLOPS between the four methods; the red dotted line indicates the number of eigengenes used by CYCLOPS on the real data (5).
+#| fig-cap: "Performance of CYCLOPS on simulated time series data sets based on mouse cortex data set (GSE151565), using eigengenes of at least 2% variance.
+(a) Absolute circular correlations between true phases and CYCLOPS estimated phases on the simulated data sets.
+(b) The number of eigengenes used by CYCLOPS; the dotted line indicates the number of eigengenes used by CYCLOPS on the real data (5).
 (c-f) Examples of CYCLOPS estimated phases on the simulated data sets.
@@
 Surprisingly, the spiked Wishart and corpcor methods improved in this metric only slightly compared to the simulations with independent genes.
-These observations demonstrate that there is room for future improvements over independent data in these techniques, possibly incorporating more recent developments in copulae [@copulae@].
+These observations demonstrate that there is room for future improvements over independent data in these techniques, possibly incorporating more recent developments in copulae [@copulae].
 Possibly, this could demonstrate the limitations of methods based on the multivariate normal distribution or of the low-rank approximation used by all three of our methods. 
@@
 
+JY received funding from National Institute of Neurological Disorders and Stroke (5R01NS048471).
 TB and GG received funding support from the National Center for Advancing Translational Sciences Grant (5UL1TR000003).

### 2025-01-23 manuscript/circular_correlation.R

@@ -6,21 +6,23 @@ library(tidyr)
 library(stringr)
 library(patchwork)
 
-TYPE_ORDER <- c("indep", "pca", "wishart", "corpcor")
-TYPE_COLORS <- RColorBrewer::brewer.pal(4, "Dark2")
+METHOD_ORDER <- c("indep", "pca", "wishart", "corpcor")
+METHOD_COLORS <- RColorBrewer::brewer.pal(4, "Dark2")
+METHOD_ORDER2 <- c("real", METHOD_ORDER) # Includes "real"
+METHOD_COLORS2 <- c("black", METHOD_COLORS)
 
 # formula for the circular correlation
 get_cir_corr <- function(X, Y) {
     sum1 <- 0
     sum2 <- 0
     sum3 <- 0
-    
+
     if (length(X) != length(Y)) {
         return("X and Y have different lengths")
     } else {
         n <- length(X)
     }
-    
+
     for (i in (1:(n-1)) ) {
         for (j in ((i+1):n)) {
             sum1 <- sum1 + sin(X[i] - X[j])*sin(Y[i] - Y[j])
@@ -28,11 +30,11 @@ get_cir_corr <- function(X, Y) {
             sum3 <- sum3 + sin(Y[i] - Y[j])^2
         }
     }
-    
+
     rho <- sum1/(sqrt(sum2)*sqrt(sum3))
-    
+
     return(rho)
-        
+
 }
 
 # compute circular correlation between CYCLOP-estimated phases and true_counts_counts_counts phases for all 40 datasets
@@ -41,35 +43,36 @@ for (method in c("indep", "pca", "wishart", "corpcor")) {
     for (j in (0:19)) {
         phaselist <- read_csv(paste0("../processed/cyclops/",method,"/batch=", j, "/cyclops_estimated_phaselist.csv"))
         for (i in (1:32)) {
-            phaselist[i, 'ID'] <- strsplit(as.character(phaselist[i, 'ID']), "_")[[1]][1] |> 
+            phaselist[i, 'ID'] <- strsplit(as.character(phaselist[i, 'ID']), "_")[[1]][1] |>
                 substring(3)
         }
         phaselist$new_ID <- as.numeric(phaselist$ID)/12*pi
-        
+
         corrs[[method]][j+1] <- get_cir_corr(phaselist$new_ID, phaselist$phase)
     }
 }
 
-# violin plot comparing circular correlation for simulated datasets with and without dependence
[…]

### 2025-01-23 manuscript/compare_to_real.R

@@ -13,8 +13,8 @@ set.seed(55)
 
 # CONFIGURATION
 HIGH_EXPR_CUTOFF <- 300
-TYPE_ORDER <- c("real", "indep", "pca", "wishart", "corpcor")
-TYPE_COLORS <- c("black", RColorBrewer::brewer.pal(length(TYPE_ORDER)-1, "Dark2"))
+METHOD_ORDER <- c("real", "indep", "pca", "wishart", "corpcor")
+METHOD_COLORS <- c("black", RColorBrewer::brewer.pal(length(METHOD_ORDER)-1, "Dark2"))
 
 # Load the real data ---------------------------------
 raw <- read_tsv("../data/GSE151923_metaReadCount_ensembl.txt.gz") |>
@@ -86,7 +86,7 @@ projected_indep_draws <- predict(pca, t(scaled_indep_draws[!constant_rows, ]))
 both_data <- data.frame(list(
     PC1 = c(projected_read_data[1:12,"PC1"], projected_draws[1:12,"PC1"], projected_corpcor[1:12,"PC1"], projected_wishart[1:12,"PC1"], projected_indep_draws[1:12,"PC1"]),
     PC2 = c(projected_read_data[1:12,"PC2"], projected_draws[1:12,"PC2"], projected_corpcor[1:12,"PC2"], projected_wishart[1:12,"PC2"], projected_indep_draws[1:12,"PC2"]),
-    type = c(
+    method = c(
         rep("real", 12),
         rep("pca", 12),
         rep("corpcor", 12),
@@ -94,11 +94,11 @@ both_data <- data.frame(list(
         rep("indep", 12)
     )
 )) %>%
-  mutate(type = factor(type, levels=TYPE_ORDER, ordered=TRUE))
-#both_data$type <- factor(both_data$type, levels=c("real", "dependent sim", "independent sim"))
-g3<-ggplot(data = both_data, aes(x=PC1, y=PC2,color=type)) +
+  mutate(method = factor(method, levels=METHOD_ORDER, ordered=TRUE))
+#both_data$method <- factor(both_data$method, levels=c("real", "dependent sim", "independent sim"))
+g3<-ggplot(data = both_data, aes(x=PC1, y=PC2,color=method)) +
    geom_point() +
-  scale_color_manual(values = TYPE_COLORS, breaks = TYPE_ORDER, name="type")
[…]

### 2025-01-24 scripts/simulate_data.Rmd
Update vignette

@@
 
-Install the required packages, and choose a simulation method (one of "indep", "pca", "corpcor" and "wishart").
+Install the required dependencies if not already installed:
+```{r}
+# The simulator
+if (!requireNamespace("remotes", quietly = TRUE))
+    install.packages("remotes")
+if (!requireNamespace("dependentsimr", quietly = TRUE))
+    remotes::install_github("tgbrooks/dependentsimr")
 
+# DESeq2 - optional dependency only needed if doing RNA-seq data sets
+# Needed for type = "DESeq2"
+if (!requireNamespace("BiocManager", quietly = TRUE))
+    install.packages("BiocManager")
+if (!requireNamespace("DESeq2", quietly = TRUE))
+    BiocManager::install("DESeq2")
+
+# Two more dependencies that are used in some simulation modes
+# Needed for method = "corpcor"
+if (!requireNamespace("corpcor", quietly)) 
+    install.packages("corpcor")
+
+# Needed for method = "spiked Wishart"
+if (!requireNamespace("sparsesvd"))
+    install.packages("sparsesvd")
+
+# Extras needed for this vignette
+if (!requireNamespace("tidyverse", quietly = TRUE))
+    install.packages("tidyverse")
+```
+
+Load the libraries:
 ```{r}
@@
 library(DESeq2)
+```
 
-method <- 'pca'
+Set a seed for reproducibility.
+```{r}
+set.seed(0)
 ```
+
+## Load data
+
+Simulations with `dependentsimr` start with a real data set that should be mimicked.
+This data can be almost any shape, including continuous, or discrete.
+Here, we will simulate RNA-seq, which has discrete counts that we handle with a special "DESeq2" model.
+Other data types can use normal, Poisson, or empirical (which uses only the values present in the data set) types, or even a combination of these.
 
@@
 
-For the given simulation method, get the dependence structure from the real data set.
+The first step is to compute a 'random structure' for the data, which encodes how the variables (genes, in this case) in the data set are dependent upon each other.
+This computes a covariance matrix which captures the dependence between genes, however, that matrix is not the same as the usual sample covariance matrix.
+It also computes the shape (such as mean and variance) of each variable individually.
 
+There are three implemented methods ("pca", "spiked Wishart", and "corpcor") with different characteristics.
+We demonstrate here each.
+
+First, the "pca" method takes a `rank` and includes only gene-gene dependence up to that many dimensions.
+Typically small values (2 or 3, for example) would be used for `rank`, similar to when performing PCA, unless a large number of input samples are available.
+Note that we put our `read_counts` data into a `list()`.
+This is done since `get_random_structure` also supports multiple data sets of different 'modes' (for example, proteomics and transcriptomics done on the same samples).
+We only have one 'mode' that we give the arbitrary name of `data`.
 ```{r}
 
-# Produce simulated data
-if (method == "pca") {
-    set.seed(1)
-    # Simulate with the PCA method
-    rs <- get_random_structure(list(data=male_read_counts), method="pca", rank=2, types="DESeq2")
-} else if (method == "corpcor") {
-    set.seed(2)
-    # Simulate with the corpcor method
-    rs <- get_random_structure(list(data=male_read_counts), method="corpcor", types="DESeq2")
-} else if (method == "wishart") {
-    set.seed(3)
-    # Simulate with the spiked Wishart method
-    rs <- get_random_structure(list(data=male_read_counts), rank=11, method="spiked Wishart", types="DESeq2")
-} else if (method == "indep") {
-    set.seed(4)
-    # Simulate without any dependence
-    dep_rs <- get_random_structure(list(data=male_read_counts), method="pca", rank=2, type="DESeq2")
-    rs <- remove_dependence(dep_rs)
-}
+rs_pca <- get_random_structure(list(data=read_counts), method="pca", rank=2, types="DESeq2")
 ```
 
-Get the read depth of samples in the real data set, and generate library sizes of the simulated data sets (5 data sets each with 20 samples).
+Second, the spiked Wishart method also takes a `rank` value but uses it differently.
+Typically, this will be larger, up to the total number of samples minus one, which is what we use here.
+For very large sample counts, `rank` might need to be capped well below the number of samples for computational efficiency reasons.
+```{r}
+rs_wishart <- get_random_structure(list(data=male_read_counts), rank=11, method="spiked Wishart", types="DESeq2")
+```
+
+Third, a method based off the `corpcor` package is implemented.
+It takes no parameters but often underestimates the amount of gene-gene dependence in the data.
+```{r}
+rs_corpcor <- get_random_structure(list(data=male_read_counts), method="corpcor", types="DESeq2")
+```
+
+Lastly, we could generate data without gene-gene dependence for comparison by removing the dependence from any of our random structures.
+```{r}
+rs_indep <- remove_dependence(rs_pca)
+```
+
+## Generate Data
+
+Real RNA-seq data sets typically have variation between samples in the total number of reads per sample.
+Here, we just copy the total number of reads from the real data set to get realistic values.
+A `library_size` of 1 corresponds to the average read depth of the data used to generate the random structure.
+This step is only needed when generating RNA-seq data with `type = "DESeq2"`.
+
 ```{r}
@@
 
-N_SAMPLES <- 100
+N_SAMPLES <- 6
 library_sizes <- sample(actual_library_sizes / mean(actual_library_sizes), size=N_SAMPLES, replace=TRUE)
@@
 
-## Simulate Control Data
+### Simulate Control Data
 
-Generate and save data+metadata (true values) from a specific random structure
+Now we generate some 'control' samples, which have the same expression levels (as well as gene-gene dependence) as was estimated from the original data set.
+We'll use `rs_pca` that we generated above for example, but any of the methods could be used.
 ```{r}
-generate <- function(rs, name) {
-  sim <- draw_from_multivariate_corr(rs, n_samples=N_SAMPLES, size_factors=library_sizes)$data
-  colnames(sim) <- paste("sample", 1:ncol(sim), sep='')
+control_sim_data <- draw_from_multivariate_corr(rs_pca, n_samples=N_SAMPLES, size_factors=library_sizes)$data
+```
+Note that we used `$data` here because that was the name we provided to `get_random_structure` and it was the only mode of data that we had.
+If multiple modes were used to generate the random structure, then we `draw_from_multivariate_corr` returns a list with each of them.
 
-sim |>
-    data.frame() |>
-    rownames_to_column("ENSEMBL_ID") |>
-    write.table(file=paste0("/home/jianingy/dependent_sim/dependent_sim_paper/notebook_data/", name, ".txt"), sep="\t", row.names=FALSE)
+### Simulate Differentially Expressed Data
 
-  # metadata
-  truth <- data.frame(
-    gene_id = rownames(sim),
-    mean = rs$marginals$data$q,
-    disperion = rs$marginals$data$dispersion
-  )
-  truth[is.na(truth$mean), 2] <- 0 # NA mean is 0
+To generate a set of 'treatment' samples which differ from the 'control' samples, we will have to modify the random structure to reflect the desired expression values of the treatment group.
+We do this by just choosing some genes at random and picking a fold-change for each at random.
 
-  return(sim)
-}
-
-male_sim <- generate(rs, name=paste0("Mouse.Cortex.Male.",method,".Control"))
-```
-
-## Simulate Differentially Expressed Data
-
-Randomly select some genes to be DE.
 ```{r}
+# fraction of all genes to modify in treatment
 FRAC_DE <- 0.05
+# Min and max value for the log fold change (LFC) for each DE gene chosen
+# non-DE genes will have a LFC of 0
 LFC_MIN <- 0.2
 LFC_MAX <- 2.0
-N_DE <- floor(FRAC_DE * nrow(male_sim))
+N_DE <- floor(FRAC_DE * nrow(control_sim_data))
 
+# Pick which genes are DE
 de_genes <- sample(nrow(male_sim), N_DE)
-```
 
-Simulate log fold changes for the case data sets.
-```{r}
+# Pick the LFCs of each of those genes
 de_lfc <- runif(N_DE, LFC_MIN, LFC_MAX) * sample(c(-1, 1), size=N_DE, replace=TRUE)
@@
 
-Generate Cases data sets.
+Now, we create a new random structure object and modify it to have the desired expression values.
+Information about the marginal distribution of each gene (e.g., mean and variance but not dependence on other genes) is stored in `rs$marginals` for a random structure `rs`.
+The specifics depend upon the `type` used in `get_random_structure`, but for `DESeq2`, the `q` value determines the mean expression, so we change that here.
+
 ```{r}
-if (method == "indep") {
-    male_de_rs <- dep_rs
-    male_de_rs$marginals$data$q[de_genes] <- 2^(de_lfc) * male_de_rs$marginals$data$q[de_genes]
+rs_treatment <- rs_pca
+rs_treatment$marginals$data$q[de_genes] <- 2^(de_lfc) * rs_treatment$marginals$data$q[de_genes]
+```
 
-    male_indep_de_rs <- remove_dependence(male_de_rs)
-    male_indep_de_sim <- generate(male_indep_de_rs, name="Mouse.Cortex.Male.indep.Case")
-} else {
-    male_de_rs <- rs
-    male_de_rs$marginals$data$q[de_genes] <- 2^(de_lfc) * male_de_rs$marginals$data$q[de_genes]
+Lastly, we generate the data for the 'treatment' samples.
+```{r}
+treatment_sim_data <- draw_from_multivariate_corr(rs_treatment, n_samples=N_SAMPLES, size_factors=library_sizes)$data
+```
 
-    male_de_sim <- generate(male_de_rs, name=paste0("Mouse.Cortex.Male.",method,".Case"))
-}
+## Plot the generated data
+
+To demonstrate the data, we show a PCA plot of the control and treatment groups
+```{r}
+sim_data <- cbind(control_sim_data, treatment_sim_data)
+pca <- prcomp(t(sim_data), center=TRUE, scale.=TRUE, rank.=2)
+pca_points <- predict(t(sim_data))
+pca_data <- tibble(x=pca_points[,1], y=pca_points[,2]) |>
+    mutate(group = c(rep("control", N_SAMPLES), rep("treatment", N_SAMPLES)))
+ggplot(pca_data, aes(x, y, color=group))
 ```

### 2025-02-06 manuscript/paper.qmd
Update affiliations

@@
       affiliations:
-        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania"
         - name: "Chronobiology and Sleep Institute, University of Pennsylvania"
@@
       affiliations:
-        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania"
+        - name: "Department of Genetics, University of Pennsylvania"
     - name: "Thomas G. Brooks"

### 2025-05-05 scripts/SPsimSeq.mouse_timeseries.R
Add mouse timeseries comparison
Exclude one subject from simulations due to being a large outlier in PCA

@@ -1,0 +1,45 @@
+library(tidyverse)
+library(SPsimSeq)
+
+N_SAMPLES <- 100
+
+read_counts <- read.csv("processed/Cortex_ZT0-counts.csv") |> 
+    select(!all_of("X331308_CRM55_WT_CTXR_26WKS_M_ZT0_L1.D704")) |>
+    column_to_rownames(var="raw_data.EnsemblID")
+
+# Produce simulated data
+not_all_zero <- rowSums(read_counts) > 0
+
+input <- read_counts[not_all_zero,] |> as.matrix()
+set.seed(1)
+results <- SPsimSeq(
+    n.sim = 1,
+    s.data = input,
+    n.genes = input |> nrow(),
+    batch.config = 1, # only one batch,
+    group.config = 1, # only one group
+    pDE = 0, # no DE since no groups
+    tot.samples = N_SAMPLES,
+    model.zero.prob = FALSE, # since bulk RNA
+    genewiseCor = TRUE, # gene-gene dependence, notes that it could be SLOW
+    log.CPM.transform = TRUE,
+    result.format = "list",
+    return.details = TRUE,
+    verbose = TRUE,
+)
+
+saveRDS(results, "processed/SPsimSeq/GSE151565.RDS")
+
+counts <- results$sim.data.list[[1]]$counts
+# Add back the original gene names onto the simulated data
+counts |>
+    as.data.frame() |>
+    rownames_to_column("temp_id") |>
+    left_join(
+        results$sim.data.list[[1]]$rowData |>
+            rownames_to_column("temp_id"),
+        by = join_by(temp_id == temp_id),
+    ) |>
+    mutate(gene_id = source.ID) |>
+    select(-temp_id, -DE.ind, -source.ID) |>
+    write_tsv("simulated_data/SPsimSeq.GSE151565.txt")

### 2025-05-05 scripts/compare_to_real.R

@@ -10,6 +10,7 @@ set.seed(55)
 
 #dataset = "GSE81142"
 #dataset = "GSE151923"
+#dataset = "GSE151565"
 dataset <- snakemake@wildcards$dataset
 
 # CONFIGURATION
@@ -36,17 +37,6 @@ if (dataset == "GSE151923") {
 
     draws_spsimseq <- read_tsv("simulated_data/SPsimSeq.GSE151923.txt") |>
         column_to_rownames("gene_id")
-    # fill in missing all-zero genes and reorder
-    missing <- rownames(read_counts)[!(rownames(read_counts) %in% rownames(draws_spsimseq))]
-    present <- rownames(draws_spsimseq)
-    zero_genes <- matrix(0, length(missing), ncol(draws_spsimseq))
-    rownames(zero_genes) <- missing
-    colnames(zero_genes) <- colnames(draws_spsimseq)
-    draws_spsimseq <- rbind(
-        draws_spsimseq,
-        zero_genes
-    )
-    draws_spsimseq <- draws_spsimseq[rownames(read_counts),] # put in the right order
 } else if (dataset == "GSE81142") {
     HIGH_EXPR_CUTOFF <- 30
     raw <- read_tsv("data/GSE81142.counts.txt.gz") |> column_to_rownames("GENE_ID")
@@ -64,27 +54,43 @@ if (dataset == "GSE151923") {
 
     draws_spsimseq <- read_tsv("simulated_data/SPsimSeq.GSE81142.txt") |>
         column_to_rownames("gene_id")
-    # fill in missing all-zero genes and reorder
-    missing <- rownames(read_counts)[!(rownames(read_counts) %in% rownames(draws_spsimseq))]
-    present <- rownames(draws_spsimseq)
-    zero_genes <- matrix(0, length(missing), ncol(draws_spsimseq))
-    rownames(zero_genes) <- missing
-    colnames(zero_genes) <- colnames(draws_spsimseq)
-    draws_spsimseq <- rbind(
-        draws_spsimseq,
-        zero_genes
-    )
-    draws_spsimseq <- draws_spsimseq[rownames(read_counts),] # put in the right order
+} else if (dataset == "GSE151565") {
+    HIGH_EXPR_CUTOFF <- 30
+    raw <- read.csv("processed/Cortex_ZT0-counts.csv") |> 
[…]

### 2025-05-05 scripts/simulate_time_data.R

@@ -9,6 +9,7 @@ method <- snakemake@wildcards$method
 
 # use ZT0 of raw time series data to get dependence structure
 male_read_counts <- read.csv("processed/Cortex_ZT0-counts.csv") |> 
+    select(!all_of("X331308_CRM55_WT_CTXR_26WKS_M_ZT0_L1.D704")) |>
     column_to_rownames(var="raw_data.EnsemblID")
 
 # Get dependence structure
@@ -23,7 +24,7 @@ if (method == "pca") {
 } else if (method == "wishart") {
     set.seed(3)
     # Simulate with the spiked Wishart method
-    male_rs <- get_random_structure(list(data=male_read_counts), rank=11, method="spiked Wishart", types="DESeq2")
+    male_rs <- get_random_structure(list(data=male_read_counts), rank=10, method="spiked Wishart", types="DESeq2")
 } else if (method == "indep") {
     set.seed(4)
     # Simulate without any dependence

### 2025-05-05 manuscript/paper.qmd
Add supplemental document

@@
 
-```{r}
-#| include: false
-source("compare_to_real.R")
-```
-
 To compare the three simulation methods with a real data set, we chose as a references data set 12 mouse cortex RNA-seq samples from accession GSE151923 [@Wang2022].
@@
 #| label: fig-compare-to-real
-#| fig-cap: !expr glue::glue("Comparison to real data run on a mouse cortex data set from GSE151923.
+#| fig-cap: "Comparison to real data run on a mouse cortex data set from GSE151923.
 (a-b) Comparison of gene (a) mean expression and (b) variance, log-scaled in real and PCA simulated data.
@@
 (c) Quantile-quantile plot comparing correlation values of gene pairs from real data and simulated data (both with and without dependence).
-Genes with at least {HIGH_EXPR_CUTOFF} reads were used.
+Genes with at least 100 reads were used.
 Values on the diagonal line indicate a match between the simulated and real data sets.
@@
 (e) Principal component analysis was performed on all data sets and the variance captured by the top components is shown.
-Unlike (d), these components were fit from each data set considered separately instead of reusing the weights from the real data.")
+Unlike (d), these components were fit from each data set considered separately instead of reusing the weights from the real data."
 #| fig-width: 10
 #| fig-height: 10
-compare_to_real_plot
+readRDS("../processed/compare_to_real_plot.GSE151923.RDS")
 ```

### 2025-05-05 manuscript/supplemental/supplemental.qmd
Add supplemental document

---
title: Generating Correlated Data for Omics Simulation
author:
    - name: "Jianing Yang"
      orcid: 0000-0002-2048-9398
      affiliations:
        - name: "Chronobiology and Sleep Institute, University of Pennsylvania"
    - name: "Gregory R. Grant"
      orcid: 0000-0002-0139-7658
      affiliations:
        - name: "Department of Genetics, University of Pennsylvania"
    - name: "Thomas G. Brooks"
      orcid: 0000-0002-6980-0079
      affiliations:
        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania"
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  cache: true
  warning: false
format:
  html:
    toc: true
    toc-expand: true
bibliography: ../references.bib
---

## Supplemental Material

### Comparisons to real data

```{r}
#| label: fig-compare-to-real
#| fig-cap: "Comparison to real data run on a fly whole body data set from GSE81142. (a-b) Comparison of gene (a) mean expression and (b) variance, log-scaled in real and PCA simulated data. The line of equality is marked in black. Points are colored according to the density of points in their region. Wishart and corpcor methods give similar results (not shown). (c) Quantile-quantile plot comparing correlation values of gene pairs from real data and simulated data (both with and without dependence). Genes with at least 30 reads were used. Values on the diagonal line indicate a match between the simulated and real data sets. (d) Projections onto the top two principal components of the real data set for both real and simulated data. All 8 simulations (96 samples for each simulation) shown. (e) Principal component analysis was performed on all data sets and the variance captured by the top components is shown. Unlike (d), these components were fit from each data set considered separately instead of reusing the weights from the real data."
#| fig-width: 10
#| fig-height: 10
readRDS("../../processed/compare_to_real_plot.GSE81142.RDS")
```

```{r}
#| label: fig-compare-to-real2
#| fig-cap: "Comparison to real data run on a mouse cortex timeseries body data set from GSE151565. Only timepoint ZT0 was used for this comparison. (a-b) Comparison of gene (a) mean expression and (b) variance, log-scaled in real and PCA simulated data. The line of equality is marked in black. Points are colored according to the density of points in their region. Wishart and corpcor methods give similar results (not shown). (c) Quantile-quantile plot comparing correlation values of gene pairs from real data and simulated data (both with and without dependence). Genes with at least 30 reads were used. Values on the diagonal line indicate a match between the simulated and real data sets. (d) Projections onto the top two principal components of the real data set for both real and simulated data. All 8 simulations (96 samples for each simulation) shown. (e) Principal component analysis was performed on all data sets and the variance captured by the top components is shown. Unlike (d), these components were fit from each data set considered separately instead of reusing the weights from the real data."
#| fig-width: 10
#| fig-height: 10
readRDS("../../processed/compare_to_real_plot.GSE151565.RDS")
```

## Supplemental Methods

### Running SPsimSeq

The R package [SPsimSeq](https://github.com/CenterForStatistics-UGent/SPsimSeq) [@Assefa2020] provides a dedicated RNA-seq and single-cell RNA-seq simulator using a Gaussian copula approach to simulate gene dependence.
We ran SPsimSeq using the following options:

``` R
    n.sim = 1,
    s.data = input,
    n.genes = input |> nrow(),
    batch.config = 1,
    group.config = 1,
    pDE = 0,
    tot.samples = 100,
    model.zero.prob = FALSE,
    genewiseCor = TRUE,
    log.CPM.transform = TRUE,
    result.format = "list",
    return.details = TRUE,
    verbose = TRUE,
```

This produces bulk RNA-seq data with no batching or group effects or differentially expressed (DE) genes.
Raw count values provided as the input matrix.
For dataset GSE151565, only the ZT0 timepoints were used.
This is the same set of samples that were used in our simulation to determine the correlation structure (copula).

SPsimSeq was not run in the two example applications, DESeq2 and CYCLOPS, due to substantial differences in how it generates DE across samples compared to our method.
DE in SPsimSeq is done by mimicking a random set of detected DE genes in the reference data set.
SPsimSeq does not report either the group mean values or the log fold changes used for these genes, so we are not able to modify our own simulation to have the same set of DE genes at the same effect sizes.

### 2025-05-05 manuscript/paper.qmd
Some updates based off review comments

@@
 
-Assume that we have a reference dataset $X$ given by an $p \times n$ data matrix of $p$ features measured in each of $n$ independent samples.
-We want to capture correlations between the $p$ features, which could represent gene expressions, protein abundances, or other measured values.
+We start with a reference data set $X$ given by an $p \times n$ data matrix of $p$ features measured in each of $n$ independent samples, which could be any real data set of interest.
+Our goal is to generate simulated data set that looks, at least on average, the same as this reference data set.
+To do this, we need to capture correlations between the $p$ features, which could represent gene expressions, protein abundances, or other measured values.
 We refer to these features as genes for simplicity.
-Our goal is to generate simulated data with the same $p$ genes, the same marginal distributions of each gene as in $X$ and realistic gene-gene dependence.
+
+Our methods are based off the well-known Gaussian copula method, which uses a multivariate normal distribution to capture aspects of the gene-gene dependence, without necessarily assuming that the data is normally distributed.
+This multivariate normal distribution is parameterized by a $p \times p$ covariance matrix, $\Sigma$.
+In omics, choosing this $\Sigma$ presents a challenge since $n < p$.
+The most obvious choice is to use the sample covariance matrix $\widehat \Sigma$ of the reference data set (possibly after some transformation of the data).
+It is important to note that using $\widehat Sigma$ itself is a poor choice for $\Sigma$.
+Indeed, we consider a choice of $\Sigma$ to be successful if data sets simulated using $\Sigma$ have sample covaraince matrices that are similar to $\widehat \Sigma$.
+Counter-intuitively, data simulated using $\Sigma := \widehat \Sigma$ have sample covariance matrices that systematically and substantially differ from $\widehat \Sigma$.
+
+This is due to the fact that $\widehat \Sigma$ is low rank, specifically at most rank $n - 1$, which is less than $p$ in typical omics studies.
+This limitation means that any data generated from the multivariate normal distribution of that $\Sigma$ will lie on a $n-1$-dimensional hyperplane.
+Therefore, newly simulated samples will actually be linear combination of the samples of the reference data set.
+Intuitively, this means that the entire probability distribution of possible $p \times n$ dimensional omics data sets have been squished into a $n-1$ dimensional space, generating much higher gene-gen correlations than in the reference data set.
+In contrast, we know that omics contains at least some level of random noise that is approximately independent from one gene to the next.
+For example, RNA-seq involves a random sampling of fragments from the true sample of RNA molecules and other sources involve various background noise effects.
+We therefore need to "inject" some level of additional randomness that is independent between genes on top of the correlations captured by $\widehat \Sigma$
+
+We present three methods - called here PCA, spiked Wishart, and corpcor - for making this choice of $\Sigma$, all which take the form
+$$ \Sigma = D^2 + PP^T $$
+where $D$ is a diagonal matrix capturing the injected independence and $PP^T$ is a low-rank correlation matrix derived from $\widehat \Sigma$.
+An additional advantage of this special form is that data may be generated very efficiently, see Methods, compared to arbitrary $\Sigma$ which typically require a Cholesky or eigenvalue decomposition.
 
@@
 
+We also ran for comparison SPsimSeq, another RNA-seq simulator for both bulk and single-cell that uses a Gaussian copula-based model of dependence, see Supplemental Methods for details.
+
 Simulated data captures the genes' mean and variance accurately (@fig-compare-to-real a-b).
@@
 
-Then, we computed the gene-gene correlation on pairs of high-expressed genes (at least 300 mean reads).
+Then, we computed the gene-gene correlation on pairs of high-expressed genes (at least 100 mean reads in the real data set).
 The simulation with independence showed the least levels of gene-gene correlations (@fig-compare-to-real d).
 However, the PCA method overshot the reference data set and the spiked Wishart and corpcor methods only slightly improved upon the independent simulation.
+The SPsimSeq simulator performed similarly to the PCA method.
 
@@
 The PCA method puts a large amount of variance into the first two components (due to using $k=2$) and then undershoots the other components.
+Like PCA, the SPsimSeq simulator, overshot the first PCA component but did a better job matching lower PCA coordinates.
 
@@
 
-We used a fly whole body RNA-Seq data set GSE81142 and selected samples of male flies without treatment and after at least 2 hours of feeding to simulate 5 "control" samples.
+We used a fly whole body RNA-Seq data set GSE81142, see Supplemental Figure 1, and selected samples of male flies without treatment and after at least 2 hours of feeding to simulate 5 "control" samples.
+Unnormalized read counts were used as the reference data set and all simulations used the DESeq2-based negative binomial model.
 We then randomly selected 5% of the genes to be differentially expressed, with absolute $\log_2$ fold change uniformly distributed between $0.2$ and $2.0$, either up or down regulated chosen randomly, and simulated 5 "experimental" samples.
@@
 We observed that DESeq2 is anti-conservative on all data sets, with similar mean true FDRs for each estimated FDR cutoff.
-However, there was a greater variance in the performance of DESeq2 on the data sets simulated with dependence, indicating that it performs less consistently on data sets with gene-gene dependence, as in real data sets.
+However, we note that this anti-conservative nature is most present at very low FDR thresholds, was typically associated with only a single digit number of false positive genes.
+While that mean behavior was consistant whether or not the simulation included gene-gene dependence, simulations with dependence showed a substantially greater variance in the performance of DESeq2.
 
-To demonstrate the application of our simulation method for another organism, we also simulated data sets using mouse cortex data set GSE151923 [@Wang2022] and selected samples from male mice.
+To demonstrate the application of our simulation method for another organism, we also simulated data sets using the mouse cortex data set GSE151923 [@Wang2022] and selected samples from male mice.
 We then simulated differential expression experiments as above and observed a similar result (@fig-DESeq2 e-f) as for the fly whole body data sets.
@@
 We next used our simulation method to benchmark CYCLOPS [@Anafi2017], which infers relative times for a set of unlabeled samples using an autoencoder to identify circular structures.
-We chose a mouse cortex time series data set GSE151565, which contains a total of 77 samples every 3 hours, for 36 hours.
-We computed the dependence structure of the genes as well as the variances of marginal distributions using the 12 time point 0 samples and computed the means of gene expressions at each time point.
+We chose a mouse cortex time series data set GSE151565, see Supplemental Figure 2, which contains a total of 77 samples every 3 hours, for 36 hours.
+As before, unnormalized read counts were used as the reference data set and all simulations used the DESeq2-based negative binomial model.
+We computed the dependence structure of the genes as well as the variances of marginal distributions using the 11 time point 0 samples after removing one outlier and computed the means of gene expressions at each time point.
 We then used these to simulate 20 time series data sets for each of the independent, PCA, Wishart, and corpcor simulation methods.
@@
 In the first, DESeq2 results were substantially more variable when simulating with gene-gene dependence.
+This indicates that DESeq2's had an increased chance of a larger simultaneous number of false positives, as well as increased chance of fewer false positives, when data included realistic dependence.
+We therefore recommend that benchmarking of differential expression methods should include gene-gene dependence whenever possible.
+This is despite the fact that differential expression is largely addressing a problem of the behavior of individual genes.
 In the second, CYCLOPS performance in estimating circadian phases depends upon gene-gene dependence, was sensitive to dependence structure of the data.
+Unlike DESeq2, CYCLOPS explicitly makes use of the correlations of genes and therefore this result is not surprising but still demonstrates the potential impact of the assumption of independence when benchmarking.
 
@@
 The R package [SPsimSeq](https://github.com/CenterForStatistics-UGent/SPsimSeq) [@Assefa2020] provides a dedicated RNA-seq and single-cell RNA-seq simulator using a Gaussian copula approach to simulate gene dependence.
-In contrast to this package, it uses WGCNA to determine the correlation matrix, which is a gene network approach.
-However, this method takes significant computational resources.
-Indeed, the `SPsimSeq` paper generated data for just 5000 genes based on a randomly sampled 5000 gene subset of the RNA-seq data and our attempts to use `SPsimSeq` to generate a full sample exhausted the memory of a 24GB computer.
-In contrast, our method runs in seconds to generate a 40,000 gene samples on the same computer.
-`SPsimSeq` is more specialized and full-featured for RNA-seq simulation, providing, for example, native differential expression (DE) options.
+In contrast to this package, it uses a two-step randomization.
+In short, it first fits distributions to each genes and then bins those distributions into discrete buckets.
+Then, new data is generated by drawing from a multivariate normal distribution whose covariance matrix equals the sample covariance matrix of the reference data set.
+These values are used to choose which bucket each gene is drawn from.
+Then, each gene's final value is drawn uniformly and independently from within its chosen bucket.
+Therefore, SPsimSeq induces correlation in the first multivarainte normal draw, but then weakens it in the second independent uniform step.
+This two-step process makes it challenging to compare on theoretical grounds to our proposed methods since it will dependend upon parameters such as the number of buckets used.
+More fine-grained buckets will induce higher correlation, and in the limit of infinite buckets, it should approach our PCA method with $k$ equal to the number of samples in the reference data set.
+`SPsimSeq` is more specialized and full-featured for RNA-seq simulation, providing, for example, native differential expression (DE) options,.
 In comparison, our `dependentsimr` package requires manually setting marginal expression values to inject DE, but also supports other marginal distributions for situations outside of RNA-seq.
@@
 Source code for the `dependentsimr` package is available at [github.com/tgbrooks/dependent_sim](https://github.com/tgbrooks/dependent_sim/).
-All data used is available with accession numbers GSE151923, GSE81142, GSE151565.
+All data used is available from the Gene Expression Omnibus (GEO) with accession numbers GSE151923, GSE81142, GSE151565.
 

### 2025-05-06 manuscript/paper.qmd
Add new figure demonstrating sample covariance matrix

@@
   cache: true
+  warning: false
 format:
@@
 The most obvious choice is to use the sample covariance matrix $\widehat \Sigma$ of the reference data set (possibly after some transformation of the data).
-It is important to note that using $\widehat Sigma$ itself is a poor choice for $\Sigma$.
-Indeed, we consider a choice of $\Sigma$ to be successful if data sets simulated using $\Sigma$ have sample covaraince matrices that are similar to $\widehat \Sigma$.
-Counter-intuitively, data simulated using $\Sigma := \widehat \Sigma$ have sample covariance matrices that systematically and substantially differ from $\widehat \Sigma$.
+However, using $\widehat Sigma$ is a poor choice for $\Sigma$.
+Indeed, we consider a choice of $\Sigma$ to be successful if data sets of $n$ samples simulated using $\Sigma$ have sample covaraince matrices that are similar to $\widehat \Sigma$.
+Counter-intuitively, data simulated using $\Sigma := \widehat \Sigma$ have sample covariance matrices that systematically and substantially differ from $\widehat \Sigma$, see @fig-intro.
+
+```{r}
+#| label: fig-intro
+#| fig-cap: "Eigenvalues of the sample covariance matrix $\\widehat \\Sigma$ when estimated from the reference data set and when estimated from simulated data using $\\Sigma$ equal to $\\widehat \\Sigma$ of the reference data set ('simulated').
+Eigenvalues are numbered by size.
+100 genes and 12 samples were used for both reference and simulated data sets.
+Both reference and simulated data follow a multivariate normal distribution."
+#| fig-width: 5
+#| fig-height: 4
+
+library(MASS)
+library(tidyverse)
+
+# create reference aka 'real' data set
+temp <- matrix(rnorm(100*100), 100, 100)
+Sigma_true <- temp %*% t(temp)
+reference_data <- mvrnorm(n=12, mu=rep(0,100), Sigma = Sigma_true) |> t()
+Sigma_hat <- cov(reference_data |> t())
+
+# Run 20 simulations and collate eigenvalues of the sample covariance matrix
+temp <- list()
+temp[[1]] <-   tibble(
+  number = 1:11,
+  eigenvalue = eigen(Sigma_hat)$values[1:11],
+  type = "reference",
+  iter = 0,
+)
+for (i in 1:20) {
+  sim_data <- mvrnorm(n=12, mu=rep(0,100), Sigma = Sigma_hat) |> t()
+  Sigma_sim <- cov(sim_data |> t())
+
+  # collate eigenvalues of sample covariance matrix of the simulated data
+  temp[[length(temp) + 1]] <- tibble(
+    number = 1:11,
+    eigenvalue = eigen(Sigma_sim)$values[1:11],
+    type = "simulated",
+    iter = i,
+  )
+}
+eigenvalue_data <- bind_rows(temp)
+
+# Plot the eigenvalues
+ggplot(eigenvalue_data |> filter(type != "reference"), aes(x=number, y = eigenvalue, color = type, group=iter)) +
+  geom_line(linewidth=2, alpha=0.5) +
+  geom_line(data = eigenvalue_data |> filter(type == "reference"), linewidth=4) +
+  labs(x = "eigenvalue number", y = "eigenvalue")
+```
 
@@
 Therefore, newly simulated samples will actually be linear combination of the samples of the reference data set.
-Intuitively, this means that the entire probability distribution of possible $p \times n$ dimensional omics data sets have been squished into a $n-1$ dimensional space, generating much higher gene-gen correlations than in the reference data set.
+Intuitively, this means that the entire probability distribution of possible $p \times n$ dimensional omics data sets have been squished into a $n-1$ dimensional space, generating much higher gene-gene correlations than in the reference data set.
 In contrast, we know that omics contains at least some level of random noise that is approximately independent from one gene to the next.
-For example, RNA-seq involves a random sampling of fragments from the true sample of RNA molecules and other sources involve various background noise effects.
+For example, RNA-seq involves a random sampling of fragments from the true sample of RNA molecules and other omics modes involve various background noise effects.
 We therefore need to "inject" some level of additional randomness that is independent between genes on top of the correlations captured by $\widehat \Sigma$

### 2025-05-08 manuscript/supplemental/supplemental.qmd
Add benchmark of runtime + memory usage

@@
 
+### Benchmark
+```{r}
+#| label: fig-benchmark
+#| fig-cap: "Comparison of (a) run time and (b) memory usage for our methods and SPsimSeq.
+100 samples were generated with a varying number of genes.
+GSE151564 dataset was randomly subsetted to have the required number of genes to be used as the reference data set."
+#| fig-width: 8
+#| fig-height: 4
+library(tidyverse)
+library(patchwork)
+METHOD_ORDER <- c("real", "indep", "pca", "wishart", "corpcor", "SPsimSeq")
+METHOD_COLORS <- c("black", RColorBrewer::brewer.pal(length(METHOD_ORDER)-1, "Dark2"))
+benchmark <- read_tsv("../../processed/benchmark/results.txt")
+g1 <- ggplot(benchmark, aes(x=n_genes, y=time, color=method)) +
+    geom_line(linewidth=2) +
+    labs(x = "N genes", y = "time (seconds)") + 
+    scale_color_manual(values = METHOD_COLORS, breaks = METHOD_ORDER, name="method") +
+    scale_x_log10() +
+    scale_y_log10()
+g2 <- ggplot(benchmark, aes(x=n_genes, y=max_rss, color=method)) +
+    geom_line(linewidth=2) +
+    labs(x = "N genes", y = "memory (MB)") +
+    scale_color_manual(values = METHOD_COLORS, breaks = METHOD_ORDER, name="method") +
+    scale_x_log10() +
+    scale_y_log10()
+g1 + g2 + plot_annotation(tag_levels="a") + plot_layout(guides = "collect")
+```
+
 ## Supplemental Methods

### 2025-05-08 manuscript/paper.qmd
Add benchmark of runtime + memory usage

@@
 The most obvious choice is to use the sample covariance matrix $\widehat \Sigma$ of the reference data set (possibly after some transformation of the data).
-However, using $\widehat Sigma$ is a poor choice for $\Sigma$.
+However, using $\widehat \Sigma$ is a poor choice for $\Sigma$.
 Indeed, we consider a choice of $\Sigma$ to be successful if data sets of $n$ samples simulated using $\Sigma$ have sample covaraince matrices that are similar to $\widehat \Sigma$.
@@
 
-This is due to the fact that $\widehat \Sigma$ is low rank, specifically at most rank $n - 1$, which is less than $p$ in typical omics studies.
+One reason this occurs is that $\widehat \Sigma$ is low rank, specifically at most rank $n - 1$, which is less than $p$ in typical omics studies.
 This limitation means that any data generated from the multivariate normal distribution of that $\Sigma$ will lie on a $n-1$-dimensional hyperplane.
 Therefore, newly simulated samples will actually be linear combination of the samples of the reference data set.
-Intuitively, this means that the entire probability distribution of possible $p \times n$ dimensional omics data sets have been squished into a $n-1$ dimensional space, generating much higher gene-gene correlations than in the reference data set.
+Intuitively, this means that the entire probability distribution of possible $p$ dimensional data have been squished into a $n-1$ dimensional space, generating much higher gene-gene correlations than in the reference data set.
 In contrast, we know that omics contains at least some level of random noise that is approximately independent from one gene to the next.
@@
 where $D$ is a diagonal matrix capturing the injected independence and $PP^T$ is a low-rank correlation matrix derived from $\widehat \Sigma$.
-An additional advantage of this special form is that data may be generated very efficiently, see Methods, compared to arbitrary $\Sigma$ which typically require a Cholesky or eigenvalue decomposition.
+An additional advantage of this special form is that data may be generated very efficiently, see Methods, compared to arbitrary $\Sigma$ which typically require a Cholesky or eigenvalue decomposition, see Supplemental Figure S3.
 
@@
 Then, each gene's final value is drawn uniformly and independently from within its chosen bucket.
-Therefore, SPsimSeq induces correlation in the first multivarainte normal draw, but then weakens it in the second independent uniform step.
+Therefore, SPsimSeq induces correlation in the first multivarainte normal draw, but then injects additional independence in the second uniform step.
 This two-step process makes it challenging to compare on theoretical grounds to our proposed methods since it will dependend upon parameters such as the number of buckets used.

### 2025-05-08 scripts/SPsimSeq.benchmark.R

@@ -1,0 +1,43 @@
+library(tidyverse)
+library(SPsimSeq)
+
+N_SAMPLES <- 100
+N_GENES <- snakemake@wildcards$n_genes
+method <- "SPsimSeq"
+
+read_counts <- read.csv("processed/Cortex_ZT0-counts.csv") |> 
+    select(!all_of("X331308_CRM55_WT_CTXR_26WKS_M_ZT0_L1.D704")) |>
+    column_to_rownames(var="raw_data.EnsemblID")
+
+# Produce simulated data
+not_all_zero <- rowSums(read_counts) > 0
+
+input <- read_counts[not_all_zero,] |> as.matrix()
+selected_rows <- sample.int(nrow(input), N_GENES, replace=TRUE)
+input <- input[selected_rows,]
+message(paste("Size of input", input |> nrow()))
+set.seed(1)
+time <- system.time({
+    results <- SPsimSeq(
+        n.sim = 1,
+        s.data = input,
+        n.genes = input |> nrow(),
+        batch.config = 1, # only one batch,
+        group.config = 1, # only one group
+        pDE = 0, # no DE since no groups
+        tot.samples = N_SAMPLES,
+        model.zero.prob = FALSE, # since bulk RNA
+        genewiseCor = TRUE, # gene-gene dependence, notes that it could be SLOW
+        log.CPM.transform = TRUE,
+        result.format = "list",
+        return.details = TRUE,
+        verbose = TRUE,
+    )
+})
+message("Elapsed time:")
+message(time[['elapsed']])
+tibble(
+    method = method,
+    n_genes = N_GENES,
+    time = time[['elapsed']],
+) |> write_tsv(paste0("processed/benchmark/", method, "/", N_GENES, ".txt"))

### 2025-05-08 scripts/benchmark.R

@@ -1,0 +1,11 @@
+library(tidyverse)
+input = snakemake@input$results
+benchmarks = snakemake@input$benchmarks
+temp <- list()
+for (i in seq_along(input)) {
+    temp[[length(temp)+1]] <- bind_cols(
+        read_tsv(input[[i]]),
+        read_tsv(benchmarks[[i]]),
+    )
+}
+bind_rows(temp) |> write_tsv("processed/benchmark/results.txt")

### 2025-05-08 scripts/simulate.benchmark.R

@@ -1,0 +1,51 @@
+library(dependentsimr)
+library(tidyverse)
+library(rlang)
+
+method <- snakemake@wildcards$method
+N_GENES <- snakemake@wildcards$n_genes
+N_SAMPLES <- 100
+
+# use ZT0 of raw time series data to get dependence structure
+male_read_counts <- read.csv("processed/Cortex_ZT0-counts.csv") |> 
+    select(!all_of("X331308_CRM55_WT_CTXR_26WKS_M_ZT0_L1.D704")) |>
+    column_to_rownames(var="raw_data.EnsemblID")
+
+selected_rows <- sample.int(nrow(male_read_counts), N_GENES)
+input <- male_read_counts[selected_rows,]
+## Generate Controls
+# With dependence
+time <- system.time({
+    actual_library_sizes <- input |> apply(2, sum)
+    # Get dependence structure
+    if (method == "pca") {
+        set.seed(1)
+        # Simulate with the PCA method
+        rs <- get_random_structure(list(data=input), method="pca", rank=2, types="DESeq2")
+    } else if (method == "corpcor") {
+        set.seed(2)
+        # Simulate with the corpcor method
+        rs <- get_random_structure(list(data=input), method="corpcor", types="DESeq2")
+    } else if (method == "wishart") {
+        set.seed(3)
+        # Simulate with the spiked Wishart method
+        rs <- get_random_structure(list(data=input), rank=10, method="spiked Wishart", types="DESeq2")
+    } else if (method == "indep") {
+        set.seed(4)
+        # Simulate without any dependence
+        dep_rs <- get_random_structure(list(data=input), method="pca", rank=2, type="DESeq2")
+        rs <- remove_dependence(dep_rs)
+    }
+
+    # Generate simulated data
+    library_sizes <- sample(actual_library_sizes / mean(actual_library_sizes), size=N_SAMPLES, replace=TRUE)
+    sim <- draw_from_multivariate_corr(rs, n_samples=N_SAMPLES, size_factors=library_sizes)$data
+})
+
+message("Elapsed time:")
+message(time[['elapsed']])
+tibble(
+    method = method,
+    n_genes = N_GENES,
+    time = time[['elapsed']],
+) |> write_tsv(paste0("processed/benchmark/", method, "/", N_GENES, ".txt"))

### 2025-05-09 manuscript/paper.qmd
Start rewrite of terminology

@@
 This multivariate normal distribution is parameterized by a $p \times p$ covariance matrix, $\Sigma$.
-In omics, choosing this $\Sigma$ presents a challenge since $n < p$.
+In omics, choosing the covariance matrix $\Sigma_{sim}$ to use during simulation presents a challenge since $n < p$.
 The most obvious choice is to use the sample covariance matrix $\widehat \Sigma$ of the reference data set (possibly after some transformation of the data).
-However, using $\widehat \Sigma$ is a poor choice for $\Sigma$.
-Indeed, we consider a choice of $\Sigma$ to be successful if data sets of $n$ samples simulated using $\Sigma$ have sample covaraince matrices that are similar to $\widehat \Sigma$.
-Counter-intuitively, data simulated using $\Sigma := \widehat \Sigma$ have sample covariance matrices that systematically and substantially differ from $\widehat \Sigma$, see @fig-intro.
+However, using $\widehat \Sigma$ is a poor choice for $\Sigma_{sim}$.
+Indeed, we consider a choice of $\Sigma_{sim}$ to be successful if data sets of $n$ samples simulated using $\Sigma_{sim}$ have sample covaraince matrices $\widehat \Sigma_{sim}$ that are similar to $\widehat \Sigma$.
+Counter-intuitively, data simulated using $\Sigma_{sim} := \widehat \Sigma$ have sample covariance matrices that systematically and substantially differ from $\widehat \Sigma$, see @fig-intro.
 
@@
 #| label: fig-intro
-#| fig-cap: "Eigenvalues of the sample covariance matrix $\\widehat \\Sigma$ when estimated from the reference data set and when estimated from simulated data using $\\Sigma$ equal to $\\widehat \\Sigma$ of the reference data set ('simulated').
-Eigenvalues are numbered by size.
-100 genes and 12 samples were used for both reference and simulated data sets.
-Both reference and simulated data follow a multivariate normal distribution."
+#| fig-cap: "Eigenvalues of the sample covariance matrix $\\widehat \\Sigma$ of reference data compared to $\\widehat \\Sigma_{sim}$ from from simulated data using $\\Sigma_{sim} = \\widehat \\Sigma$.
+100 genes and 12 samples were used for both reference and the 20 simulated data sets.
+Both reference and simulated data follow a multivariate normal distribution.
+Eigenvalues are numbered from largest to smallest."
 #| fig-width: 5
@@
 One reason this occurs is that $\widehat \Sigma$ is low rank, specifically at most rank $n - 1$, which is less than $p$ in typical omics studies.
-This limitation means that any data generated from the multivariate normal distribution of that $\Sigma$ will lie on a $n-1$-dimensional hyperplane.
+This means that any data generated using $\Sigma_{sim} = \widehat \Sigma$ will lie on a $n-1$-dimensional hyperplane.
 Therefore, newly simulated samples will actually be linear combination of the samples of the reference data set.
@@
 
-We present three methods - called here PCA, spiked Wishart, and corpcor - for making this choice of $\Sigma$, all which take the form
-$$ \Sigma = D^2 + PP^T $$
-where $D$ is a diagonal matrix capturing the injected independence and $PP^T$ is a low-rank correlation matrix derived from $\widehat \Sigma$.
-An additional advantage of this special form is that data may be generated very efficiently, see Methods, compared to arbitrary $\Sigma$ which typically require a Cholesky or eigenvalue decomposition, see Supplemental Figure S3.
+We present three methods - called here PCA, spiked Wishart, and corpcor - for making this choice of $\Sigma_{sim}$, all which take the form
+$$ \Sigma_{sim} = D^2 + UWU^T $$
+where $D$ is a diagonal matrix capturing the injected independence and $UWU^T$ is a low-rank correlation matrix derived from $\widehat \Sigma$.
+An additional advantage of this special form is that data may be generated very efficiently, Supplemental Figure S3 and see Methods, compared to arbitrary $\Sigma_{sim}$ which typically require a Cholesky or eigenvalue decomposition.
 
@@
 
-Below, we describe the three methods for selecting the components of the covariance matrix $\Sigma := D^2 + P P^T$.
+Below, we describe the three methods for selecting the components of the covariance matrix $\Sigma_{sim} := D^2 + P P^T$.
 
@@
 Specifically, let $u_1, \ldots, u_k$ be the left singular vectors of $Z$ with $\lambda_1, \ldots, \lambda_k$ the corresponding top $k$ signular values.
-This method computes $\Sigma$ such that $u_i^T \Sigma u_i = \lambda_i^2$, i.e. that the variance in the direction of $u_i$ exactly matches of the reference dataset's variance in that same direction.
+This method computes $\Sigma_{sim}$ such that $u_i^T \Sigma_{sim} u_i = \lambda_i^2$, i.e.
+that the variance in the direction of $u_i$ exactly matches of the reference dataset's variance in that same direction.
 One solution is to use the sample covariance matrix, but that is not full rank and would match for all $i \leq n$ instead of just $i \leq k$.
@@
 
-Steps 1 and 2 give that $u_i \Sigma u_i^T = \lambda_i$ for $i = 1, \ldots, k$.
-Step 4 ensures that $e_j \Sigma e_j^T = 1$ for $j = 1, \ldots, p$.
+Steps 1 and 2 give that $u_i \Sigma_{sim} u_i^T = \lambda_i$ for $i = 1, \ldots, k$.
+Step 4 ensures that $e_j \Sigma_{sim} e_j^T = 1$ for $j = 1, \ldots, p$.
 
@@
 The second method also makes use of PCA but has a different objective.
-If $n$ samples are drawn from $N(0, \Sigma)$ then we want the variances of their PCA components to match those of the reference dataset.
-Specifically, let $\lambda_1, \ldots, \lambda_{n-1}$ be the ${n-1}$ non-zero singular values of $Z$ ($\lambda_n$ is always approximately zero due to the normalization procedure), and let $\lambda'_1, \ldots, \lambda'_{n-1}$ be the singular values of $Z'$ where $Z'$ has $n-1$ columns each iid $N(0, \Sigma)$.
-Then we want to choose $\Sigma$ such that $E[\lambda'_i] = \lambda_i$ for each $i$, where $E[Y]$ denotes the expectation of the random variable $Y$.
+If $n$ samples are drawn from $N(0, \Sigma_{sim})$ then we want the variances of their PCA components to match those of the reference dataset.
+Specifically, let $\lambda_1, \ldots, \lambda_{n-1}$ be the ${n-1}$ non-zero singular values of $Z$ ($\lambda_n$ is always approximately zero due to the normalization procedure), and let $\lambda'_1, \ldots, \lambda'_{n-1}$ be the singular values of $Z'$ where $Z'$ has $n-1$ columns each iid $N(0, \Sigma_{sim})$.
+Then we want to choose $\Sigma_{sim}$ such that $E[\lambda'_i] = \lambda_i$ for each $i$, where $E[Y]$ denotes the expectation of the random variable $Y$.
 
 Since the distribution of the $\lambda'_i$ does not have a known analytic solution, we approximate this situation with the spiked Wishart distribution.
-The rank $k-1$ Wishart distribution is that of the sample covariance matrix of $Y$ where the $n$ columns of $Y$ are iid $N(0, \Sigma)$.
-The spiked Wishart is the special case where $\Sigma$ has $k$ arbitrary eigenvalues and the remaining are all equal to a constant.
+The rank $k-1$ Wishart distribution is that of the sample covariance matrix of $Y$ where the $n$ columns of $Y$ are iid $N(0, \Sigma_{sim})$.
+The spiked Wishart is the special case where $\Sigma_{sim}$ has $k$ arbitrary eigenvalues and the remaining are all equal to a constant.
 Note that the singular values of $Y$ are the square roots of the eigenvalues of its sample covariance matrix.
-While our case has $\Sigma$ non-diagonal, $\Sigma$ may be diagonalized by orthogonal rotations due the sepctral theorem, and orthogonal rotations do not change the singular values of $Y$.
-Therefore, the distribution of singular values is not affected by the assumption that $\Sigma$ is diagonal.
-Moreover, for the form $\Sigma = D^2 + U W U^T$ where $p$ is very large, each column of $U$ is typically very close to orthogonal to any $e_i$, a standard basis vector.
-Therefore, when $D = c^2I$ for some constant $c$, we can approximate $\Sigma$ as having $k$ arbitrary eigenvalues from $W$ and $n$ remaining eigenvalues all equal to $c$ corresponding to $D$.
+While our case has $\Sigma_{sim}$ non-diagonal, $\Sigma_{sim}$ may be diagonalized by orthogonal rotations due the sepctral theorem, and orthogonal rotations do not change the singular values of $Y$.
+Therefore, the distribution of singular values is not affected by the assumption that $\Sigma_{sim}$ is diagonal.
+Moreover, for the form $\Sigma_{sim} = D^2 + U W U^T$ where $p$ is very large, each column of $U$ is typically very close to orthogonal to any $e_i$, a standard basis vector.
+Therefore, when $D = c^2I$ for some constant $c$, we can approximate $\Sigma_{sim}$ as having $k$ arbitrary eigenvalues from $W$ and $n$ remaining eigenvalues all equal to $c$ corresponding to $D$.
 This is a spiked Wishart distribution.
@@
 Therefore, we use an efficient sampling and stochastic gradient descent method that we recently described [@wishart].
-Since the dataset has been normalized, $c$ will be close to one and $e_i \Sigma e_i^T \approx 1$.
+Since the dataset has been normalized, $c$ will be close to one and $e_i \Sigma_{sim} e_i^T \approx 1$.
 
@@
 2.
-Compute $w_1, \ldots, w_k$ and $c$ by stochastic gradient descent minimizing $\sum_{i} (E[\lambda'_i] - \lambda_i)^2$ for $\Sigma$ diagonal with entries $w_1, \ldots, w_k, c^2, \ldots, c^2$ [@wishart].
+Compute $w_1, \ldots, w_k$ and $c$ by stochastic gradient descent minimizing $\sum_{i} (E[\lambda'_i] - \lambda_i)^2$ for $\Sigma_{sim}$ diagonal with entries $w_1, \ldots, w_k, c^2, \ldots, c^2$ [@wishart].
 4. Set $W$ diagonal with the entries $w_1, \ldots, w_k$.
@@
 It then uses $\lambda_1$ to linearly interpolate the sample covariance matrix towards the identity matrix $I$ and $\lambda_2$ to interpolate the vector of variances towards the median variance value.
-Since the sample covariance matrix is rank at most $n$, we again obtain a matrix of the form $\Sigma = D^2 + UWU^T$.
+Since the sample covariance matrix is rank at most $n$, we again obtain a matrix of the form $\Sigma_{sim} = D^2 + UWU^T$.
 

### 2025-05-20 manuscript/paper.qmd
Revise intro and discussion

@@
 Omics data involves large numbers of measured features on each samples and these measures are generally correlated with each other.
-However, simulation too often ignores these correlations, perhaps due to the inconvenience and computational hurdles of doing so.
-To alleviate this, we describe in detail three approaches for quickly generating omics-scale data with correlated measures which mimic real data sets.
+However, simulation too often ignores these correlations, perhaps due to computational and statistical hurdles of doing so.
+To alleviate this, we describe three approaches for efficiently generating omics-scale data with correlated measures which mimic real datasets.
 These approaches all are based on a Gaussian copula approach with a covariance matrix that decomposes into a diagonal part and a low-rank part.
@@
 This creates dual challenges in generating realistic simulated data for the purposes of benchmarking.
-First, there isn’t enough data to be able to compute a dependence structure (e.g., a full-rank correlation matrix).
-Second, generating omics-scale data with a specified correlation matrix is slow due to the typical $O(n^3)$ nature of these algorithms, where $n$ is the number of measurements per sample.
+First, there is not enough data to be able to compute a meaningful dependence structure (e.g., a correlation matrix).
+Second, generating omics-scale data with a specified correlation matrix is slow due to the typical $O(p^3)$ nature of these algorithms, where $p$ is the number of measurements per sample.
 Moreover, there is a lack of practical guidance on how to generate simulated data with realistic dependence.
@@
 
-Here, we give an introduction to the theory and practice of generating dependent data and describe three related solutions which all offer good performance and ease-of-use even for large omics-scale problems.
+Here, we describe three related solutions which all offer good performance and ease-of-use even for large omics-scale problems.
 This expands off a discussion we originally wrote as part of a larger discussion on best practices in omics benchmarking [@Brooks2024].
-Our goal here is to produce guidelines that show that generating correlated data does not have to be onerous and instead should be considered a baseline requirement when simulating data.
+Our goal here is to show that generating correlated data does not have to be onerous and instead should be considered a baseline requirement when simulating data.
 
-We present three methods that operate by inferring a covariance matrix that decomposes into a diagonal part and a low-rank part.
 Using a Gaussian copula [@Nelsen1998-nq] approach (also referred to as NORTA, for "normal to anything" [@norta]), the marginal (univariate) distributions can have realistic forms.
-These solutions operate by taking a real dataset and mimicking it.
-For ease of use, we implement this in an R package which supports normal, Poisson, DESeq2-based (negative binomial with sample-specific size factors), and empirical (for ordinal data) marginal distributions.
+These solutions operate by taking a real, reference dataset and producing simulations that mimic it.
+However, a major challenge when simulating datasets is choice of the copula, which is a mathematical object that encodes the dependence between two or more variables.
+For Guassian copulas, this is a choice a covariance matrix.
+However, the sample covariance matrix of a reference dataset is a poor choice and present three alternative choices.
 
-We implemented three different strategies for determining the diagonal and low-rank parts of the covariance matrix.
+There is a large body of literature on determining covariance matrices for various purposes.
+Shrinkage methods [@SchaferStrimmer2005; @OpgenRheinStrimmer2007] improve estimates and guarantee that the estimated matrix is well-conditioned (and hence invertible).
+Sparse matrix methods [@Bien2011-nx, @Bickel_sparse, @Cai_adaptive] produce a matrices where many of the pairwise correlations are zero.
+Similarly, other methods produce a sparse estimate of the inverse of the covariance matrix [@sparse_inverse_covariance], which imparts pairwise independence of some variables conditional on all the others.
+
+Copula approaches are flexible and have a long history [@Sklar] of use in fields such as economics [@MacKenzie_copula] 
+More recently, copula approaches have been used for modelling or classification in multi-omics fields [@He_copula; @Ma_copula] and metagenomics [@Deek_microbiome].
+Projects simulating omics with copulas include SPsimSeq [@Assefa2020] and scDesign2 [@Sun2021].
+Some methods have employed the "vine copula" approach in the context of single-cell or spatial omics simulation [@fuetterer2019constructing; @scDesign3].
+This provides an alternative approach for choosing a copula that avoids the difficulty of choosing the covariance matrix, however it comes at significant computational cost when done at the omics-scale.
+SeqNet uses instead a network approach for gene dependence [@SeqNet].
+
+Here, we describe three different strategies for determining covariance matrices, all of which enable efficient generation of simulated data due to following a specific form.
+Specifically, these all decompose into a diagonal part and a low-rank part, which speeds up generation of random vectors.
 First, the 'PCA' method uses principal component analysis (PCA) and picks the low-rank part such that the simulated data has the same variance in the top $k$ PCA components of the reference dataset.
 Second, the 'spiked Wishart' method fits $k$ components such that simulations with the same number of samples as the reference dataset will have, on average, the same PCA component variances as the reference.
-Unlike 'PCA', these variances are computed with resepect to the simulated data's own PCA and not using the PCA weights of the reference dataset.
+Unlike 'PCA', these variances are computed with respect to the simulated data's own PCA and not using the PCA weights of the reference dataset.
 Third, the 'corpcor' method uses the popular R library `corpcor` [@SchaferStrimmer2005; @OpgenRheinStrimmer2007], which implements a James-Stein type shrinkage estimator for the covariance matrix as a linear interpolation of the sample covariance matrix and a diagonal matrix.
 No method exactly captures the input data, indicating room for future research, but all improve upon the common approach of assuming independence.
+For ease of use, we implement this in an R package which supports normal, Poisson, DESeq2-based (negative binomial with sample-specific size factors), and empirical (for ordinal data) marginal distributions.
 
@@
 
-We start with a reference data set $X$ given by an $p \times n$ data matrix of $p$ features measured in each of $n$ independent samples, which could be any real data set of interest.
-Our goal is to generate simulated data set that looks, at least on average, the same as this reference data set.
+We start with a reference dataset $X$ given by an $p \times n$ data matrix of $p$ features measured in each of $n$ independent samples, which could be any real dataset of interest.
+Our goal is to generate simulated dataset that looks, at least on average, the same as this reference dataset.
 To do this, we need to capture correlations between the $p$ features, which could represent gene expressions, protein abundances, or other measured values.
@@
 In omics, choosing the covariance matrix $\Sigma_{sim}$ to use during simulation presents a challenge since $n < p$.
-The most obvious choice is to use the sample covariance matrix $\widehat \Sigma$ of the reference data set (possibly after some transformation of the data).
-However, using $\widehat \Sigma$ is a poor choice for $\Sigma_{sim}$.
-Indeed, we consider a choice of $\Sigma_{sim}$ to be successful if data sets of $n$ samples simulated using $\Sigma_{sim}$ have sample covaraince matrices $\widehat \Sigma_{sim}$ that are similar to $\widehat \Sigma$.
-Counter-intuitively, data simulated using $\Sigma_{sim} := \widehat \Sigma$ have sample covariance matrices that systematically and substantially differ from $\widehat \Sigma$, see @fig-intro.
+The most obvious choice is to use the sample covariance matrix $\widehat \Sigma_{ref}$ of the reference dataset (possibly after some transformation of the data).
+However, using $\widehat \Sigma_{ref}$ is a poor choice for $\Sigma_{sim}$.
+Indeed, we consider a choice of $\Sigma_{sim}$ to be successful if datasets of $n$ samples simulated using $\Sigma_{sim}$ have sample covaraince matrices $\widehat \Sigma_{sim}$ that are similar to $\widehat \Sigma_{ref}$.
+Counter-intuitively, data simulated using $\Sigma_{sim} := \widehat \Sigma_{ref}$ have sample covariance matrices that systematically and substantially differ from $\widehat \Sigma_{ref}$, see @fig-intro.
 
@@
 #| label: fig-intro
-#| fig-cap: "Eigenvalues of the sample covariance matrix $\\widehat \\Sigma$ of reference data compared to $\\widehat \\Sigma_{sim}$ from from simulated data using $\\Sigma_{sim} = \\widehat \\Sigma$.
-100 genes and 12 samples were used for both reference and the 20 simulated data sets.
+#| fig-cap: "Eigenvalues of the sample covariance matrix $\\widehat \\Sigma_{ref}$ of reference data compared to $\\widehat \\Sigma_{sim}$ from from simulated data using $\\Sigma_{sim} = \\widehat \\Sigma$.
+100 genes and 12 samples were used for both reference and the 20 simulated datasets.
 Both reference and simulated data follow a multivariate normal distribution.
@@
 
-# create reference aka 'real' data set
+# create reference aka 'real' dataset
 temp <- matrix(rnorm(100*100), 100, 100)
@@
 
-One reason this occurs is that $\widehat \Sigma$ is low rank, specifically at most rank $n - 1$, which is less than $p$ in typical omics studies.
-This means that any data generated using $\Sigma_{sim} = \widehat \Sigma$ will lie on a $n-1$-dimensional hyperplane.
-Therefore, newly simulated samples will actually be linear combination of the samples of the reference data set.
-Intuitively, this means that the entire probability distribution of possible $p$ dimensional data have been squished into a $n-1$ dimensional space, generating much higher gene-gene correlations than in the reference data set.
+One reason this occurs is that $\widehat \Sigma_{ref}$ is low rank, specifically at most rank $n - 1$, which is less than $p$ in typical omics studies.
+This means that any data generated using $\Sigma_{sim} = \widehat \Sigma_{ref}$ will lie on a $n-1$-dimensional hyperplane.
+Intuitively, this means that the entire probability distribution of possible $p$ dimensional data have been squished into a $n-1$ dimensional space, generating much higher gene-gene correlations than in the reference dataset.
 In contrast, we know that omics contains at least some level of random noise that is approximately independent from one gene to the next.
 For example, RNA-seq involves a random sampling of fragments from the true sample of RNA molecules and other omics modes involve various background noise effects.
-We therefore need to "inject" some level of additional randomness that is independent between genes on top of the correlations captured by $\widehat \Sigma$
+We therefore need to "inject" some level of additional randomness that is independent between genes on top of the correlations captured by $\widehat \Sigma_{ref}$
 
 We present three methods - called here PCA, spiked Wishart, and corpcor - for making this choice of $\Sigma_{sim}$, all which take the form
-$$ \Sigma_{sim} = D^2 + UWU^T $$
-where $D$ is a diagonal matrix capturing the injected independence and $UWU^T$ is a low-rank correlation matrix derived from $\widehat \Sigma$.
+$$ \Sigma_{sim} = D^2 + UW^2U^T $$
+where $D$ is a diagonal matrix capturing the injected independence and $UW^2U^T$ is a low-rank correlation matrix derived from $\widehat \Sigma_{ref}$.
 An additional advantage of this special form is that data may be generated very efficiently, Supplemental Figure S3 and see Methods, compared to arbitrary $\Sigma_{sim}$ which typically require a Cholesky or eigenvalue decomposition.
@@
 
-We first discuss the simplest case, where our data set is multivariate normally distributed.
+We first discuss the simplest case, where our dataset is multivariate normally distributed.
 The distribution $N(\mu, \Sigma)$ is the multivariate normal distribution with mean vector $\mu$ and covariance matrix $\Sigma$.
@@
 
-The difficulty then is that we only know $\hat \Sigma$ from our reference data set, but we need to choose a $\Sigma$ with which we can simulate data and the obvious choice of $\hat \Sigma$ is inadequate.
+The difficulty then is that we only know $\hat \Sigma$ from our reference dataset, but we need to choose a $\Sigma$ with which we can simulate data and the obvious choice of $\hat \Sigma$ is inadequate.
 The most common choice is to assume $\Sigma$ is a diagonal matrix.
@@
 All three of these rely on a specific form of $\Sigma$, namely that
-$$ \Sigma = D^2 + PP^T $$
-where $D$ is a $p \times p$ diagonal matrix and $P$ is $p \times k$ for some $k \ll p$.
-This is a combination of an independent part (the diagonal matrix) and a low-rank part ($P P^T$).
+$$ \Sigma = D^2 + UW^2U^T $$
+where $D$ is a $p \times p$ diagonal matrix, $U$ is $p \times k$ for some $k \ll p$ and $W$ is a diagonal $k \times k$ matrix.
+This is a combination of an independent part (the diagonal matrix) and a low-rank part ($U W^2 U^T$ is rank at most $k$).
 The low-rank part is also simple to generate data for.
-If $x$ is a vector of $k$ independent univariate standard normal values, then $P x \sim N(0, P P^T)$.
-The choice of specific $D$ and $P$ matrices is more in-depth and we leave the details for the Methods section.
+If $x$ is a vector of $k$ independent univariate standard normal values, then $U W x \sim N(0, P P^T)$.
+The choice of specific $D$, $U$, $W$ matrices is more in-depth and we leave the details for the Methods section.
 We have three alternative means for doing so, which we refer to as the PCA, spiked Wishart, and corpcor methods.
@@
 
-Here, we describe the approach using the form of covariance matrix $\Sigma = D^2 + PP^T$ as above.
+Here, we describe the approach using the form of covariance matrix $\Sigma = D^2 + UW^2U^T$ as above.
 Once data is obtained $Z \sim N(0, \Sigma)$, then one can undo the normalization process to obtain data with the same marginal distributions as the fit marginal distributions but with dependence determined by $\Sigma$.
@@
 
-The generated data $Z'$ has covariance matrix $\Sigma = D^2 + U W U^T = D^2 + P P^T$, where $P := U \sqrt{W}$.
-Moreover, we require that $\Sigma$ satisfies that $e_i \Sigma e_i^T$ is approximately 1.
+The generated data $Z'$ has covariance matrix $\Sigma_{sim} = D^2 + U W U^T = D^2 + U W^2 U^T$.
+Moreover, we require that $\Sigma$ satisfies that $(\Sigma_{sim})_{ii}$ is approximately 1.
 That guarantees that the output $X'$ has each entry with the same marginal distributions $F_i$ as was originally fit and inherits gene-gene dependence from $Z'$.
@@
 
-To compare the three simulation methods with a real data set, we chose as a references data set 12 mouse cortex RNA-seq samples from accession GSE151923 [@Wang2022].
+To compare the three simulation methods with a real dataset, we chose as a references dataset 12 mouse cortex RNA-seq samples from accession GSE151923 [@Wang2022].
 We then simulated data mimicking this reference using all three simulation methods (PCA, spiked Wishart, and corpcor) as well as a simulation with independent genes.
+All data were simulated with negative binomial distributions fit by DESeq2.
 We repeated the simulations, each of 12 samples, a total of 8 times to estimate variance.
@@
 Simulated data captures the genes' mean and variance accurately (@fig-compare-to-real a-b).
-Next, we compared to the real data set when projected onto the top two principal components of the real data set (@fig-compare-to-real c).
+Next, we compared to the real dataset when projected onto the top two principal components of the real dataset (@fig-compare-to-real c).
 The simulations with dependence are distributed around the entire space like the real data, but the independent simulations have unrealistically low variance in these components, clustering tightly around the origin.
 
-Then, we computed the gene-gene correlation on pairs of high-expressed genes (at least 100 mean reads in the real data set).
+Then, we computed the gene-gene correlation on pairs of high-expressed genes (at least 100 mean reads in the real dataset).
 The simulation with independence showed the least levels of gene-gene correlations (@fig-compare-to-real d).
-However, the PCA method overshot the reference data set and the spiked Wishart and corpcor methods only slightly improved upon the independent simulation.
+However, the PCA method overshot the reference dataset and the spiked Wishart and corpcor methods only slightly improved upon the independent simulation.
 The SPsimSeq simulator performed similarly to the PCA method.
 
-Lastly, we compared the variances of principal components on each data set (@fig-compare-to-real e).
-These were computed separately for each data set, unlike (@fig-compare-to-real c) which used the reference data set's PCA weights for all data sets.
-The independent data has much lower variance than the real data set in the top four principal components.
-The spiked Wishart method comes closest to the real data set, as it optimizes for fitting these values.
+Lastly, we compared the variances of principal components on each dataset (@fig-compare-to-real e).
+These were computed separately for each dataset, unlike (@fig-compare-to-real c) which used the reference dataset's PCA weights for all datasets.
+The independent data has much lower variance than the real dataset in the top four principal components.
+The spiked Wishart method comes closest to the real dataset, as it optimizes for fitting these values.
 Surprisingly, the corpcor method performs only somewhat better than the independent method.
@@
 #| label: fig-compare-to-real
-#| fig-cap: "Comparison to real data run on a mouse cortex data set from GSE151923.
+#| fig-cap: "Comparison to real data run on a mouse cortex dataset from GSE151923.
 (a-b) Comparison of gene (a) mean expression and (b) variance, log-scaled in real and PCA simulated data.
@@
 Genes with at least 100 reads were used.
-Values on the diagonal line indicate a match between the simulated and real data sets.
-(d) Projections onto the top two principal components of the real data set for both real and simulated data.
+Values on the diagonal line indicate a match between the simulated and real datasets.
+(d) Projections onto the top two principal components of the real dataset for both real and simulated data.
 All 8 simulations (96 samples for each simulation) shown.
-(e) Principal component analysis was performed on all data sets and the variance captured by the top components is shown.
-Unlike (d), these components were fit from each data set considered separately instead of reusing the weights from the real data."
+(e) Principal component analysis was performed on all datasets and the variance captured by the top components is shown.
+Unlike (d), these components were fit from each dataset considered separately instead of reusing the weights from the real data."
 #| fig-width: 10
@@
 
-We benchmarked DESeq2 [@Love2014], a popular differential expression analysis tool, using data sets simulated with dependence and ones simulated without dependence to compare its performances on both.
+We benchmarked DESeq2 [@Love2014], a popular differential expression analysis tool, using datasets simulated with dependence and ones simulated without dependence to compare its performances on both.
 DESeq2 presents an interesting case because several aspects of it assume independence of genes and so may be adversely affected by gene-gene dependence.
@@
 
-We used a fly whole body RNA-Seq data set GSE81142, see Supplemental Figure 1, and selected samples of male flies without treatment and after at least 2 hours of feeding to simulate 5 "control" samples.
-Unnormalized read counts were used as the reference data set and all simulations used the DESeq2-based negative binomial model.
+We used a fly whole body RNA-Seq dataset GSE81142, see Supplemental Figure 1, and selected samples of male flies without treatment and after at least 2 hours of feeding to simulate 5 "control" samples.
+Unnormalized read counts were used as the reference dataset and all simulations used the DESeq2-based negative binomial model.
 We then randomly selected 5% of the genes to be differentially expressed, with absolute $\log_2$ fold change uniformly distributed between $0.2$ and $2.0$, either up or down regulated chosen randomly, and simulated 5 "experimental" samples.
@@
 Finally, we ran DESeq2 on each simulated 5 vs 5 experiment and compared the output FDR with the true percentages of genes that are differential expressed (@fig-DESeq2 a-d).
-We observed that DESeq2 is anti-conservative on all data sets, with similar mean true FDRs for each estimated FDR cutoff.
+We observed that DESeq2 is anti-conservative on all datasets, with similar mean true FDRs for each estimated FDR cutoff.
 However, we note that this anti-conservative nature is most present at very low FDR thresholds, was typically associated with only a single digit number of false positive genes.
-While that mean behavior was consistant whether or not the simulation included gene-gene dependence, simulations with dependence showed a substantially greater variance in the performance of DESeq2.
+While that mean behavior was consistent whether or not the simulation included gene-gene dependence, simulations with dependence showed a substantially greater variance in the performance of DESeq2.
 
-To demonstrate the application of our simulation method for another organism, we also simulated data sets using the mouse cortex data set GSE151923 [@Wang2022] and selected samples from male mice.
-We then simulated differential expression experiments as above and observed a similar result (@fig-DESeq2 e-f) as for the fly whole body data sets.
+To demonstrate the application of our simulation method for another organism, we also simulated datasets using the mouse cortex dataset GSE151923 [@Wang2022] and selected samples from male mice.
+We then simulated differential expression experiments as above and observed a similar result (@fig-DESeq2 e-f) as for the fly whole body datasets.
 
@@
 #| fig-cap: "Performance of DESeq2 on simulated datasets.
-(a-d) Comparison of true false discovery proportions and DESeq2 reported False Discovery Rates, plotted on a log scale, for data sets simulated from the fly whole body data set (GSE81142), (a) without dependence, (b) using PCA, (c) using Wishart and (d) using corpcor.
+(a-d) Comparison of true false discovery proportions and DESeq2 reported False Discovery Rates, plotted on a log scale, for datasets simulated from the fly whole body dataset (GSE81142), (a) without dependence, (b) using PCA, (c) using Wishart and (d) using corpcor.
 Diagonal line represents perfect estimation of FDR.
-(e-h) Comparison of true false discovery proportions and DESeq2 reported FDR for data sets simulated from the mouse cortex data set (GSE151923), (e) without dependence, (f) using PCA, (g) using Wishart and (h) using corpcor."
+(e-h) Comparison of true false discovery proportions and DESeq2 reported FDR for datasets simulated from the mouse cortex dataset (GSE151923), (e) without dependence, (f) using PCA, (g) using Wishart and (h) using corpcor."
 #| fig-width: 12
@@
 We next used our simulation method to benchmark CYCLOPS [@Anafi2017], which infers relative times for a set of unlabeled samples using an autoencoder to identify circular structures.
-We chose a mouse cortex time series data set GSE151565, see Supplemental Figure 2, which contains a total of 77 samples every 3 hours, for 36 hours.
-As before, unnormalized read counts were used as the reference data set and all simulations used the DESeq2-based negative binomial model.
+We chose a mouse cortex time series dataset GSE151565, see Supplemental Figure 2, which contains a total of 77 samples every 3 hours, for 36 hours.
+As before, unnormalized read counts were used as the reference dataset and all simulations used the DESeq2-based negative binomial model.
 We computed the dependence structure of the genes as well as the variances of marginal distributions using the 11 time point 0 samples after removing one outlier and computed the means of gene expressions at each time point.
-We then used these to simulate 20 time series data sets for each of the independent, PCA, Wishart, and corpcor simulation methods.
+We then used these to simulate 20 time series datasets for each of the independent, PCA, Wishart, and corpcor simulation methods.
 
-We ran CYCLOPS on each data set with a list of cyclic mouse genes (from [@Zhang2014], JTK p-value $<0.05$), which yielded an estimated relative time for each sample.
+We ran CYCLOPS on each dataset with a list of cyclic mouse genes (from [@Zhang2014], JTK p-value $<0.05$), which yielded an estimated relative time for each sample.
 We evaluated CYCLOPS' performance compared to true circadian time using the circular correlation [@correlation_coeff], defined as follows:
@@
 At that setting, CYCLOPS performance is much higher in the PCA method and moderately improved in Wishart method, compared to the independent method. 
-This difference is likely driven by the difference in the number of eigengenes used (@fig-CYCLOPS b), which is a measure of how much dependence is present in the data set.
+This difference is likely driven by the difference in the number of eigengenes used (@fig-CYCLOPS b), which is a measure of how much dependence is present in the dataset.
 This demonstrates that the correlational structure of the transcriptome can have a major impact on performance.
@@
 #| label: fig-CYCLOPS
-#| fig-cap: "Performance of CYCLOPS on simulated time series data sets based on mouse cortex data set (GSE151565), using eigengenes of at least 2% variance.
-(a) Absolute circular correlations between true phases and CYCLOPS estimated phases on the simulated data sets.
+#| fig-cap: "Performance of CYCLOPS on simulated time series datasets based on mouse cortex dataset (GSE151565), using eigengenes of at least 2% variance.
+(a) Absolute circular correlations between true phases and CYCLOPS estimated phases on the simulated datasets.
 (b) The number of eigengenes used by CYCLOPS; the dotted line indicates the number of eigengenes used by CYCLOPS on the real data (5).
-(c-f) Examples of CYCLOPS estimated phases on the simulated data sets.
+(c-f) Examples of CYCLOPS estimated phases on the simulated datasets.
 CYCLOPS shows good performance when it separates out points by color (true circadian time)."
@@
 
-Below, we describe the three methods for selecting the components of the covariance matrix $\Sigma_{sim} := D^2 + P P^T$.
+Below, we describe the three methods for selecting the components of the covariance matrix $\Sigma_{sim} := D^2 + U W^2 U^T$.
 
@@
 Since the distribution of the $\lambda'_i$ does not have a known analytic solution, we approximate this situation with the spiked Wishart distribution.
-The rank $k-1$ Wishart distribution is that of the sample covariance matrix of $Y$ where the $n$ columns of $Y$ are iid $N(0, \Sigma_{sim})$.
+The rank $k-1$ Wishart distribution is the distribution of sample covariance matrices of $Z$ whose $n$ columns of $Z$ are iid $N(0, \Sigma_{sim})$.
 The spiked Wishart is the special case where $\Sigma_{sim}$ has $k$ arbitrary eigenvalues and the remaining are all equal to a constant.
-Note that the singular values of $Y$ are the square roots of the eigenvalues of its sample covariance matrix.
-While our case has $\Sigma_{sim}$ non-diagonal, $\Sigma_{sim}$ may be diagonalized by orthogonal rotations due the sepctral theorem, and orthogonal rotations do not change the singular values of $Y$.
+Note that the singular values of $Z$ are the square roots of the eigenvalues of $\widehat \Sigma_{ref}.
+While our case has $\Sigma_{sim}$ non-diagonal, $\Sigma_{sim}$ may be diagonalized by orthogonal rotations due the sepctral theorem, and orthogonal rotations do not change the singular values of $Z$.
 Therefore, the distribution of singular values is not affected by the assumption that $\Sigma_{sim}$ is diagonal.
-Moreover, for the form $\Sigma_{sim} = D^2 + U W U^T$ where $p$ is very large, each column of $U$ is typically very close to orthogonal to any $e_i$, a standard basis vector.
-Therefore, when $D = c^2I$ for some constant $c$, we can approximate $\Sigma_{sim}$ as having $k$ arbitrary eigenvalues from $W$ and $n$ remaining eigenvalues all equal to $c$ corresponding to $D$.
+Moreover, for the form $\Sigma_{sim} = D^2 + U W^2 U^T$ where $p$ is very large, each column of $U$ is typically very close to orthogonal to any $e_i$, a standard basis vector.
+Therefore, when $D = c^2I$ for some constant $c$, we can approximate $\Sigma_{sim}$ as having $k$ arbitrary eigenvalues from $W^2$ and $n$ remaining eigenvalues all equal to $c^2$ corresponding to $D$.
 This is a spiked Wishart distribution.
@@
 2.
-Compute $w_1, \ldots, w_k$ and $c$ by stochastic gradient descent minimizing $\sum_{i} (E[\lambda'_i] - \lambda_i)^2$ for $\Sigma_{sim}$ diagonal with entries $w_1, \ldots, w_k, c^2, \ldots, c^2$ [@wishart].
+Compute $w_1, \ldots, w_k$ and $c$ by stochastic gradient descent minimizing $\sum_{i} (E[\lambda'_i] - \lambda_i)^2$ for $\Sigma_{sim}$ diagonal with entries $w_1^2, \ldots, w_k^2, c^2, \ldots, c^2$ [@wishart].
 4. Set $W$ diagonal with the entries $w_1, \ldots, w_k$.
-3. Set $D = cI$
+3. Set $D = cI$.
 
@@
 It then uses $\lambda_1$ to linearly interpolate the sample covariance matrix towards the identity matrix $I$ and $\lambda_2$ to interpolate the vector of variances towards the median variance value.
-Since the sample covariance matrix is rank at most $n$, we again obtain a matrix of the form $\Sigma_{sim} = D^2 + UWU^T$.
+Since the sample covariance matrix is rank at most $n$, we again obtain a matrix of the form $\Sigma_{sim} = D^2 + UW^2U^T$.
 
@@
 We described the well-known Gaussian copula approach and recommended a specific form of covariance matrix which is well tailored to omics data simulation.
-We developed three methods using this form of covariance matrix which can be used to mimic a reference data set for simulation.
+We developed three methods using this form of covariance matrix which can be used to mimic a reference dataset for simulation.
 All of these methods use a multivariate normal distribution as an intermediate step and therefore substantially restrict the kinds of dependence that can be simulated.
@@
 To encourage adoption of dependence in simulated omics data, we developed `dependentsimr`, an R package that generates omics-scale data with realistic correlation.
-This implementation is efficient and simple, requiring just two lines of code to fit a model to a reference data set and then simulate data from it.
+This implementation is efficient and simple, requiring just two lines of code to fit a model to a reference dataset and then simulate data from it.
 We demonstrated this package on RNA-seq data, using the DESeq2 method to fit negative binomial marginal distributions.
@@
 
+We highlight that our methods are highly efficient and easily scale to the large feature counts typical of omics data sets.
+While runtime of simulators is typically not critical, we hypothesize that it introduces an inconvenience that has slowed the adoption of dependence in omics simulation and could be particularly cumbersome whenever large numbers of simulated datasets are needed.
+Of note is that alternative methods are typically run with only a subset of genes chosen, for example, 500-200 genes for scDesign2 [@Sun2021], 1000 genes for scDesign3 [@scDesign3], and 621 for SeqNet [@SeqNet].
+In contrast, our methods can operate on the entire space of genes, avoiding an arbitrary choice of cutoff, though it should be noted that correlation is most meaningful in high-expressed, high variance genes so subsetting genes is also a reasonable approach.
+
 We demonstrated the importance of including gene-gene dependence in simulated data by two application benchmarks.
 In the first, DESeq2 results were substantially more variable when simulating with gene-gene dependence.
-This indicates that DESeq2's had an increased chance of a larger simultaneous number of false positives, as well as increased chance of fewer false positives, when data included realistic dependence.
+This indicates that DESeq2 had an increased chance of a larger simultaneous number of false positives, as well as increased chance of fewer false positives, when data included realistic dependence.
 We therefore recommend that benchmarking of differential expression methods should include gene-gene dependence whenever possible.
@@
 
-### Alternatives
-
-We highlight some alternative approaches and software packages that have been taken below.
-
-The R package [SPsimSeq](https://github.com/CenterForStatistics-UGent/SPsimSeq) [@Assefa2020] provides a dedicated RNA-seq and single-cell RNA-seq simulator using a Gaussian copula approach to simulate gene dependence.
-In contrast to this package, it uses a two-step randomization.
+We compared our simulated data to that of the R package SPsimSeq [@Assefa2020].
+In contrast to our package, SPsimSeq uses a two-step randomization.
 In short, it first fits distributions to each genes and then bins those distributions into discrete buckets.
-Then, new data is generated by drawing from a multivariate normal distribution whose covariance matrix equals the sample covariance matrix of the reference data set.
+Then, new data is generated by drawing from a multivariate normal distribution whose covariance matrix equals the sample covariance matrix of the reference dataset.
 These values are used to choose which bucket each gene is drawn from.
 Then, each gene's final value is drawn uniformly and independently from within its chosen bucket.
-Therefore, SPsimSeq induces correlation in the first multivarainte normal draw, but then injects additional independence in the second uniform step.
+Therefore, SPsimSeq induces correlation in the first multivariate normal draw, but then injects additional independence in the second uniform step.
 This two-step process makes it challenging to compare on theoretical grounds to our proposed methods since it will dependend upon parameters such as the number of buckets used.
-More fine-grained buckets will induce higher correlation, and in the limit of infinite buckets, it should approach our PCA method with $k$ equal to the number of samples in the reference data set.
-`SPsimSeq` is more specialized and full-featured for RNA-seq simulation, providing, for example, native differential expression (DE) options,.
-In comparison, our `dependentsimr` package requires manually setting marginal expression values to inject DE, but also supports other marginal distributions for situations outside of RNA-seq.
+More fine-grained buckets will induce higher correlation, and in the limit of infinite buckets, it should approach our PCA method with $k$ equal to the number of samples in the reference dataset.
+SPsimSeq is more specialized and full-featured for RNA-seq simulation, providing, for example, native differential expression (DE) options,.
+In comparison, our dependentsimr package requires manually setting marginal expression values to inject DE, but also supports other marginal distributions for situations outside of RNA-seq.
 
-The [scDesign2](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-021-02367-2) simulator [@Sun2021] for single-cell RNA-seq also uses Gaussian copula and, like our method, uses the approach of estimating the correlation matrix from the normalized dataset.
-However, it limits this correlation matrix to top-expressed genes.
-Since correlation is most discernible in high-expressed genes, this approach is reasonable but requires making certain arbitrary cutoffs that our methods avoid.
-
-Other Gaussian copula-based R packages that may be applicable, at least for datasets with smaller numbers of features, include `bindata`, `GenOrd`, and `SimMultiCorrData`, the last of these being the most comprehensive.
+Other Gaussian copula-based R packages include `bindata`, `GenOrd`, and `SimMultiCorrData`, the last of these being the most comprehensive.
 The `bigsimr` package provides faster implementations of these methods to scale up to omics-level data.
@@
 Source code for the `dependentsimr` package is available at [github.com/tgbrooks/dependent_sim](https://github.com/tgbrooks/dependent_sim/).
-All data used is available from the Gene Expression Omnibus (GEO) with accession numbers GSE151923, GSE81142, GSE151565.
+All data used is available from the Gene Expression Omnibus (GEO) with accession numbers GSE151923, GSE81142, and GSE151565.
 

### 2025-05-21 manuscript/paper.qmd
Add benchmarking of vine copula method
And various minor edits in manuscript

@@
 In order for this to work, $\Sigma$ must be symmetric and positive semi-definite, meaning that all of its eigenvalues are non-negative. 
-This matrix is conceptually simple, since $\Sigma_{ij}$ gives the the covariance of $x_i$ and $x_j$ when $x \sim N(\mu, \Sigma)$.
+This matrix is conceptually simple, since $\Sigma_{ij}$ gives the covariance of $x_i$ and $x_j$ when $x \sim N(\mu, \Sigma)$.
 More specifically, this is the population covariance matrix of $N(\mu, \Sigma)$, which does not generally equal the sample covariance matrix.
@@
 This difference is likely driven by the difference in the number of eigengenes used (@fig-CYCLOPS b), which is a measure of how much dependence is present in the dataset.
-This demonstrates that the correlational structure of the transcriptome can have a major impact on performance.
+This demonstrates that the correlation structure of the transcriptome can have a major impact on performance.
 
@@
 The first of our three methods attempts to match the top $k$ PCA components of the normalized reference dataset $Z$.
-Specifically, let $u_1, \ldots, u_k$ be the left singular vectors of $Z$ with $\lambda_1, \ldots, \lambda_k$ the corresponding top $k$ signular values.
+Specifically, let $u_1, \ldots, u_k$ be the left singular vectors of $Z$ with $\lambda_1, \ldots, \lambda_k$ the corresponding top $k$ singular values.
 This method computes $\Sigma_{sim}$ such that $u_i^T \Sigma_{sim} u_i = \lambda_i^2$, i.e.
@@
 Note that the singular values of $Z$ are the square roots of the eigenvalues of $\widehat \Sigma_{ref}.
-While our case has $\Sigma_{sim}$ non-diagonal, $\Sigma_{sim}$ may be diagonalized by orthogonal rotations due the sepctral theorem, and orthogonal rotations do not change the singular values of $Z$.
+While our case has $\Sigma_{sim}$ non-diagonal, $\Sigma_{sim}$ may be diagonalized by orthogonal rotations due the spectral theorem, and orthogonal rotations do not change the singular values of $Z$.
 Therefore, the distribution of singular values is not affected by the assumption that $\Sigma_{sim}$ is diagonal.
@@
 Therefore, SPsimSeq induces correlation in the first multivariate normal draw, but then injects additional independence in the second uniform step.
-This two-step process makes it challenging to compare on theoretical grounds to our proposed methods since it will dependend upon parameters such as the number of buckets used.
+This two-step process makes it challenging to compare on theoretical grounds to our proposed methods since it will depended upon parameters such as the number of buckets used.
 More fine-grained buckets will induce higher correlation, and in the limit of infinite buckets, it should approach our PCA method with $k$ equal to the number of samples in the reference dataset.

### 2025-05-21 manuscript/supplemental/supplemental.qmd
Add benchmarking of vine copula method
And various minor edits in manuscript

--- 2025-05-05 manuscript/supplemental/supplemental.qmd
@@
 
+### Benchmark
+```{r}
+#| label: fig-benchmark
+#| fig-cap: "Comparison of (a) run time and (b) memory usage for our methods, SPsimSeq, and vine copula (implemented by the rvinecopulib R package).
+100 samples were generated with a varying number of genes.
+GSE151564 dataset was randomly subsetted to have the required number of genes to be used as the reference data set.
+Vine copula method was only run up to 2000 genes due to the memory use required at higher gene counts."
+#| fig-width: 8
+#| fig-height: 4
+library(tidyverse)
+library(patchwork)
+METHOD_ORDER <- c("real", "indep", "pca", "wishart", "corpcor", "SPsimSeq", "vinecopula")
+METHOD_COLORS <- c("black", RColorBrewer::brewer.pal(length(METHOD_ORDER)-1, "Dark2"))
+benchmark <- read_tsv("../../processed/benchmark/results.txt") |>
+    filter(!is.na(time)) # drop the ones that aren't run completely due to time/memory constraints
+
+g1 <- ggplot(benchmark, aes(x=n_genes, y=time, color=method)) +
+    geom_line(linewidth=2) +
+    labs(x = "N genes", y = "time (seconds)") + 
+    scale_color_manual(values = METHOD_COLORS, breaks = METHOD_ORDER, name="method") +
+    scale_x_log10() +
+    scale_y_log10()
+g2 <- ggplot(benchmark, aes(x=n_genes, y=max_rss, color=method)) +
+    geom_line(linewidth=2) +
+    labs(x = "N genes", y = "memory (MB)") +
+    scale_color_manual(values = METHOD_COLORS, breaks = METHOD_ORDER, name="method") +
+    scale_x_log10() +
+    scale_y_log10()
+g1 + g2 + plot_annotation(tag_levels="a") + plot_layout(guides = "collect")
+```
+
 ## Supplemental Methods
@@
 
-The R package [SPsimSeq](https://github.com/CenterForStatistics-UGent/SPsimSeq) [@Assefa2020] provides a dedicated RNA-seq and single-cell RNA-seq simulator using a Gaussian copula approach to simulate gene dependence.
+The R package SPsimSeq [@Assefa2020] provides a dedicated RNA-seq and single-cell RNA-seq simulator using a Gaussian copula approach to simulate gene dependence.
 We ran SPsimSeq using the following options:
@@
 SPsimSeq does not report either the group mean values or the log fold changes used for these genes, so we are not able to modify our own simulation to have the same set of DE genes at the same effect sizes.
+
+### Running vine copula
+
+We used the rvinecopulib package to fit and generate random data from a vine copula distribution.
+This package is a R wrapper of the high-performance vinecopulib C++ package.
+This package expects data normalized to the range of \[0, 1\].
+Therefore, we chose not to use the real RNA-seq count data and instead generated correlated random number as input which follow a multivariate normal distribution with covariance matrix constructed with a randomly generated singular vector matrix and singular values 1, 2, ..., 100.
+These values were then normalized to the interval \[0,1\] by the `pnorm` function.
+A vine copula was fit using just the "guassian" family.
+While rvinecopula supports multi-threaded operation, we ran using a single thread as with the other methods.

### 2025-05-21 scripts/vinecopulib.benchmark.R
Add benchmarking of vine copula method
And various minor edits in manuscript

@@ -1,0 +1,42 @@
+library(rvinecopulib)
+library(tidyverse)
+N_GENES <- strtoi(snakemake@wildcards$n_genes)
+N_SAMPLES <- 100
+method <- "vinecopula"
+
+# Generate some very basic data
+X <- runif(N_GENES * N_SAMPLES) |> matrix(N_GENES, N_SAMPLES)
+udv <- svd(X)
+true_d <- seq(N_SAMPLES, 1)
+U <- udv$u
+temp <- rnorm(length(true_d) * N_SAMPLES) |> matrix(true_d, N_SAMPLES)
+data <- U %*% diag(true_d) %*% temp
+norm_data <- pnorm(data) |> t()
+print(norm_data |> dim())
+
+if (N_GENES > 2000) {
+    # Don't run slow ones
+    tibble(
+        method = method,
+        n_genes = N_GENES,
+        time = NA,
+    ) |> write_tsv(paste0("processed/benchmark/", method, "/", N_GENES, ".txt"))
+} else {
+    set.seed(1)
+    time <- system.time({
+        # Fit a vinecopula model to it
+        vc <- rvinecopulib::vinecop(norm_data, family_set = "gaussian")
+
+        # Simulate data using that vinecopula
+        sim_data <- rvinecopulib::rvinecop(vc, n= N_SAMPLES)
+    })
+    print(sim_data |> dim())
+    message("Elapsed time:")
+    message(time[['elapsed']])
+    tibble(
+        method = method,
+        n_genes = N_GENES,
+        time = time[['elapsed']],
+    ) |> write_tsv(paste0("processed/benchmark/", method, "/", N_GENES, ".txt"))
+
+}

### 2025-05-23 manuscript/paper.qmd
Add mvrnorm and diagram and many revisions

@@
 Our methods are based off the well-known Gaussian copula method, which uses a multivariate normal distribution to capture aspects of the gene-gene dependence, without necessarily assuming that the data is normally distributed.
+See @fig-intro a for a high-level overview.
 This multivariate normal distribution is parameterized by a $p \times p$ covariance matrix, $\Sigma$.
@@
 Indeed, we consider a choice of $\Sigma_{sim}$ to be successful if datasets of $n$ samples simulated using $\Sigma_{sim}$ have sample covaraince matrices $\widehat \Sigma_{sim}$ that are similar to $\widehat \Sigma_{ref}$.
-Counter-intuitively, data simulated using $\Sigma_{sim} := \widehat \Sigma_{ref}$ have sample covariance matrices that systematically and substantially differ from $\widehat \Sigma_{ref}$, see @fig-intro.
+Counter-intuitively, data simulated using $\Sigma_{sim} := \widehat \Sigma_{ref}$ have sample covariance matrices that systematically and substantially differ from $\widehat \Sigma_{ref}$, see @fig-intro b.
 
@@
 #| label: fig-intro
-#| fig-cap: "Eigenvalues of the sample covariance matrix $\\widehat \\Sigma_{ref}$ of reference data compared to $\\widehat \\Sigma_{sim}$ from from simulated data using $\\Sigma_{sim} = \\widehat \\Sigma$.
+#| fig-cap: "(a) Schematic of our Gaussian copula-based approach.
+(b) Eigenvalues of the sample covariance matrix $\\widehat \\Sigma_{ref}$ of reference data compared to $\\widehat \\Sigma_{sim}$ from simulated data using $\\Sigma_{sim} = \\widehat \\Sigma_{ref}$.
 100 genes and 12 samples were used for both reference and the 20 simulated datasets.
 Both reference and simulated data follow a multivariate normal distribution.
-Eigenvalues are numbered from largest to smallest."
-#| fig-width: 5
-#| fig-height: 4
+Eigenvalues are numbered from largest to smallest.
+(c-d) Comparison of (c) run time and (d) memory usage for our methods, SPsimSeq, multivariate normal sampling (using the mvrnorm function from the MASS package), and vine copula (implemented by the rvinecopulib R package).
+100 samples were generated with a varying number of genes.
+A mouse cortex dataset GSE151564 was randomly subsetted to have the required number of genes to be used as the reference data set.
+Vine copula method was only run up to 2000 genes due to the memory use required at higher gene counts."
+#| fig-width: 8
+#| fig-height: 5
 
@@
 library(tidyverse)
+library(patchwork)
 
@@
 # Plot the eigenvalues
-ggplot(eigenvalue_data |> filter(type != "reference"), aes(x=number, y = eigenvalue, color = type, group=iter)) +
+gsigma <- ggplot(eigenvalue_data |> filter(type != "reference"), aes(x=number, y = eigenvalue, color = type, group=iter)) +
   geom_line(linewidth=2, alpha=0.5) +
@@
   labs(x = "eigenvalue number", y = "eigenvalue")
+
+# diagram
+gdiagram <- ggplot() +
+    annotation_custom(grid::rasterGrob(png::readPNG("diagram.png")), xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf) +
+    theme(panel.background = element_blank())
+
+# Plot benchmarking
+METHOD_ORDER <- c("real", "indep", "pca", "wishart", "corpcor", "SPsimSeq", "vinecopula", "mvrnorm")
+METHOD_COLORS <- c("black", RColorBrewer::brewer.pal(length(METHOD_ORDER)-1, "Dark2"))
+benchmark <- read_tsv("../processed/benchmark/results.txt") |>
+    filter(method != "indep") |> # We don't want to compare indep for this, whose runtime hasn't been optimized anyway
+    filter(!is.na(time)) # drop the ones that aren't run completely due to time/memory constraints
+
+g1 <- ggplot(benchmark, aes(x=n_genes, y=time, color=method)) +
+    geom_line(linewidth=2) +
+    labs(x = "N genes", y = "time (seconds)") + 
+    scale_color_manual(values = METHOD_COLORS, breaks = METHOD_ORDER, name="method") +
+    scale_x_log10() +
+    scale_y_log10() +
+    coord_cartesian(ylim=c(3,10000))
+g2 <- ggplot(benchmark, aes(x=n_genes, y=max_vms / 1000, color=method)) +
+    geom_line(linewidth=2) +
+    labs(x = "N genes", y = "memory (GB)") +
+    scale_color_manual(values = METHOD_COLORS, breaks = METHOD_ORDER, name="method") +
+    scale_x_log10() +
+    scale_y_log10() +
+    coord_cartesian(ylim=c(0.600,100))
+
+(gdiagram) / (gsigma + g1 + g2 + plot_layout(guides = "collect")) + plot_annotation(tag_levels="a") + plot_layout(heights=c(1,1.5))
 ```
@@
 where $D$ is a diagonal matrix capturing the injected independence and $UW^2U^T$ is a low-rank correlation matrix derived from $\widehat \Sigma_{ref}$.
-An additional advantage of this special form is that data may be generated very efficiently, Supplemental Figure S3 and see Methods, compared to arbitrary $\Sigma_{sim}$ which typically require a Cholesky or eigenvalue decomposition.
-
-### Multivariate normal distribution
-
-We first discuss the simplest case, where our dataset is multivariate normally distributed.
-The distribution $N(\mu, \Sigma)$ is the multivariate normal distribution with mean vector $\mu$ and covariance matrix $\Sigma$.
-Here $\mu$ is a $p \times 1$ column vector and $\Sigma$ is a $p \times p$ matrix.
-In order for this to work, $\Sigma$ must be symmetric and positive semi-definite, meaning that all of its eigenvalues are non-negative. 
-This matrix is conceptually simple, since $\Sigma_{ij}$ gives the covariance of $x_i$ and $x_j$ when $x \sim N(\mu, \Sigma)$.
-More specifically, this is the population covariance matrix of $N(\mu, \Sigma)$, which does not generally equal the sample covariance matrix.
-Indeed, if $X$ is $n$ samples from $N(\mu, \Sigma)$, then $\hat \Sigma := (X - \bar X) (X - \bar X)^T / (n - 1)$ is the sample covariance matrix where $\bar X$ is the average of the $n$ samples.
-For large $n$, $\hat \Sigma$ will closely approximate $\Sigma$, but we care primarily about the situation where $n$ is small.
-In particular, $\hat \Sigma$ is at most a rank $n - 1$ matrix while $\Sigma$ could be up to rank $p$.
-This means that $N(\mu, \hat \Sigma)$ and $N(\mu, \Sigma)$ are quite different distributions: every sample from $N(\mu, \hat \Sigma)$ is contained in an $n-1$ dimensional plane.
-If $n$ is small, then this is very unlike real data, which typically is close to $p$ dimensional.
-
-The difficulty then is that we only know $\hat \Sigma$ from our reference dataset, but we need to choose a $\Sigma$ with which we can simulate data and the obvious choice of $\hat \Sigma$ is inadequate.
-The most common choice is to assume $\Sigma$ is a diagonal matrix.
-This is the situation we want to avoid where the generated data is independent: $x_i$ and $x_j$ have zero covariance unless $i = j$.
-However, this has some nice properties, such as being simple and fast to simulate (just generate univariate normal data for each variable).
-
-We describe three alternative approaches in Methods of how to choose a $\Sigma$ that will produce data similar to the input data.
-All three of these rely on a specific form of $\Sigma$, namely that
-$$ \Sigma = D^2 + UW^2U^T $$
-where $D$ is a $p \times p$ diagonal matrix, $U$ is $p \times k$ for some $k \ll p$ and $W$ is a diagonal $k \times k$ matrix.
-This is a combination of an independent part (the diagonal matrix) and a low-rank part ($U W^2 U^T$ is rank at most $k$).
-The low-rank part is also simple to generate data for.
-If $x$ is a vector of $k$ independent univariate standard normal values, then $U W x \sim N(0, P P^T)$.
-The choice of specific $D$, $U$, $W$ matrices is more in-depth and we leave the details for the Methods section.
-We have three alternative means for doing so, which we refer to as the PCA, spiked Wishart, and corpcor methods.
-
-Lastly, we emphasize that multivariate normal distributions do not capture all, or even most, types of possible dependence.
-Indeed, we see this even in the 2-dimensional case where it is well known that correlation describes only a linear relationship between two variables while in reality they may have much more complex relations.
-In higher dimensions, the problem is only worse.
-So any method based off multivariate normal distributions are making large assumptions about distribution.
-However, it is necessary to make some assumption like this.
-In the next section, though, we see that "normal" part is actually not a large obstacle.
-
-### Gaussian copula
-
-Building on the multivariate normal distribution, a popular approach to describe dependence in a high-dimensional settings is called the Gaussian copula approach.
-The idea of this approach is that by normalizing and later reversing the normalization, data that does not fit a normal distribution can still have its dependence structure described using a multivariate normal distribution.
-This allows the marginal (i.e., univariate) distributions of each genes to be specified separately from the dependence between genes.
-This operates first by normalizing each gene by fitting a distribution (such as a normal distribution, Poisson, negative binomial, or other form), and then applying the fit cumulative distribution function (CDF) to the observed values.
-Finally, those are fed to a standard normal distribution's inverse CDF to obtain values that are approximately normally distributed.
-These values are then used to compute a covariance matrix $\Sigma$ and the data is assumed to follow a multivariate normal distribution in $p$ dimensions with that covariance matrix.
-
-Here, we describe the approach using the form of covariance matrix $\Sigma = D^2 + UW^2U^T$ as above.
-Once data is obtained $Z \sim N(0, \Sigma)$, then one can undo the normalization process to obtain data with the same marginal distributions as the fit marginal distributions but with dependence determined by $\Sigma$.
-We describe this in detail:
-
-1.  Fit marginal distributions to each feature in $X$ to determine CDFs $F_{i}$ for each feature.
-2.  Transform $X$ to normalized values by $Z_{ij} = \Phi^{-1}(F_{i}(X_{ij}))$ where $\Phi$ is the CDF of the standard normal distribution.
-3.  Compute $D$, $U$, $W$ matrices from $X$ by one of three methods (see Methods).
-4.  Generate $k$ i.i.d. standard normally distributed values $u$ and $p$ i.i.d standard normally distributed values $v$.
-5.  Set $Z' = UWu + D v$.
-6.  Output the vector $X'$ where $X'_i = F_i^{-1}(\Phi(Z'))$.
-
-The generated data $Z'$ has covariance matrix $\Sigma_{sim} = D^2 + U W U^T = D^2 + U W^2 U^T$.
-Moreover, we require that $\Sigma$ satisfies that $(\Sigma_{sim})_{ii}$ is approximately 1.
-That guarantees that the output $X'$ has each entry with the same marginal distributions $F_i$ as was originally fit and inherits gene-gene dependence from $Z'$.
-This method is computationally efficient, taking hardly any more time or memory than simulations without dependence.
+An additional advantage of this special form is that data may be generated very efficiently, @fig-intro b,c, compared to arbitrary $\Sigma_{sim}$ which typically require a Cholesky or eigenvalue decomposition.
+Specifically, random samples are generated using the fact that if $u,v$ are independent vectors with iid standard normal entries, then $Dv + UWu$ has covariance matrix equal to $D^2 + UW^2U^T$.
 
@@
 
-The first of our three methods attempts to match the top $k$ PCA components of the normalized reference dataset $Z$.
+The first of our three methods attempts to match the top $k$ PCA components of $Z$, the reference data set after applying the normalizing transform.
 Specifically, let $u_1, \ldots, u_k$ be the left singular vectors of $Z$ with $\lambda_1, \ldots, \lambda_k$ the corresponding top $k$ singular values.
@@
 Therefore, the distribution of singular values is not affected by the assumption that $\Sigma_{sim}$ is diagonal.
-Moreover, for the form $\Sigma_{sim} = D^2 + U W^2 U^T$ where $p$ is very large, each column of $U$ is typically very close to orthogonal to any $e_i$, a standard basis vector.
-Therefore, when $D = c^2I$ for some constant $c$, we can approximate $\Sigma_{sim}$ as having $k$ arbitrary eigenvalues from $W^2$ and $n$ remaining eigenvalues all equal to $c^2$ corresponding to $D$.
+Moreover, for the form $\Sigma_{sim} = D^2 + U W^2 U^T$ where $p$ is very large, each column of $U$ is typically very close to orthogonal to any standard basis vector.
+Therefore, when $D = cI$ for some constant $c$, we can approximate $\Sigma_{sim}$ as having $k$ arbitrary eigenvalues from $W^2$ and $n$ remaining eigenvalues all equal to $c^2$.
 This is a spiked Wishart distribution.
@@
 Therefore, we use an efficient sampling and stochastic gradient descent method that we recently described [@wishart].
-Since the dataset has been normalized, $c$ will be close to one and $e_i \Sigma_{sim} e_i^T \approx 1$.
+Since the normalizing transform has been applied, $c$ will be close to one and $\left(\Sigma_{sim}\right)_{ii} \approx 1$ for each $i$.
 

### 2025-05-23 manuscript/supplemental/supplemental.qmd
Add mvrnorm and diagram and many revisions

--- 2025-05-05 manuscript/supplemental/supplemental.qmd
@@
 
+Here, we provide an introduction to the Gaussian copula approach.
+
+### Multivariate normal distribution
+
+We first discuss the simplest case, where our dataset is multivariate normally distributed.
+The distribution $N(\mu, \Sigma)$ is the multivariate normal distribution with mean vector $\mu$ and covariance matrix $\Sigma$.
+Here $\mu$ is a $p \times 1$ column vector and $\Sigma$ is a $p \times p$ matrix.
+In order for this to work, $\Sigma$ must be symmetric and positive semi-definite, meaning that all of its eigenvalues are non-negative. 
+This matrix is conceptually simple, since $\Sigma_{ij}$ gives the covariance of $x_i$ and $x_j$ when $x \sim N(\mu, \Sigma)$.
+More specifically, this is the population covariance matrix of $N(\mu, \Sigma)$, which does not generally equal the sample covariance matrix.
+Indeed, if $X$ is $n$ samples from $N(\mu, \Sigma)$, then $\widehat \Sigma := (X - \bar X) (X - \bar X)^T / (n - 1)$ is the sample covariance matrix where $\bar X$ is the average of the $n$ samples.
+For large $n$, $\widehat \Sigma$ will closely approximate $\Sigma$, but we care primarily about the situation where $n$ is small.
+In particular, $\widehat \Sigma$ is at most a rank $n - 1$ matrix while $\Sigma$ could be up to rank $p$.
+This means that $N(\mu, \widehat \Sigma)$ and $N(\mu, \Sigma)$ are quite different distributions: every sample from $N(\mu, \hat \Sigma)$ is contained in an $n-1$ dimensional plane.
+If $n$ is small, then this is very unlike real data, which typically is close to $p$ dimensional.
+
+The difficulty then is that we only know $\widehat \Sigma_{ref}$ from our reference dataset, but we need to choose a $\Sigma_{sim}$ with which we can simulate data and the obvious choice of $\widehat \Sigma_{ref}$ is inadequate.
+The most common choice is to assume $\Sigma_{sim}$ is a diagonal matrix.
+This is the situation we want to avoid where the generated data is independent: $x_i$ and $x_j$ have zero covariance unless $i = j$.
+However, this has some nice properties, such as being simple and fast to simulate (just generate univariate normal data for each variable).
+
+We describe three alternative approaches in Methods of how to choose a $\Sigma_{sim}$ that will produce data similar to the input data.
+All three of these rely on a specific form of $\Sigma$, namely that
+$$ \Sigma_{sim} = D^2 + UW^2U^T $$
+where $D$ is a $p \times p$ diagonal matrix, $U$ is $p \times k$ for some $k \ll p$ and $W$ is a diagonal $k \times k$ matrix.
+This is a combination of an independent part (the diagonal matrix) and a low-rank part ($U W^2 U^T$ is rank at most $k$).
+
+Generating data for $\Sigma_{sim}$ of this form is highly efficient.
+We use two basic facts about the multivariate distribution.
+First, if $A$ is a matrix and $u ~ N(0, \Sigma)$ then $Au ~ N(0, A \Sigma A^T)$.
+Second, if $u ~ N(0, \Sigma_1)$ and $v ~ N(0, \Sigma_2)$, then $u + v ~ N(0, \Sigma_1 + \Sigma_2)$.
+Therefore, given matrices $D, U, W$ and $u ~ N(0, I_k)$ and $v ~ N(0, I_p)$ (where $I_k$ and $I_p$ are the $k \times k$ and $p \times p$ identity matrices), then
+$$ D v + U W u ~ N(0, D^2 + UW^2 U^T) = N(0, \Sigma) $$
+as desired.
+
+In contrast, when given an arbitray $\Sigma$, to generate a random value in $N(0, \Sigma)$, one must compute a matrix $V$ such that $VV^T = W$.
+Then, using $v ~ N(0,I)$ , we have $Vv ~ N(0, \Sigma)$ as desired.
+However, computing $V$ is done using a Cholesky decomposition or eigenvector decomposition, both of which are computationally expensive when $\Sigma$ is large.
+
+Lastly, we emphasize that multivariate normal distributions do not capture all, or even most, types of possible dependence.
+Indeed, we see this even in the 2-dimensional case where it is well known that correlation describes only a linear relationship between two variables while in reality they may have much more complex relations.
+In higher dimensions, the problem is only worse.
+So any method based off multivariate normal distributions are making large assumptions about distribution.
+However, it is necessary to make some assumption like this.
+In the next section, though, we see that "normal" part is actually not a large obstacle.
+
+### Gaussian copula
+
+Building on the multivariate normal distribution, a popular approach to describe dependence in a high-dimensional settings is called the Gaussian copula approach.
+The idea of this approach is that by applying a normalizing transform and later reversing the transformation, data that does not fit a normal distribution can still have its dependence structure described using a multivariate normal distribution.
+This allows the marginal (i.e., univariate) distributions of each genes to be specified separately from the dependence between genes.
+This operates first by transforming each gene by fitting a distribution (such as a normal distribution, Poisson, negative binomial, or other form), and then applying the fit cumulative distribution function (CDF) to the observed values.
+Finally, those are fed to a standard normal distribution's inverse CDF to obtain values that are approximately normally distributed.
+These values are then used to compute a covariance matrix $\Sigma$ and the data is assumed to follow a multivariate normal distribution in $p$ dimensions with that covariance matrix.
+
+Here, we describe the approach using the form of covariance matrix $\Sigma = D^2 + UW^2U^T$ as above.
+Once data is obtained $Z \sim N(0, \Sigma)$, then one can undo the normalizing transformation to obtain data with the same marginal distributions as the fit marginal distributions but with dependence determined by $\Sigma$.
+We describe this in detail:
+
+1.  Fit marginal distributions to each feature in $X$ to determine CDFs $F_{i}$ for each feature.
+2.  Apply normalizing transform to $X$ by setting $Z_{ij} = \Phi^{-1}(F_{i}(X_{ij}))$ where $\Phi$ is the CDF of the standard normal distribution.
+3.  Compute $D$, $U$, $W$ matrices from $X$ by one of three methods (see Methods).
+4.  Generate $k$ i.i.d. standard normally distributed values $u$ and $p$ i.i.d standard normally distributed values $v$.
+5.  Set $z' = UWu + D v$.
+6.  Output the vector $x'$ where $x'_i = F_i^{-1}(\Phi(z'))$.
+
+The generated data vector $z'$ has covariance matrix $\Sigma_{sim} = D^2 + U W^2 U^T$.
+Moreover, we require that $\Sigma$ satisfies that $\left(\Sigma_{sim}\right)_{ii}$ is approximately 1 for each $i$.
+That guarantees that the output $x'$ has each entry with the same marginal distributions $F_i$ as was originally fit and inherits gene-gene dependence from $Z'$.
+This method is computationally efficient, taking hardly any more time or memory than simulations without dependence.
+
 ### Running SPsimSeq
 
-The R package [SPsimSeq](https://github.com/CenterForStatistics-UGent/SPsimSeq) [@Assefa2020] provides a dedicated RNA-seq and single-cell RNA-seq simulator using a Gaussian copula approach to simulate gene dependence.
+The R package SPsimSeq [@Assefa2020] provides a dedicated RNA-seq and single-cell RNA-seq simulator using a Gaussian copula approach to simulate gene dependence.
 We ran SPsimSeq using the following options:
@@
 SPsimSeq does not report either the group mean values or the log fold changes used for these genes, so we are not able to modify our own simulation to have the same set of DE genes at the same effect sizes.
+
+### Running vine copula
+
+We used the rvinecopulib package to fit and generate random data from a vine copula distribution.
+This package is a R wrapper of the high-performance vinecopulib C++ package.
+This package expects data normalized to the range of \[0, 1\].
+Therefore, we chose not to use the real RNA-seq count data and instead generated correlated random number as input which follow a multivariate normal distribution with covariance matrix constructed with a randomly generated singular vector matrix and singular values 1, 2, ..., 100.
+These values were then normalized to the interval \[0,1\] by the `pnorm` function.
+A vine copula was fit using just the "guassian" family.
+While rvinecopula supports multi-threaded operation, we ran using a single thread as with the other methods.
+
+### Running mvrnorm
+
+To assess a simple multivariate normal distribution simulation option, we ran the `mvrnorm` function from the MASS R package.
+We set means to be the mean expression of each gene and used the sample covariance matrix as the `Sigma` parameter.

### 2025-05-23 scripts/mvrnorm.benchmark.R

@@ -1,0 +1,33 @@
+library(MASS)
+library(tidyverse)
+library(rlang)
+
+method <- "mvrnorm"
+N_GENES <- snakemake@wildcards$n_genes
+N_SAMPLES <- 100
+
+# use ZT0 of raw time series data to get dependence structure
+male_read_counts <- read.csv("processed/Cortex_ZT0-counts.csv") |> 
+    column_to_rownames(var="raw_data.EnsemblID")
+
+selected_rows <- sample.int(nrow(male_read_counts), N_GENES)
+input <- male_read_counts[selected_rows,]
+
+## Generate Controls
+# With dependence
+time <- system.time({
+    Sigma <- cov(input |> t())
+    data <- mvrnorm(
+        n = N_SAMPLES,
+        mu = rowMeans(input),
+        Sigma = Sigma,
+    )
+})
+
+message("Elapsed time:")
+message(time[['elapsed']])
+tibble(
+    method = method,
+    n_genes = N_GENES,
+    time = time[['elapsed']],
+) |> write_tsv(paste0("processed/benchmark/", method, "/", N_GENES, ".txt"))

### 2025-05-23 manuscript/paper.qmd
Minor correction for corpcor methods
Square-roots were in the wrong place, essentially.
This made very little difference in practice due to the
normalization that already set things to be approximately 1.

@@
 2.
-Set $D$ to be diagonal with $D_{ii} = \sqrt{\lambda_1} (\lambda_2 \sigma_{med} + (1 - \lambda_2) \sigma_i)$ where $\sigma_i$ is the standard deviation of the $Z_{i\cdot}$ and $\sigma_{med}$ is the median of the $\sigma_i$.
-3. Set $U$ to be $\sqrt{1-\lambda} S Z / \sqrt{n-1}$ where $S$ is the diagonal matrix with $S_{ii} = \lambda_2 \sigma_{med} / \sigma_i + (1 - \lambda_2)$.
+Set $D$ to be diagonal with $D_{ii}^2 = \lambda_1 (\lambda_2 \sigma_{med} + (1 - \lambda_2) \sigma_i)$ where $\sigma_i$ is the standard deviation of the $Z_{i\cdot}$ and $\sigma_{med}$ is the median of the $\sigma_i$.
+3. Set $U$ to be $\sqrt{1-\lambda_1} S Z / \sqrt{n-1}$ where $S$ is the diagonal matrix with $S_{ii}^2 = \lambda_2 \sigma_{med} / \sigma_i + (1 - \lambda_2)$.
 4. Set $W$ to the identity.

### 2025-05-23 manuscript/paper.qmd
Move back copula stuff to Methods
Otherwise, we lack too much context to explain our own methods

@@
 
-To demonstrate the application of our simulation method for another organism, we also simulated datasets using the mouse cortex dataset GSE151923 [@Wang2022] and selected samples from male mice.
+To demonstrate the application of our simulation method for another organism, we also simulated datasets using the mouse cortex dataset GSE151923 [@Wang2022] and selected samples from male mice, as in @fig-compare-to-real.
 We then simulated differential expression experiments as above and observed a similar result (@fig-DESeq2 e-f) as for the fly whole body datasets.
@@
 
+Here, we provide an introduction to the Gaussian copula approach before describing our three methods for choosing $\Sigma_{sim}$.
+
+### Multivariate normal distribution
+
+We first discuss the simplest case, where our dataset is multivariate normally distributed.
+The distribution $N(\mu, \Sigma)$ is the multivariate normal distribution with mean vector $\mu$ and covariance matrix $\Sigma$.
+Here $\mu$ is a $p \times 1$ column vector and $\Sigma$ is a $p \times p$ matrix.
+In order for this to work, $\Sigma$ must be symmetric and positive semi-definite, meaning that all of its eigenvalues are non-negative. 
+This matrix is conceptually simple, since $\Sigma_{ij}$ gives the covariance of $x_i$ and $x_j$ when $x \sim N(\mu, \Sigma)$.
+More specifically, this is the population covariance matrix of $N(\mu, \Sigma)$, which does not generally equal the sample covariance matrix.
+Indeed, if $X$ is $n$ samples from $N(\mu, \Sigma)$, then $\widehat \Sigma := (X - \bar X) (X - \bar X)^T / (n - 1)$ is the sample covariance matrix where $\bar X$ is the average of the $n$ samples.
+For large $n$, $\widehat \Sigma$ will closely approximate $\Sigma$, but we care primarily about the situation where $n$ is small.
+In particular, $\widehat \Sigma$ is at most a rank $n - 1$ matrix while $\Sigma$ could be up to rank $p$.
+This means that $N(\mu, \widehat \Sigma)$ and $N(\mu, \Sigma)$ are quite different distributions: every sample from $N(\mu, \hat \Sigma)$ is contained in an $n-1$ dimensional plane.
+If $n$ is small, then this is very unlike real data, which typically is close to $p$ dimensional.
+
+The difficulty then is that we only know $\widehat \Sigma_{ref}$ from our reference dataset, but we need to choose a $\Sigma_{sim}$ with which we can simulate data and the obvious choice of $\widehat \Sigma_{ref}$ is inadequate.
+The most common choice is to assume $\Sigma_{sim}$ is a diagonal matrix.
+This is the situation we want to avoid where the generated data is independent: $x_i$ and $x_j$ have zero covariance unless $i = j$.
+However, this has some nice properties, such as being simple and fast to simulate (just generate univariate normal data for each variable).
+
+We describe three alternative approaches in Methods of how to choose a $\Sigma_{sim}$ that will produce data similar to the input data.
+All three of these rely on a specific form of $\Sigma$, namely that
+$$ \Sigma_{sim} = D^2 + UW^2U^T $$
+where $D$ is a $p \times p$ diagonal matrix, $U$ is $p \times k$ for some $k \ll p$ and $W$ is a diagonal $k \times k$ matrix.
+This is a combination of an independent part (the diagonal matrix) and a low-rank part ($U W^2 U^T$ is rank at most $k$).
+
+Generating data for $\Sigma_{sim}$ of this form is highly efficient.
+We use two basic facts about the multivariate distribution.
+First, if $A$ is a matrix and $u ~ N(0, \Sigma)$ then $Au ~ N(0, A \Sigma A^T)$.
+Second, if $u ~ N(0, \Sigma_1)$ and $v ~ N(0, \Sigma_2)$, then $u + v ~ N(0, \Sigma_1 + \Sigma_2)$.
+Therefore, given matrices $D, U, W$ and $u ~ N(0, I_k)$ and $v ~ N(0, I_p)$ (where $I_k$ and $I_p$ are the $k \times k$ and $p \times p$ identity matrices), then
+$$ D v + U W u ~ N(0, D^2 + UW^2 U^T) = N(0, \Sigma) $$
+as desired.
+
+In contrast, when given an arbitrary $\Sigma$, to generate a random value in $N(0, \Sigma)$, one must compute a matrix $V$ such that $VV^T = \Sigma$.
+Then, using $v ~ N(0,I)$ , we have $Vv ~ N(0, \Sigma)$ as desired.
+However, computing $V$ is done using a Cholesky decomposition or eigenvector decomposition, both of which are computationally expensive when $\Sigma$ is large.
+
+Lastly, we emphasize that multivariate normal distributions do not capture all, or even most, types of possible dependence.
+Indeed, we see this even in the 2-dimensional case where it is well known that correlation describes only a linear relationship between two variables while in reality they may have much more complex relations.
+In higher dimensions, the problem is only worse.
+So any method based off multivariate normal distributions are making large assumptions about distribution.
+However, it is necessary to make some assumption like this.
+In the next section, though, we see that "normal" part is actually not a large obstacle.
+
+### Gaussian copula
+
+Building on the multivariate normal distribution, a popular approach to describe dependence in a high-dimensional settings is called the Gaussian copula approach.
+The idea of this approach is that by applying a normalizing transform and later reversing the transformation, data that does not fit a normal distribution can still have its dependence structure described using a multivariate normal distribution.
+This allows the marginal (i.e., univariate) distributions of each genes to be specified separately from the dependence between genes.
+This operates first by transforming each gene by fitting a distribution (such as a normal distribution, Poisson, negative binomial, or other form), and then applying the fit cumulative distribution function (CDF) to the observed values.
+Finally, those are fed to a standard normal distribution's inverse CDF to obtain values that are approximately normally distributed.
+These values are then used to compute a covariance matrix $\Sigma$ and the data is assumed to follow a multivariate normal distribution in $p$ dimensions with that covariance matrix.
+
+Here, we describe the approach using the form of covariance matrix $\Sigma = D^2 + UW^2U^T$ as above.
+Once data is obtained $Z \sim N(0, \Sigma)$, then one can undo the normalizing transformation to obtain data with the same marginal distributions as the fit marginal distributions but with dependence determined by $\Sigma$.
+We describe this in detail:
+
+1.  Fit marginal distributions to each feature in $X$ to determine CDFs $F_{i}$ for each feature.
+2.  Apply normalizing transform to $X$ by setting $Z_{ij} = \Phi^{-1}(F_{i}(X_{ij}))$ where $\Phi$ is the CDF of the standard normal distribution.
+3.  Compute $D$, $U$, $W$ matrices from $X$ by one of three methods (see Methods).
+4.  Generate $k$ i.i.d. standard normally distributed values $u$ and $p$ i.i.d standard normally distributed values $v$.
+5.  Set $z' = UWu + D v$.
+6.  Output the vector $x'$ where $x'_i = F_i^{-1}(\Phi(z'))$.
+
+The generated data vector $z'$ has covariance matrix $\Sigma_{sim} = D^2 + U W^2 U^T$.
+Moreover, we require that $\Sigma$ satisfies that $\left(\Sigma_{sim}\right)_{ii}$ is approximately 1 for each $i$.
+That guarantees that the output $x'$ has each entry with the same marginal distributions $F_i$ as was originally fit and inherits gene-gene dependence from $Z'$.
+This method is computationally efficient, taking hardly any more time or memory than simulations without dependence.
+
 Below, we describe the three methods for selecting the components of the covariance matrix $\Sigma_{sim} := D^2 + U W^2 U^T$.
@@
 1.
-Compute $A_{ij} = \delta_{ij} - \sum_\ell U_{\ell,i}^2 U_{\ell,j}^2$ and $B_{i} = \lambda_i^2/(n-1)^2 - \sum_\ell U_{\ell,i}^2 V_{\ell\ell}$ where $\delta_{ij}$ is the Kronecker delta and $V = Z^T Z / (n-1)$ is the covariance matrix of $Z$.
+Compute $A_{ij} = \delta_{ij} - \sum_\ell U_{\ell,i}^2 U_{\ell,j}^2$ and $B_{i} = \lambda_i^2/(n-1)^2 - \sum_\ell U_{\ell,i}^2 \left(\widehat \Sigma_{ref}\right)_{\ell\ell}$ where $\delta_{ij}$ is the Kronecker delta and $\widehat \Sigma_{ref} = Z Z^T / (n-1)$ is the covariance matrix of $Z$.
 2. Solve $A w = B$ and set $W$ to be the diagonal matrix with $w$ along its diagonal.
 3. Set $U$ to be the $p \times k$ matrix with columns $u_i$.
-4. Set $D$ to be the diagonal matrix with $D_{ii} = \sqrt{V_{ii} - (UW^2U^T)_{ii}}$, which is the remaining variance.
+4. Set $D$ to be the diagonal matrix with $D_{ii}^2 = \left(\widehat \Sigma_{ref}\right)_{ii} - (UW^2U^T)_{ii}$, which is the remaining variance.
 
 Steps 1 and 2 give that $u_i \Sigma_{sim} u_i^T = \lambda_i$ for $i = 1, \ldots, k$.
-Step 4 ensures that $e_j \Sigma_{sim} e_j^T = 1$ for $j = 1, \ldots, p$.
+Step 4 ensures that $\left(\Sigma_{sim}\right)_{jj} = 1$ for $j = 1, \ldots, p$.
 

### 2025-05-23 manuscript/supplemental/supplemental.qmd
Move back copula stuff to Methods
Otherwise, we lack too much context to explain our own methods

--- 2025-05-21 manuscript/supplemental/supplemental.qmd
@@
 
-### Benchmark
-```{r}
-#| label: fig-benchmark
-#| fig-cap: "Comparison of (a) run time and (b) memory usage for our methods, SPsimSeq, and vine copula (implemented by the rvinecopulib R package).
-100 samples were generated with a varying number of genes.
-GSE151564 dataset was randomly subsetted to have the required number of genes to be used as the reference data set.
-Vine copula method was only run up to 2000 genes due to the memory use required at higher gene counts."
-#| fig-width: 8
-#| fig-height: 4
-library(tidyverse)
-library(patchwork)
-METHOD_ORDER <- c("real", "indep", "pca", "wishart", "corpcor", "SPsimSeq", "vinecopula")
-METHOD_COLORS <- c("black", RColorBrewer::brewer.pal(length(METHOD_ORDER)-1, "Dark2"))
-benchmark <- read_tsv("../../processed/benchmark/results.txt") |>
-    filter(!is.na(time)) # drop the ones that aren't run completely due to time/memory constraints
-
-g1 <- ggplot(benchmark, aes(x=n_genes, y=time, color=method)) +
-    geom_line(linewidth=2) +
-    labs(x = "N genes", y = "time (seconds)") + 
-    scale_color_manual(values = METHOD_COLORS, breaks = METHOD_ORDER, name="method") +
-    scale_x_log10() +
-    scale_y_log10()
-g2 <- ggplot(benchmark, aes(x=n_genes, y=max_rss, color=method)) +
-    geom_line(linewidth=2) +
-    labs(x = "N genes", y = "memory (MB)") +
-    scale_color_manual(values = METHOD_COLORS, breaks = METHOD_ORDER, name="method") +
-    scale_x_log10() +
-    scale_y_log10()
-g1 + g2 + plot_annotation(tag_levels="a") + plot_layout(guides = "collect")
-```
-
 ## Supplemental Methods
@@
 This package is a R wrapper of the high-performance vinecopulib C++ package.
-This package expects data normalized to the range of \[0, 1\].
+This package expects data scaled to the range of \[0, 1\].
 Therefore, we chose not to use the real RNA-seq count data and instead generated correlated random number as input which follow a multivariate normal distribution with covariance matrix constructed with a randomly generated singular vector matrix and singular values 1, 2, ..., 100.
-These values were then normalized to the interval \[0,1\] by the `pnorm` function.
+These values were then scaled to the interval \[0,1\] by the `pnorm` function.
 A vine copula was fit using just the "guassian" family.
 While rvinecopula supports multi-threaded operation, we ran using a single thread as with the other methods.
+
+### Running mvrnorm
+
+To assess a simple multivariate normal distribution simulation option, we ran the `mvrnorm` function from the MASS R package.
+We set means to be the mean expression of each gene and used the sample covariance matrix as the `Sigma` parameter.

### 2025-05-23 manuscript/paper.qmd
Add author summary and more discussion

@@
 
+## Author Summary
+
+Modern techniques, including high-throughput sequencing, produce more data than ever before.
+To determine the optimal computational analysis methods for these data, benchmarks are often performed using simulated data.
+This simulated data needs to closely match realistic data in order for benchmarking to meaningful.
+An often neglected aspect of this is that measurements of different values are often correlated or dependent upon each other.
+Two possible reasons for this neglect could be that there is a lack of guidelines on how to produce such data and also that methods to produce it are computationally expensive to run.
+We describe here three related methods that are both conceptually relatively simple and also highly computationally efficient.
+We demonstrated these on two applications which show how inclusion of these dependencies can affect the results of benchmarking.
+Lastly, we provide a software package to act as reference implementations of these.
+
 ## Introduction
@@
 Some methods have employed the "vine copula" approach in the context of single-cell or spatial omics simulation [@fuetterer2019constructing; @scDesign3].
-This provides an alternative approach for choosing a copula that avoids the difficulty of choosing the covariance matrix, however it comes at significant computational cost when done at the omics-scale.
+This provides an alternative approach for choosing a copula which decomposes it into a set of 2-dimensional dependencies, however it comes at significant computational cost when done at the omics-scale.
 SeqNet uses instead a network approach for gene dependence [@SeqNet].
@@
 Moreover, it can support multi-modal data such as is increasingly common in multi-omics.
+In addition, the techniques described here are relatively simple to implement and we hope they can see broader adoption outside of our package.
 
@@
 
+While we describe one special form of $\Sigma_{sim}$ where generation is highly efficient, there exist many other possiblecovariance matrix forms.
+Block diagonal covariance matrices are one such example which can efficiently represent even high-dimensional data [@Devijver02012018] and lend themselves to efficient simulation.
+Sparse matrix methods [@Bien2011-nx, @Bickel_sparse, @Cai_adaptive] yield exactly zero correlation between some variables, but it is less clear how to simulate data from these efficiently.
+Iterative sampling methods [@Krylov_subspaces] are one approach that could exploit the sparse nature and may even extend to sparse inverse covariance methods [@sparse_inverse_covariance].
+
 ## Data availability
 
-Source code for all simulations and figures in this plot is available at [github.com/itmat/dependent_sim_paper/](https://github.com/itmat/dependent_sim_paper/).
+Source code for all simulations and figures in this paper is available at [github.com/itmat/dependent_sim_paper/](https://github.com/itmat/dependent_sim_paper/).
 Source code for the `dependentsimr` package is available at [github.com/tgbrooks/dependent_sim](https://github.com/tgbrooks/dependent_sim/).

### 2025-05-27 manuscript/paper.qmd
Update intro fig

@@
 #| fig-cap: "(a) Schematic of our Gaussian copula-based approach.
+The goal is to pick $\\Sigma_{sim}$ so that $\\widehat \\Sigma_{sim} \\approx \\widehat \\Sigma_{ref}$ while using a special, highly efficient form.
 (b) Eigenvalues of the sample covariance matrix $\\widehat \\Sigma_{ref}$ of reference data compared to $\\widehat \\Sigma_{sim}$ from simulated data using $\\Sigma_{sim} = \\widehat \\Sigma_{ref}$.
@@
 Eigenvalues are numbered from largest to smallest.
-(c-d) Comparison of (c) run time and (d) memory usage for our methods, SPsimSeq, multivariate normal sampling (using the mvrnorm function from the MASS package), and vine copula (implemented by the rvinecopulib R package).
+(c-d) Comparison of (c) run time and (d) memory usage for our methods (pca, wishart, coprcor), SPsimSeq, multivariate normal sampling (using the mvrnorm function from the MASS package), and vine copula (implemented by the rvinecopulib R package).
 100 samples were generated with a varying number of genes.
@@
 #| fig-width: 8
-#| fig-height: 5
+#| fig-height: 6
 
@@
 
-(gdiagram) / (gsigma + g1 + g2 + plot_layout(guides = "collect")) + plot_annotation(tag_levels="a") + plot_layout(heights=c(1,1.5))
+(gdiagram) / ((gsigma + g1 + g2) + plot_layout(guides = "collect")) + plot_annotation(tag_levels="a") + plot_layout(heights=c(1.2,1))
 ```

### 2025-05-27 manuscript/paper.qmd
Discuss CYCLOPS performance, revise SPsimSeq discussion

@@
 This demonstrates that the correlation structure of the transcriptome can have a major impact on performance.
+In this application, the differences between corpcor and Wishart were such that, while Wishart improved upon the number of eigengenes used compared to corpcor, both undershot the level of correlation in the real data set (see Supplemental Figure 2 d) and the difference between them was too small to affect the actual performance of CYCLOPS as measured by the circular correlation.
 
@@
 Therefore, SPsimSeq induces correlation in the first multivariate normal draw, but then injects additional independence in the second uniform step.
-This two-step process makes it challenging to compare on theoretical grounds to our proposed methods since it will depended upon parameters such as the number of buckets used.
-More fine-grained buckets will induce higher correlation, and in the limit of infinite buckets, it should approach our PCA method with $k$ equal to the number of samples in the reference dataset.
-SPsimSeq is more specialized and full-featured for RNA-seq simulation, providing, for example, native differential expression (DE) options,.
+This two-step process makes it challenging to compare on theoretical grounds to our proposed methods since it will depended upon parameters such as the number of buckets used (more fine-grained buckets will produce higher correlation).
+SPsimSeq is more specialized and full-featured for RNA-seq simulation, providing, for example, native differential expression (DE) options.
 In comparison, our dependentsimr package requires manually setting marginal expression values to inject DE, but also supports other marginal distributions for situations outside of RNA-seq.

### 2025-06-02 manuscript/paper.qmd
Fix ~

@@
 We use two basic facts about the multivariate distribution.
-First, if $A$ is a matrix and $u ~ N(0, \Sigma)$ then $Au ~ N(0, A \Sigma A^T)$.
-Second, if $u ~ N(0, \Sigma_1)$ and $v ~ N(0, \Sigma_2)$, then $u + v ~ N(0, \Sigma_1 + \Sigma_2)$.
-Therefore, given matrices $D, U, W$ and $u ~ N(0, I_k)$ and $v ~ N(0, I_p)$ (where $I_k$ and $I_p$ are the $k \times k$ and $p \times p$ identity matrices), then
-$$ D v + U W u ~ N(0, D^2 + UW^2 U^T) = N(0, \Sigma) $$
+First, if $A$ is a matrix and $u \sim N(0, \Sigma)$ then $Au \sim N(0, A \Sigma A^T)$.
+Second, if $u \sim N(0, \Sigma_1)$ and $v \sim N(0, \Sigma_2)$, then $u + v \sim N(0, \Sigma_1 + \Sigma_2)$.
+Therefore, given matrices $D, U, W$ and $u \sim N(0, I_k)$ and $v \sim N(0, I_p)$ (where $I_k$ and $I_p$ are the $k \times k$ and $p \times p$ identity matrices), then
+$$ D v + U W u \sim N(0, D^2 + UW^2 U^T) = N(0, \Sigma) $$
 as desired.
@@
 In contrast, when given an arbitrary $\Sigma$, to generate a random value in $N(0, \Sigma)$, one must compute a matrix $V$ such that $VV^T = \Sigma$.
-Then, using $v ~ N(0,I)$ , we have $Vv ~ N(0, \Sigma)$ as desired.
+Then, using $v \sim N(0,I)$ , we have $Vv \sim N(0, \Sigma)$ as desired.
 However, computing $V$ is done using a Cholesky decomposition or eigenvector decomposition, both of which are computationally expensive when $\Sigma$ is large.

### 2025-06-02 manuscript/paper.qmd
Incorporate metabolomics dataset

@@
 
+We further demonstrate our methods on a different data type, metabolomics, where we use normal marginal distributions and with far fewer features measured (42), see Supplemental Figure 3.
+
 ### DESeq2 application
@@
 While runtime of simulators is typically not critical, we hypothesize that it introduces an inconvenience that has slowed the adoption of dependence in omics simulation and could be particularly cumbersome whenever large numbers of simulated datasets are needed.
-Of note is that alternative methods are typically run with only a subset of genes chosen, for example, 500-200 genes for scDesign2 [@Sun2021], 1000 genes for scDesign3 [@scDesign3], and 621 for SeqNet [@SeqNet].
-In contrast, our methods can operate on the entire space of genes, avoiding an arbitrary choice of cutoff, though it should be noted that correlation is most meaningful in high-expressed, high variance genes so subsetting genes is also a reasonable approach.
+Moreover, high memory use of the alternatives could be an even larger barrier than run-time.
+Of note is that alternative methods are typically run with only a subset of genes chosen, for example, 500-200 genes for scDesign2 [@Sun2021], 1000 genes for scDesign3 [@scDesign3], and 621 for SeqNet [@SeqNet], presumably due to the resource requirements.
+In contrast, our methods can operate on the entire space of genes, avoiding an arbitrary choice of cutoff, though it should be noted that correlation is most meaningful in high-expressed, high variance genes and so subsetting genes is a reasonable approach.
 
@@
 
-While we describe one special form of $\Sigma_{sim}$ where generation is highly efficient, there exist many other possiblecovariance matrix forms.
+While we describe one special form of $\Sigma_{sim}$ where generation is highly efficient, there exist many other possible covariance matrix forms.
 Block diagonal covariance matrices are one such example which can efficiently represent even high-dimensional data [@Devijver02012018] and lend themselves to efficient simulation.
@@
 Source code for the `dependentsimr` package is available at [github.com/tgbrooks/dependent_sim](https://github.com/tgbrooks/dependent_sim/).
-All data used is available from the Gene Expression Omnibus (GEO) with accession numbers GSE151923, GSE81142, and GSE151565.
+All sequencing data used is available from the Gene Expression Omnibus (GEO) with accession numbers GSE151923, GSE81142, and GSE151565.
+Metabolomics data was obtained from MetaboAnalyst (https://www.metaboanalyst.ca/MetaboAnalyst/upload/RocUploadView.xhtml).
 

### 2025-06-02 scripts/compare_to_real_metabolomics.R

@@ -17,43 +17,6 @@ TYPE_ORDER <- c("real", "indep", "pca", "wishart", "corpcor")
 TYPE_COLORS <- c("black", RColorBrewer::brewer.pal(length(TYPE_ORDER)-1, "Dark2"))
 
 # Load the real data ---------------------------------
-# raw <- read_tsv("../data/GSE151923_metaReadCount_ensembl.txt.gz") |>
-#   select(-GeneName, -Description, -Chromosome, -Strand)
-# read_counts <- as.matrix(raw[,2:13])
-# rownames(read_counts) <- raw[[1]]
-# 
-# actual_library_sizes <- apply(read_counts, 2, sum)
-# 
-# # Run dependentsimr on this data --------------------
-# N_SAMPLES <- 96
-# library_sizes <- (actual_library_sizes + rep(0, N_SAMPLES)) / mean(actual_library_sizes) # Recycle the existing library sizes
-# # Simulate with the PCA method
-# rs <- get_random_structure(list(counts=read_counts), method="pca", rank=2, type="DESeq2")
-# draws <- draw_from_multivariate_corr(rs, n_samples=N_SAMPLES, size_factors=library_sizes)$counts
-# 
-# # Simulate with the corpcor method
-# rs_corpcor <- get_random_structure(list(counts=read_counts), method="corpcor", type="DESeq2")
-# draws_corpcor <- draw_from_multivariate_corr(rs_corpcor, n_samples=N_SAMPLES, size_factors=library_sizes)$counts
-# 
-# # Simulate with the spiked Wishart method
-# rs_wishart <- get_random_structure(list(counts=read_counts), rank=11, method="spiked Wishart", type="DESeq2")
-# draws_wishart <- draw_from_multivariate_corr(rs_wishart, n_samples=N_SAMPLES, size_factors=library_sizes)$counts
-# 
-# # Simulate without any dependence
-# rs_indep <- remove_dependence(rs)
-# indep_draws <- draw_from_multivariate_corr(rs_indep, n_samples=N_SAMPLES, size_factors=library_sizes)$counts
-# 
-# 
-# # Scale to Counts Per Million ------------------------
-# cpm <- function(x) { # counts per million
-#   return(t(t(x) / apply(x, 2, sum) * 1e6))
-# }
-# scaled_read_data <- cpm(read_counts)
-# scaled_draws <- cpm(draws)
[…]

### 2025-06-02 manuscript/supplemental/supplemental.qmd
Incorporate metabolomics dataset

@@
 
+```{r}
+#| label: fig-compare-to-real3
+#| fig-cap: "Comparison to plasma metabolomics data set, obtained from MetaboAnalyst (https://www.metaboanalyst.ca/MetaboAnalyst/upload/RocUploadView.xhtml).
+Methods were run using normal marginal distributions.
+Unlike other examples, this included more samples (59) than measured features (42 metabolites), demonstrating that these methods also work in that case."
+#| fig-width: 10
+#| fig-height: 10
+readRDS("../../processed/compare_to_real_plot_metabolomics.RDS")
+```
+
 ## Supplemental Methods

### 2025-06-02 scripts/simulate_metabolomics.R

@@ -1,34 +1,26 @@
 library(dependentsimr)
 library(tidyverse)
 library(DESeq2)
-#library(rlang)
-library(ggplot2)
+library(rlang)
 
-#method <- snakemake@wildcards$method
-method <- "wishart"
+method <- snakemake@wildcards$method
 
-#data <- read.delim("data/ST002639_metabolomics.txt", sep='\t')
 data <- read.csv("data/plasma_nmr.csv")
 metadata <- data |> 
     filter(Label == 0) |> 
     select(Samples, Label)
-#read_counts <- data[2:nrow(data)-1,] |> 
-#    select(c(1,3:12))
-read_counts <- data |> 
+real_data <- data |> 
     filter(Label == 0) |> 
     select(-Label)
 
-ggplot(read_counts |> select(2), aes(log(X2.Hydroxybutyrate))) +
-    geom_histogram(bins=20)
-
-transposed_df <- read_counts |> 
+transposed_df <- real_data |> 
     pivot_longer(cols = -Samples, names_to = "Metabolites", values_to = "value") |> 
     pivot_wider(names_from = Samples, values_from = value) |> 
     select(-C60)
-read_counts <- transposed_df |> column_to_rownames(var="Metabolites") |> 
+real_data <- transposed_df |> column_to_rownames(var="Metabolites") |> 
     log()
-write_csv(read_counts, "data/Plasma_metabolomics_real.csv")
-read_counts <- read_counts |> 
+write_csv(real_data, "processed/Plasma_metabolomics_real.csv")
+real_data <- real_data |> 
     as.matrix()
 
 # Produce simulated data
@@ -36,19 +28,19 @@ N_SAMPLES <- 59*8
 if (method == "pca") {
     set.seed(1)
     # Simulate with the PCA method
-    rs <- get_random_structure(list(data=read_counts), method="pca", rank=2, types="normal")
+    rs <- get_random_structure(list(data=real_data), method="pca", rank=2, types="normal")
 } else if (method == "corpcor") {
     set.seed(2)
     # Simulate with the corpcor method
-    rs <- get_random_structure(list(data=read_counts), method="corpcor", types="normal")
+    rs <- get_random_structure(list(data=real_data), method="corpcor", types="normal")
 } else if (method == "wishart") {
     set.seed(3)
     # Simulate with the spiked Wishart method
-    rs <- get_random_structure(list(data=read_counts), rank=11, method="spiked Wishart", types="normal")
[…]

### 2025-06-17 manuscript/paper.qmd
Edit MS and figures
And fix metabolomics figure

@@
 format:
+  docx:
+    fig-dpi: 800
   html:
@@
 We start with a reference dataset $X$ given by an $p \times n$ data matrix of $p$ features measured in each of $n$ independent samples, which could be any real dataset of interest.
-Our goal is to generate simulated dataset that looks, at least on average, the same as this reference dataset.
+Our goal is to generate a simulated dataset that looks, at least on average, the same as this reference dataset.
 To do this, we need to capture correlations between the $p$ features, which could represent gene expressions, protein abundances, or other measured values.
@@
 We observed that DESeq2 is anti-conservative on all datasets, with similar mean true FDRs for each estimated FDR cutoff.
-However, we note that this anti-conservative nature is most present at very low FDR thresholds, was typically associated with only a single digit number of false positive genes.
-While that mean behavior was consistent whether or not the simulation included gene-gene dependence, simulations with dependence showed a substantially greater variance in the performance of DESeq2.
+While the mean behavior was consistent whether or not the simulation included gene-gene dependence, simulations with dependence showed a substantially greater variance in the performance of DESeq2.
 
@@
 This is despite the fact that differential expression is largely addressing a problem of the behavior of individual genes.
-In the second, CYCLOPS performance in estimating circadian phases depends upon gene-gene dependence, was sensitive to dependence structure of the data.
+We note that DESeq2's observed anti-conservative nature is most present at very low FDR thresholds and was typically associated with only a single digit number of false positive genes.
+Moreover, more modern methods to control FDR are compatible with DESeq2 and have been shown to be appropriately conservative [@Korthauer2019-lx].
+
+In the second application, CYCLOPS performance in estimating circadian phases depends upon gene-gene dependence, was sensitive to dependence structure of the data.
 Unlike DESeq2, CYCLOPS explicitly makes use of the correlations of genes and therefore this result is not surprising but still demonstrates the potential impact of the assumption of independence when benchmarking.
@@
 Nonetheless, these methods represent significant improvements by other metrics and we recommend the inclusion of some dependence in nearly every simulated omics dataset.
+
+We observe that our PCA method generally overshoots the level of correlation in the real data set.
+The corpcor method typically undershoots.
+The spiked Wishart method also undershoots but less so than the corpcor method and therefore we believe represents the best starting point for users.
+However, we emphasize that in any application, the users should compare their simulated data to the reference data set and this can be used to determine the best method for the specific application at hand, which could vary depending on data set.
 
@@
 Block diagonal covariance matrices are one such example which can efficiently represent even high-dimensional data [@Devijver02012018] and lend themselves to efficient simulation.
-Sparse matrix methods [@Bien2011-nx, @Bickel_sparse, @Cai_adaptive] yield exactly zero correlation between some variables, but it is less clear how to simulate data from these efficiently.
+Sparse matrix methods [@Bien2011-nx; @Bickel_sparse; @Cai_adaptive] yield exactly zero correlation between some variables, but it is less clear how to simulate data from these efficiently.
 Iterative sampling methods [@Krylov_subspaces] are one approach that could exploit the sparse nature and may even extend to sparse inverse covariance methods [@sparse_inverse_covariance].

### 2025-06-17 manuscript/supplemental/supplemental.qmd
Edit MS and figures
And fix metabolomics figure

--- 2025-05-23 manuscript/supplemental/supplemental.qmd
@@
 
+```{r}
+#| label: fig-compare-to-real3
+#| fig-cap: "Comparison to plasma metabolomics data set, obtained from MetaboAnalyst (https://www.metaboanalyst.ca/MetaboAnalyst/upload/RocUploadView.xhtml).
+Methods were run using normal marginal distributions.
+Unlike other examples, this included more samples (59) than measured features (42 metabolites), demonstrating that these methods also work in that case.
+Data were log-scaled prior to simulation."
+#| fig-width: 10
+#| fig-height: 10
+readRDS("../../processed/compare_to_real_plot_metabolomics.RDS")
+```
+
 ## Supplemental Methods

### 2025-06-17 scripts/compare_to_real_metabolomics.R

@@ -12,12 +12,12 @@ library(rlang)
 set.seed(55)
 
 # CONFIGURATION
-HIGH_EXPR_CUTOFF <- 1
 TYPE_ORDER <- c("real", "indep", "pca", "wishart", "corpcor")
 TYPE_COLORS <- c("black", RColorBrewer::brewer.pal(length(TYPE_ORDER)-1, "Dark2"))
 
 # Load the real data ---------------------------------
-scaled_read_data <- read_csv("data/Plasma_metabolomics_real.csv")
+scaled_read_data <- read_csv("processed/Plasma_metabolomics_real.csv") |>
+    as.matrix()
 scaled_draws <- read_delim("simulated_data/Plasma_metabolomics.pca.txt", delim = '\t') |> 
     select(-Metabolites) |> 
     as.matrix()
@@ -33,8 +33,8 @@ scaled_indep_draws <- read_delim("simulated_data/Plasma_metabolomics.indep.txt",
 
 ## Generate the figures
 
-simulated_mean <- apply(scaled_draws[,1:ncol(read_counts)], 1, mean)
-simulated_variance <- apply(scaled_draws[,1:ncol(read_counts)], 1, var)
+simulated_mean <- apply(scaled_draws[,1:ncol(scaled_read_data)], 1, mean)
+simulated_variance <- apply(scaled_draws[,1:ncol(scaled_read_data)], 1, var)
 real_mean <- apply(scaled_read_data, 1, mean)
 real_variance <- apply(scaled_read_data, 1, var)
 marg_dist <- data.frame(
@@ -53,7 +53,7 @@ g2<-ggplot(marg_dist|> filter(real_mean > 0.1))+
   scale_color_viridis_c(option="inferno")
 
 # Plot the top principal components --------------------
-constant_rows <- apply(read_counts, 1, function(x) all(x - mean(x) == 0))
+constant_rows <- apply(scaled_read_data, 1, function(x) all(x - mean(x) == 0))
 pca <- prcomp(t(scaled_read_data[!constant_rows,]), rank.=2, scale.=TRUE)
 projected_read_data <- predict(pca, t(scaled_read_data[!constant_rows, ]))
 projected_draws <- predict(pca, t(scaled_draws[!constant_rows, ]))
@@ -78,7 +78,7 @@ g3<-ggplot(data = both_data, aes(x=PC1, y=PC2,color=type)) +
   scale_color_manual(values = TYPE_COLORS, breaks = TYPE_ORDER, name="type")
 
[…]

### 2025-06-17 manuscript/paper.qmd
Edits based off coauthor comments

@@
 To alleviate this, we describe three approaches for efficiently generating omics-scale data with correlated measures which mimic real datasets.
-These approaches all are based on a Gaussian copula approach with a covariance matrix that decomposes into a diagonal part and a low-rank part.
+These approaches are all based on a Gaussian copula approach with a covariance matrix that decomposes into a diagonal part and a low-rank part.
 We use these approaches to demonstrate the importance of including correlation in two benchmarking applications.
@@
 Second, we demonstrate that CYCLOPS, a method for inferring circadian time of collection from transcriptomics, improves in performance when given gene-gene dependencies in some circumstances.
-We provide an R package, dependentsimr, that has efficient implementations of these methods and can generate dependent data with arbitrary distributions, including discrete (binary, ordered categorical, Poisson, negative binomial), continuous (normal), or with an empirical distribution.
+We provide an R package, dependentsimr, that has efficient implementations of these methods and can generate dependent data with arbitrary marginal distributions, including discrete (binary, ordered categorical, Poisson, negative binomial), continuous (normal), or with an empirical distribution.
 
@@
 This creates dual challenges in generating realistic simulated data for the purposes of benchmarking.
-First, there is not enough data to be able to compute a meaningful dependence structure (e.g., a correlation matrix).
+First, there is not enough data to be able to estimate an accurate dependence structure (e.g., a correlation matrix).
 Second, generating omics-scale data with a specified correlation matrix is slow due to the typical $O(p^3)$ nature of these algorithms, where $p$ is the number of measurements per sample.
 Moreover, there is a lack of practical guidance on how to generate simulated data with realistic dependence.
-These often mean that simulators assume independence of the measurements, which does not reflect reality.
+Simulators therefore often assume independence of the measurements, which does not reflect reality.
 
 Here, we describe three related solutions which all offer good performance and ease-of-use even for large omics-scale problems.
-This expands off a discussion we originally wrote as part of a larger discussion on best practices in omics benchmarking [@Brooks2024].
+This expands off part of a larger discussion we published on best practices in omics benchmarking [@Brooks2024].
 Our goal here is to show that generating correlated data does not have to be onerous and instead should be considered a baseline requirement when simulating data.
@@
 Some methods have employed the "vine copula" approach in the context of single-cell or spatial omics simulation [@fuetterer2019constructing; @scDesign3].
-This provides an alternative approach for choosing a copula which decomposes it into a set of 2-dimensional dependencies, however it comes at significant computational cost when done at the omics-scale.
+Vine copulas provide an alternative approach for choosing a copula which decomposes it into a set of 2-dimensional dependencies, however it comes at significant computational cost when done at the omics-scale.
 SeqNet uses instead a network approach for gene dependence [@SeqNet].
@@
 Second, the 'spiked Wishart' method fits $k$ components such that simulations with the same number of samples as the reference dataset will have, on average, the same PCA component variances as the reference.
-Unlike 'PCA', these variances are computed with respect to the simulated data's own PCA and not using the PCA weights of the reference dataset.
-Third, the 'corpcor' method uses the popular R library `corpcor` [@SchaferStrimmer2005; @OpgenRheinStrimmer2007], which implements a James-Stein type shrinkage estimator for the covariance matrix as a linear interpolation of the sample covariance matrix and a diagonal matrix.
-No method exactly captures the input data, indicating room for future research, but all improve upon the common approach of assuming independence.
-For ease of use, we implement this in an R package which supports normal, Poisson, DESeq2-based (negative binomial with sample-specific size factors), and empirical (for ordinal data) marginal distributions.
+Unlike 'PCA', these variances are computed with respect to the simulated data's own PCA rather than using the PCA weights of the reference dataset.
+Third, the 'corpcor' method builds off the popular R library `corpcor` [@SchaferStrimmer2005; @OpgenRheinStrimmer2007], which implements a James-Stein type shrinkage estimator for the covariance matrix as a linear interpolation of the sample covariance matrix and a diagonal matrix.
+No method exactly replicates the reference dataset's properties, indicating room for future research, but all improve upon the common approach of assuming independence.
+We implement this in an R package which supports normal, Poisson, DESeq2-based (negative binomial with sample-specific size factors), and empirical (for ordinal data) marginal distributions.
 
-We show two applications which demonstrate the effects of including dependence of measurements in simulated data when benchmarking computational pipelines.
+We include two applications which demonstrate the effects of including dependence of measurements in simulated data when benchmarking computational pipelines.
 First, we simulate RNA-seq data with differential expression between two conditions.
@@
 Second, we simulated a time series of RNA-seq data points and used the CYCLOPS method [@Anafi2017] to infer collection time from the RNA-seq data, without time labels.
-Depending upon settings used, performance of CYCLOPS dependent substantially on the dependence structure of the data, and surprisingly showed worst performance when given data with independent genes.
+Depending upon settings used, the performance of CYCLOPS depends substantially on the dependence structure of the data, and surprisingly showed improved performance when given data with dependence.
 
@@
 
-We start with a reference dataset $X$ given by an $p \times n$ data matrix of $p$ features measured in each of $n$ independent samples, which could be any real dataset of interest.
-Our goal is to generate a simulated dataset that looks, at least on average, the same as this reference dataset.
+We start with a reference dataset $X$ given by a $p \times n$ data matrix of $p$ features measured in each of $n$ independent samples, which could be any real dataset of interest.
+Our goal is to be able to generate simulated datasets that look the same as this reference dataset, on average and when $n$ samples were simulated.
 To do this, we need to capture correlations between the $p$ features, which could represent gene expressions, protein abundances, or other measured values.
@@
 This multivariate normal distribution is parameterized by a $p \times p$ covariance matrix, $\Sigma$.
-In omics, choosing the covariance matrix $\Sigma_{sim}$ to use during simulation presents a challenge since $n < p$.
-The most obvious choice is to use the sample covariance matrix $\widehat \Sigma_{ref}$ of the reference dataset (possibly after some transformation of the data).
-However, using $\widehat \Sigma_{ref}$ is a poor choice for $\Sigma_{sim}$.
-Indeed, we consider a choice of $\Sigma_{sim}$ to be successful if datasets of $n$ samples simulated using $\Sigma_{sim}$ have sample covaraince matrices $\widehat \Sigma_{sim}$ that are similar to $\widehat \Sigma_{ref}$.
-Counter-intuitively, data simulated using $\Sigma_{sim} := \widehat \Sigma_{ref}$ have sample covariance matrices that systematically and substantially differ from $\widehat \Sigma_{ref}$, see @fig-intro b.
+In omics, choosing the covariance matrix $\Sigma_{\text{sim}}$ to use during simulation presents a challenge since $n < p$.
+The most obvious choice is to use the sample covariance matrix $\widehat \Sigma_{\text{ref}}$ of the reference dataset (possibly after some transformation of the data).
+However, using $\widehat \Sigma_{\text{ref}}$ is a poor choice for $\Sigma_{\text{sim}}$.
+Indeed, we consider a choice of $\Sigma_{\text{sim}}$ to be successful if datasets of $n$ samples simulated using $\Sigma_{\text{sim}}$ have sample covaraince matrices $\widehat \Sigma_{\text{sim}}$ that are similar to $\widehat \Sigma_{\text{ref}}$.
+Counter-intuitively, data simulated using $\Sigma_{\text{sim}} := \widehat \Sigma_{\text{ref}}$ have sample covariance matrices that systematically and substantially differ from $\widehat \Sigma_{\text{ref}}$, see @fig-intro b.
 
@@
 #| fig-cap: "(a) Schematic of our Gaussian copula-based approach.
-The goal is to pick $\\Sigma_{sim}$ so that $\\widehat \\Sigma_{sim} \\approx \\widehat \\Sigma_{ref}$ while using a special, highly efficient form.
-(b) Eigenvalues of the sample covariance matrix $\\widehat \\Sigma_{ref}$ of reference data compared to $\\widehat \\Sigma_{sim}$ from simulated data using $\\Sigma_{sim} = \\widehat \\Sigma_{ref}$.
+The goal is to pick $\\Sigma_{\text{sim}}$ so that $\\widehat \\Sigma_{\text{sim}} \\approx \\widehat \\Sigma_{\text{ref}}$ while using a special, highly efficient form.
+(b) Eigenvalues of the sample covariance matrix $\\widehat \\Sigma_{\text{ref}}$ of reference data compared to $\\widehat \\Sigma_{\text{sim}}$ from simulated data using $\\Sigma_{\text{sim}} = \\widehat \\Sigma_{\text{ref}}$.
 100 genes and 12 samples were used for both reference and the 20 simulated datasets.
@@
 
-One reason this occurs is that $\widehat \Sigma_{ref}$ is low rank, specifically at most rank $n - 1$, which is less than $p$ in typical omics studies.
-This means that any data generated using $\Sigma_{sim} = \widehat \Sigma_{ref}$ will lie on a $n-1$-dimensional hyperplane.
-Intuitively, this means that the entire probability distribution of possible $p$ dimensional data have been squished into a $n-1$ dimensional space, generating much higher gene-gene correlations than in the reference dataset.
-In contrast, we know that omics contains at least some level of random noise that is approximately independent from one gene to the next.
+One reason this occurs is that $\widehat \Sigma_{\text{ref}}$ is low rank, specifically at most rank $n - 1$, which is much less than $p$ in typical omics studies.
+This means that any data generated using $\Sigma_{\text{sim}} = \widehat \Sigma_{\text{ref}}$ will lie on an $n-1$-dimensional hyperplane.
+Intuitively, this means that the entire probability distribution of potentially $p$ dimensional data has been squished into a $n-1$ dimensional space, generating much higher gene-gene correlations than in the reference dataset.
+In contrast, we know that omics contains at least some level of random noise that is essentially independent from one gene to the next.
 For example, RNA-seq involves a random sampling of fragments from the true sample of RNA molecules and other omics modes involve various background noise effects.
-We therefore need to "inject" some level of additional randomness that is independent between genes on top of the correlations captured by $\widehat \Sigma_{ref}$
+We therefore need to "inject" some level of additional randomness that is independent between genes on top of the correlations captured by $\widehat \Sigma_{\text{ref}}$
 
-We present three methods - called here PCA, spiked Wishart, and corpcor - for making this choice of $\Sigma_{sim}$, all which take the form
-$$ \Sigma_{sim} = D^2 + UW^2U^T $$
-where $D$ is a diagonal matrix capturing the injected independence and $UW^2U^T$ is a low-rank correlation matrix derived from $\widehat \Sigma_{ref}$.
-An additional advantage of this special form is that data may be generated very efficiently, @fig-intro b,c, compared to arbitrary $\Sigma_{sim}$ which typically require a Cholesky or eigenvalue decomposition.
-Specifically, random samples are generated using the fact that if $u,v$ are independent vectors with iid standard normal entries, then $Dv + UWu$ has covariance matrix equal to $D^2 + UW^2U^T$.
+We present three methods - to be called PCA, spiked Wishart, and corpcor - for making the choice of $\Sigma_{\text{sim}}$
+All three take the form
+$$ \Sigma_{\text{sim}} = D^2 + UW^2U^T $$
+where $D$ is a diagonal matrix capturing the injected independence and $UW^2U^T$ is a low-rank correlation matrix derived from $\widehat \Sigma_{\text{ref}}$.
+An additional advantage of this special form is that data may be generated very efficiently, @fig-intro b,c, compared to arbitrary $\Sigma_{\text{sim}}$ which typically require a Cholesky or eigenvalue decomposition.
+Specifically, random samples are generated using the fact that if $u,v$ are independent vectors with iid standard normal entries, then $Dv + UWu$ has covariance matrix equal to $D^2 + UW^2U^T$, see Methods.
 
@@
 
-We also ran for comparison SPsimSeq, another RNA-seq simulator for both bulk and single-cell that uses a Gaussian copula-based model of dependence, see Supplemental Methods for details.
+For comparison we also ran SPsimSeq, another RNA-seq simulator for both bulk and single-cell that uses a Gaussian copula-based model of dependence, see Supplemental Methods for details.
 
@@
 #| label: fig-compare-to-real
-#| fig-cap: "Comparison to real data run on a mouse cortex dataset from GSE151923.
+#| fig-cap: "Comparison to real reference data run on a mouse cortex dataset from GSE151923.
 (a-b) Comparison of gene (a) mean expression and (b) variance, log-scaled in real and PCA simulated data.
@@
 
-We further demonstrate our methods on a different data type, metabolomics, where we use normal marginal distributions and with far fewer features measured (42), see Supplemental Figure 3.
+We further demonstrated our methods on metabolomics data, where we use normal marginal distributions and with just 42 features measured, see Supplemental Figure 3.
 
@@
 
-We used a fly whole body RNA-Seq dataset GSE81142, see Supplemental Figure 1, and selected samples of male flies without treatment and after at least 2 hours of feeding to simulate 5 "control" samples.
+We used a fly whole body RNA-Seq dataset GSE81142, see Supplemental Figure 1.
+To select a "control" dataset samples were selected if they met the following criteria: male flies; without treatment; and after at least 2 hours of feeding.
 Unnormalized read counts were used as the reference dataset and all simulations used the DESeq2-based negative binomial model.
 We then randomly selected 5% of the genes to be differentially expressed, with absolute $\log_2$ fold change uniformly distributed between $0.2$ and $2.0$, either up or down regulated chosen randomly, and simulated 5 "experimental" samples.
-This was repeated 20 times for each of four dependence conditions (independent, PCA, Wishart, and corpcor).
+This was repeated 20 times for each of four dependence methods (independent, PCA, Wishart, and corpcor).
 
@@
 To demonstrate the application of our simulation method for another organism, we also simulated datasets using the mouse cortex dataset GSE151923 [@Wang2022] and selected samples from male mice, as in @fig-compare-to-real.
-We then simulated differential expression experiments as above and observed a similar result (@fig-DESeq2 e-f) as for the fly whole body datasets.
+We then simulated differential expression experiments as above and observed a similar result (@fig-DESeq2 e,f) as for the fly whole body datasets.
 
@@
 As before, unnormalized read counts were used as the reference dataset and all simulations used the DESeq2-based negative binomial model.
-We computed the dependence structure of the genes as well as the variances of marginal distributions using the 11 time point 0 samples after removing one outlier and computed the means of gene expressions at each time point.
-We then used these to simulate 20 time series datasets for each of the independent, PCA, Wishart, and corpcor simulation methods.
+We computed the dependence structure of the genes as well as the variances of marginal distributions using the 12 time point 0 samples.
+We computed the means of gene expressions at each time point.
+We then used these to simulate 20 time series datasets for each of the four simulation methods: independent, PCA, Wishart, and corpcor.
+
+We note that, unlike many timeseries methods that consider correlation across time points, the dependence here is between genes within the same sample and no dependence between samples or timepoints is generated.
+This reflects scenarios where CYCLOPS is used, for example, post-mortem biopsy samples of unrelated individuals collected at different times [@CYCLOPS2].
 
@@
 By default, CYCLOPS performs dimension reduction so that each dimension (called an "eigengene") contains at least 3% of the total variance.
-We found that CYCLOPS performance depended significantly on this parameter, with the default producing good performance across all simulations.
-However, when dropping CYCLOPS to require just 2% variance in each eigengene, we found that its performance depends significantly on the dependence structure of the simulated time series data (@fig-CYCLOPS a).
-At that setting, CYCLOPS performance is much higher in the PCA method and moderately improved in Wishart method, compared to the independent method. 
+We found that CYCLOPS accuracy depended significantly on this parameter, with the default producing good accuracy across all simulations.
+However, when dropping CYCLOPS to require just 1% variance in each eigengene, we found that its accuracy depends significantly on the dependence structure of the simulated time series data (@fig-CYCLOPS a).
+At that setting, CYCLOPS accuracy is much higher in the PCA method and moderately improved in Wishart method, compared to the 'independent' method. 
 This difference is likely driven by the difference in the number of eigengenes used (@fig-CYCLOPS b), which is a measure of how much dependence is present in the dataset.
-This demonstrates that the correlation structure of the transcriptome can have a major impact on performance.
-In this application, the differences between corpcor and Wishart were such that, while Wishart improved upon the number of eigengenes used compared to corpcor, both undershot the level of correlation in the real data set (see Supplemental Figure 2 d) and the difference between them was too small to affect the actual performance of CYCLOPS as measured by the circular correlation.
+This demonstrates that the correlation structure of the transcriptome can have a major impact on accuracy.
+In this application, the differences between corpcor and Wishart were such that, while Wishart improved upon the number of eigengenes used compared to corpcor, both undershot the level of correlation in the real data set (see Supplemental Figure 2 d) and the difference between them was too small to affect the actual accuracy of CYCLOPS as measured by the circular correlation.
 
@@
 #| label: fig-CYCLOPS
-#| fig-cap: "Performance of CYCLOPS on simulated time series datasets based on mouse cortex dataset (GSE151565), using eigengenes of at least 2% variance.
+#| fig-cap: "Performance of CYCLOPS on simulated time series datasets based on mouse cortex dataset (GSE151565), using eigengenes of at least 1% variance.
 (a) Absolute circular correlations between true phases and CYCLOPS estimated phases on the simulated datasets.
@@
 
-Here, we provide an introduction to the Gaussian copula approach before describing our three methods for choosing $\Sigma_{sim}$.
+Here, we provide an introduction to the Gaussian copula approach before describing our three methods for choosing $\Sigma_{\text{sim}}$.
 
@@
 
-The difficulty then is that we only know $\widehat \Sigma_{ref}$ from our reference dataset, but we need to choose a $\Sigma_{sim}$ with which we can simulate data and the obvious choice of $\widehat \Sigma_{ref}$ is inadequate.
-The most common choice is to assume $\Sigma_{sim}$ is a diagonal matrix.
+The difficulty then is that we only know $\widehat \Sigma_{\text{ref}}$ from our reference dataset, but we need to choose a $\Sigma_{\text{sim}}$ with which we can simulate data and the obvious choice of $\widehat \Sigma_{\text{ref}}$ is inadequate.
+The most common choice is to assume $\Sigma_{\text{sim}}$ is a diagonal matrix.
 This is the situation we want to avoid where the generated data is independent: $x_i$ and $x_j$ have zero covariance unless $i = j$.
@@
 
-We describe three alternative approaches in Methods of how to choose a $\Sigma_{sim}$ that will produce data similar to the input data.
+We describe three alternative approaches in Methods of how to choose a $\Sigma_{\text{sim}}$ that will produce data similar to the input data.
 All three of these rely on a specific form of $\Sigma$, namely that
-$$ \Sigma_{sim} = D^2 + UW^2U^T $$
+$$ \Sigma_{\text{sim}} = D^2 + UW^2U^T $$
 where $D$ is a $p \times p$ diagonal matrix, $U$ is $p \times k$ for some $k \ll p$ and $W$ is a diagonal $k \times k$ matrix.
@@
 
-Generating data for $\Sigma_{sim}$ of this form is highly efficient.
+Generating data for $\Sigma_{\text{sim}}$ of this form is highly efficient.
 We use two basic facts about the multivariate distribution.
@@
 
-The generated data vector $z'$ has covariance matrix $\Sigma_{sim} = D^2 + U W^2 U^T$.
-Moreover, we require that $\Sigma$ satisfies that $\left(\Sigma_{sim}\right)_{ii}$ is approximately 1 for each $i$.
+The generated data vector $z'$ has covariance matrix $\Sigma_{\text{sim}} = D^2 + U W^2 U^T$.
+Moreover, we require that $\Sigma$ satisfies that $\left(\Sigma_{\text{sim}}\right)_{ii}$ is approximately 1 for each $i$.
 That guarantees that the output $x'$ has each entry with the same marginal distributions $F_i$ as was originally fit and inherits gene-gene dependence from $Z'$.
@@
 
-Below, we describe the three methods for selecting the components of the covariance matrix $\Sigma_{sim} := D^2 + U W^2 U^T$.
+Below, we describe the three methods for selecting the components of the covariance matrix $\Sigma_{\text{sim}} := D^2 + U W^2 U^T$.
 
@@
 The first of our three methods attempts to match the top $k$ PCA components of $Z$, the reference data set after applying the normalizing transform.
-Specifically, let $u_1, \ldots, u_k$ be the left singular vectors of $Z$ with $\lambda_1, \ldots, \lambda_k$ the corresponding top $k$ singular values.
-This method computes $\Sigma_{sim}$ such that $u_i^T \Sigma_{sim} u_i = \lambda_i^2$, i.e.
-that the variance in the direction of $u_i$ exactly matches of the reference dataset's variance in that same direction.
+Specifically, let $u_1, \ldots, u_k$ be the left singular vectors of $Z$ with $\lambda_1, \ldots, \lambda_k$ the corresponding top $k$ singular values and let $U$ be the $p \times k$ matrix with columns $u_i$.
+This method computes $\Sigma_{\text{sim}}$ such that $u_i^T \Sigma_{\text{sim}} u_i = \lambda_i^2$, i.e.
+that the variance in the direction of $u_i$ exactly matches the reference dataset's variance in that same direction.
 One solution is to use the sample covariance matrix, but that is not full rank and would match for all $i \leq n$ instead of just $i \leq k$.
@@
 1.
-Compute $A_{ij} = \delta_{ij} - \sum_\ell U_{\ell,i}^2 U_{\ell,j}^2$ and $B_{i} = \lambda_i^2/(n-1)^2 - \sum_\ell U_{\ell,i}^2 \left(\widehat \Sigma_{ref}\right)_{\ell\ell}$ where $\delta_{ij}$ is the Kronecker delta and $\widehat \Sigma_{ref} = Z Z^T / (n-1)$ is the covariance matrix of $Z$.
+Compute $A_{ij} = \delta_{ij} - \sum_\ell U_{\ell,i}^2 U_{\ell,j}^2$ and $B_{i} = \lambda_i^2/(n-1)^2 - \sum_\ell U_{\ell,i}^2 \left(\widehat \Sigma_{\text{ref}}\right)_{\ell\ell}$ where $\delta_{ij}$ is the Kronecker delta and $\widehat \Sigma_{\text{ref}} = Z Z^T / (n-1)$ is the covariance matrix of $Z$.
 2. Solve $A w = B$ and set $W$ to be the diagonal matrix with $w$ along its diagonal.
 3. Set $U$ to be the $p \times k$ matrix with columns $u_i$.
-4. Set $D$ to be the diagonal matrix with $D_{ii}^2 = \left(\widehat \Sigma_{ref}\right)_{ii} - (UW^2U^T)_{ii}$, which is the remaining variance.
+4. Set $D$ to be the diagonal matrix with $D_{ii}^2 = \left(\widehat \Sigma_{\text{ref}}\right)_{ii} - (UW^2U^T)_{ii}$, which is the remaining variance.
 
-Steps 1 and 2 give that $u_i \Sigma_{sim} u_i^T = \lambda_i$ for $i = 1, \ldots, k$.
-Step 4 ensures that $\left(\Sigma_{sim}\right)_{jj} = 1$ for $j = 1, \ldots, p$.
+Steps 1 and 2 give that $u_i \Sigma_{\text{sim}} u_i^T = \lambda_i$ for $i = 1, \ldots, k$.
+Step 4 ensures that $\left(\Sigma_{\text{sim}}\right)_{jj} = 1$ for $j = 1, \ldots, p$.
 
@@
 The second method also makes use of PCA but has a different objective.
-If $n$ samples are drawn from $N(0, \Sigma_{sim})$ then we want the variances of their PCA components to match those of the reference dataset.
-Specifically, let $\lambda_1, \ldots, \lambda_{n-1}$ be the ${n-1}$ non-zero singular values of $Z$ ($\lambda_n$ is always approximately zero due to the normalization procedure), and let $\lambda'_1, \ldots, \lambda'_{n-1}$ be the singular values of $Z'$ where $Z'$ has $n-1$ columns each iid $N(0, \Sigma_{sim})$.
-Then we want to choose $\Sigma_{sim}$ such that $E[\lambda'_i] = \lambda_i$ for each $i$, where $E[Y]$ denotes the expectation of the random variable $Y$.
+If $n$ samples are drawn from $N(0, \Sigma_{\text{sim}})$ then we want the variances of their PCA components to match those of the reference dataset.
+Specifically, let $\lambda_1, \ldots, \lambda_{n-1}$ be the ${n-1}$ non-zero singular values of $Z$ ($\lambda_n$ is always approximately zero due to the normalization procedure), and let $\lambda'_1, \ldots, \lambda'_{n-1}$ be the singular values of $Z'$ where $Z'$ has $n-1$ columns each iid $N(0, \Sigma_{\text{sim}})$.
+Then we want to choose $\Sigma_{\text{sim}}$ such that $E[\lambda'_i] = \lambda_i$ for each $i$, where $E[Y]$ denotes the expectation of the random variable $Y$.
 
 Since the distribution of the $\lambda'_i$ does not have a known analytic solution, we approximate this situation with the spiked Wishart distribution.
-The rank $k-1$ Wishart distribution is the distribution of sample covariance matrices of $Z$ whose $n$ columns of $Z$ are iid $N(0, \Sigma_{sim})$.
-The spiked Wishart is the special case where $\Sigma_{sim}$ has $k$ arbitrary eigenvalues and the remaining are all equal to a constant.
-Note that the singular values of $Z$ are the square roots of the eigenvalues of $\widehat \Sigma_{ref}.
-While our case has $\Sigma_{sim}$ non-diagonal, $\Sigma_{sim}$ may be diagonalized by orthogonal rotations due the spectral theorem, and orthogonal rotations do not change the singular values of $Z$.
-Therefore, the distribution of singular values is not affected by the assumption that $\Sigma_{sim}$ is diagonal.
-Moreover, for the form $\Sigma_{sim} = D^2 + U W^2 U^T$ where $p$ is very large, each column of $U$ is typically very close to orthogonal to any standard basis vector.
-Therefore, when $D = cI$ for some constant $c$, we can approximate $\Sigma_{sim}$ as having $k$ arbitrary eigenvalues from $W^2$ and $n$ remaining eigenvalues all equal to $c^2$.
+The rank $k-1$ Wishart distribution is the distribution of sample covariance matrices of $Z$ whose $n$ columns of $Z$ are iid $N(0, \Sigma_{\text{sim}})$.
+The spiked Wishart is the special case where $\Sigma_{\text{sim}}$ has $k$ arbitrary eigenvalues and the remaining are all equal to a constant.
+Note that the singular values of $Z$ are the square roots of the eigenvalues of $\widehat \Sigma_{\text{ref}}$.
+While our case has $\Sigma_{\text{sim}}$ non-diagonal, $\Sigma_{\text{sim}}$ may be diagonalized by orthogonal rotations due the spectral theorem, and orthogonal rotations do not change the singular values of $Z$.
+Therefore, the distribution of singular values is not affected by the assumption that $\Sigma_{\text{sim}}$ is diagonal.
+Moreover, for the form $\Sigma_{\text{sim}} = D^2 + U W^2 U^T$ where $p$ is very large, each column of $U$ is typically very close to orthogonal to any standard basis vector.
+Therefore, when $D = cI$ for some constant $c$, we can approximate $\Sigma_{\text{sim}}$ as having $k$ arbitrary eigenvalues from $W^2$ and $n$ remaining eigenvalues all equal to $c^2$.
 This is a spiked Wishart distribution.
@@
 Therefore, we use an efficient sampling and stochastic gradient descent method that we recently described [@wishart].
-Since the normalizing transform has been applied, $c$ will be close to one and $\left(\Sigma_{sim}\right)_{ii} \approx 1$ for each $i$.
+Since the normalizing transform has been applied, $c$ will be close to one and $\left(\Sigma_{\text{sim}}\right)_{ii} \approx 1$ for each $i$.
 
@@
 2.
-Compute $w_1, \ldots, w_k$ and $c$ by stochastic gradient descent minimizing $\sum_{i} (E[\lambda'_i] - \lambda_i)^2$ for $\Sigma_{sim}$ diagonal with entries $w_1^2, \ldots, w_k^2, c^2, \ldots, c^2$ [@wishart].
+Compute $w_1, \ldots, w_k$ and $c$ by stochastic gradient descent minimizing $\sum_{i} (E[\lambda'_i] - \lambda_i)^2$ for $\Sigma_{\text{sim}}$ diagonal with entries $w_1^2, \ldots, w_k^2, c^2, \ldots, c^2$ [@wishart].
 4. Set $W$ diagonal with the entries $w_1, \ldots, w_k$.
@@
 It then uses $\lambda_1$ to linearly interpolate the sample covariance matrix towards the identity matrix $I$ and $\lambda_2$ to interpolate the vector of variances towards the median variance value.
-Since the sample covariance matrix is rank at most $n$, we again obtain a matrix of the form $\Sigma_{sim} = D^2 + UW^2U^T$.
+Since the sample covariance matrix is rank at most $n$, we again obtain a matrix of the form $\Sigma_{\text{sim}} = D^2 + UW^2U^T$.
 
@@
 
-We described the well-known Gaussian copula approach and recommended a specific form of covariance matrix which is well tailored to omics data simulation.
+Here we have described the well-known Gaussian copula approach and argued that a specific form of covariance matrix which is well tailored to omics data simulation.
 We developed three methods using this form of covariance matrix which can be used to mimic a reference dataset for simulation.
 All of these methods use a multivariate normal distribution as an intermediate step and therefore substantially restrict the kinds of dependence that can be simulated.
-However, when operating in a high-dimensional space some simplification is likely required.
+However, when operating in a high-dimensional space some simplification is required.
 
@@
 This implementation is efficient and simple, requiring just two lines of code to fit a model to a reference dataset and then simulate data from it.
-We demonstrated this package on RNA-seq data, using the DESeq2 method to fit negative binomial marginal distributions.
-However, this package is actually quite general and supports normal, Poisson, negative binomial, and arbitrary ordered discrete distributions using the empirical CDF.
+We demonstrated this package on RNA-seq data, using the DESeq2 method to fit negative binomial marginal distributions as well as a metabolomics dataset using normal marginal distributions.
+Our package is even more general and supports normal, Poisson, negative binomial, and arbitrary ordered discrete distributions using the empirical CDF.
 Moreover, it can support multi-modal data such as is increasingly common in multi-omics.
@@
 
-We highlight that our methods are highly efficient and easily scale to the large feature counts typical of omics data sets.
-While runtime of simulators is typically not critical, we hypothesize that it introduces an inconvenience that has slowed the adoption of dependence in omics simulation and could be particularly cumbersome whenever large numbers of simulated datasets are needed.
+We demonstrated that our methods are highly efficient and easily scale to the large feature counts typical of omics data sets.
+While runtime of simulators is typically not critical, we posit that it introduces a sufficient inconvenience to have slowed the adoption of dependence in omics simulation and could be particularly cumbersome whenever large numbers of simulated datasets are needed.
 Moreover, high memory use of the alternatives could be an even larger barrier than run-time.
-Of note is that alternative methods are typically run with only a subset of genes chosen, for example, 500-200 genes for scDesign2 [@Sun2021], 1000 genes for scDesign3 [@scDesign3], and 621 for SeqNet [@SeqNet], presumably due to the resource requirements.
-In contrast, our methods can operate on the entire space of genes, avoiding an arbitrary choice of cutoff, though it should be noted that correlation is most meaningful in high-expressed, high variance genes and so subsetting genes is a reasonable approach.
+Of note is that alternative methods are typically run with only a subset of genes chosen, for example, 500-2000 genes for scDesign2 [@Sun2021], 1000 genes for scDesign3 [@scDesign3], and 621 for SeqNet [@SeqNet]; presumably these limitations are due to the high resource requirements.
+In contrast, our methods can operate on the entire space of genes, avoiding an arbitrary choice of cutoff, though it should be noted that correlation is most meaningful in high-expressed, high-variance genes and so subsetting genes is reasonable.
 
-We demonstrated the importance of including gene-gene dependence in simulated data by two application benchmarks.
+We demonstrated the importance of including gene-gene dependence in simulated data by two benchmarking applications.
 In the first, DESeq2 results were substantially more variable when simulating with gene-gene dependence.
 This indicates that DESeq2 had an increased chance of a larger simultaneous number of false positives, as well as increased chance of fewer false positives, when data included realistic dependence.
-We therefore recommend that benchmarking of differential expression methods should include gene-gene dependence whenever possible.
+We therefore recommend that benchmarking of differential expression methods include gene-gene dependence whenever possible.
 This is despite the fact that differential expression is largely addressing a problem of the behavior of individual genes.
@@
 These observations demonstrate that there is room for future improvements over independent data in these techniques, possibly incorporating more recent developments in copulae [@copulae].
-Possibly, this could demonstrate the limitations of methods based on the multivariate normal distribution or of the low-rank approximation used by all three of our methods. 
+Possibly, this could demonstrate the limitations of methods based on the multivariate normal distribution or of the low-rank approximation used by all three proposed methods. 
 Nonetheless, these methods represent significant improvements by other metrics and we recommend the inclusion of some dependence in nearly every simulated omics dataset.
@@
 The spiked Wishart method also undershoots but less so than the corpcor method and therefore we believe represents the best starting point for users.
-However, we emphasize that in any application, the users should compare their simulated data to the reference data set and this can be used to determine the best method for the specific application at hand, which could vary depending on data set.
+However, we emphasize that in any application, the users should compare their simulated data to the reference dataset and this can be used to determine the best method for the specific application at hand, which could vary depending on data set.
 
@@
 Other Gaussian copula-based R packages include `bindata`, `GenOrd`, and `SimMultiCorrData`, the last of these being the most comprehensive.
-The `bigsimr` package provides faster implementations of these methods to scale up to omics-level data.
+The `bigsimr` package provides faster implementations of these methods which scale up to omics-level data.
 However, even this is computationally demanding; their paper references generating 20,000-dimensional vectors in “under an hour” using 16 threads.
 The `copula` package provides even more flexible dependence options through use of copulas.
-All of these packages provide more flexibility in specifying dependence than our package, which can only mimic existing datasets, and therefore the longer run-times may be unavoidable for use cases where researchers need to parameterize the dependence structure.
+All of these packages provide more flexibility in specifying dependence and therefore the longer run-times may be unavoidable for use cases where researchers need to parameterize the dependence structure.
 
-While we describe one special form of $\Sigma_{sim}$ where generation is highly efficient, there exist many other possible covariance matrix forms.
+While we describe one special form of $\Sigma_{\text{sim}}$ where generation is highly efficient, there exist many other possible covariance matrix forms.
 Block diagonal covariance matrices are one such example which can efficiently represent even high-dimensional data [@Devijver02012018] and lend themselves to efficient simulation.

### 2025-06-17 manuscript/paper.qmd
Fix \text elements

@@
     toc-expand: true
+  pdf:
+    keep-tex: true
+    template-partials: 
+      - title.tex
+    include-in-header:
+      text: |
+        \usepackage[noblocks]{authblk}
+        \renewcommand*{\Authsep}{, }
+        \renewcommand*{\Authand}{, }
+        \renewcommand*{\Authands}{, }
+        \renewcommand\Affilfont{\small}
 bibliography: references.bib
@@
 #| fig-cap: "(a) Schematic of our Gaussian copula-based approach.
-The goal is to pick $\\Sigma_{\text{sim}}$ so that $\\widehat \\Sigma_{\text{sim}} \\approx \\widehat \\Sigma_{\text{ref}}$ while using a special, highly efficient form.
-(b) Eigenvalues of the sample covariance matrix $\\widehat \\Sigma_{\text{ref}}$ of reference data compared to $\\widehat \\Sigma_{\text{sim}}$ from simulated data using $\\Sigma_{\text{sim}} = \\widehat \\Sigma_{\text{ref}}$.
+The goal is to pick $\\Sigma_{\\text{sim}}$ so that $\\widehat \\Sigma_{\\text{sim}} \\approx \\widehat \\Sigma_{\\text{ref}}$ while using a special, highly efficient form.
+(b) Eigenvalues of the sample covariance matrix $\\widehat \\Sigma_{\\text{ref}}$ of reference data compared to $\\widehat \\Sigma_{\\text{sim}}$ from simulated data using $\\Sigma_{\\text{sim}} = \\widehat \\Sigma_{\\text{ref}}$.
 100 genes and 12 samples were used for both reference and the 20 simulated datasets.

### 2025-06-17 manuscript/paper.qmd
Narrower margins for pdf
Finalizing revisions

@@
         \renewcommand\Affilfont{\small}
+    margin-left: 0.75in
+    margin-right: 0.75in
+    margin-top: 0.75in
+    margin-bottom: 0.75in
 bibliography: references.bib
@@
 Simulation of realistic omics data is a key input for benchmarking studies that help users obtain optimal computational pipelines.
-Omics data involves large numbers of measured features on each samples and these measures are generally correlated with each other.
+Omics data involves large numbers of measured features on each sample and these measures are generally correlated with each other.
 However, simulation too often ignores these correlations, perhaps due to computational and statistical hurdles of doing so.
-To alleviate this, we describe three approaches for efficiently generating omics-scale data with correlated measures which mimic real datasets.
+To alleviate this, we describe three approaches for generating omics-scale data with correlated measures which mimic real datasets.
 These approaches are all based on a Gaussian copula approach with a covariance matrix that decomposes into a diagonal part and a low-rank part.
+This decomposition allows for extremely efficient simulation, overcoming a hurdle for adoption of past methods.
 We use these approaches to demonstrate the importance of including correlation in two benchmarking applications.
@@
 We demonstrated these on two applications which show how inclusion of these dependencies can affect the results of benchmarking.
-Lastly, we provide a software package to act as reference implementations of these.
+Lastly, we provide a software package to act as a reference implementations of these.
 
@@
 However, a major challenge when simulating datasets is choice of the copula, which is a mathematical object that encodes the dependence between two or more variables.
-For Guassian copulas, this is a choice a covariance matrix.
-However, the sample covariance matrix of a reference dataset is a poor choice and present three alternative choices.
+For Guassian copulas, this is a choice of a covariance matrix.
+However, the sample covariance matrix of a reference dataset is a poor choice and we present three alternative choices.
 
@@
 Shrinkage methods [@SchaferStrimmer2005; @OpgenRheinStrimmer2007] improve estimates and guarantee that the estimated matrix is well-conditioned (and hence invertible).
-Sparse matrix methods [@Bien2011-nx, @Bickel_sparse, @Cai_adaptive] produce a matrices where many of the pairwise correlations are zero.
+Sparse matrix methods [@Bien2011-nx, @Bickel_sparse, @Cai_adaptive] produce matrices where many of the pairwise correlations are zero.
 Similarly, other methods produce a sparse estimate of the inverse of the covariance matrix [@sparse_inverse_covariance], which imparts pairwise independence of some variables conditional on all the others.
 
-Copula approaches are flexible and have a long history [@Sklar] of use in fields such as economics [@MacKenzie_copula] 
+Copula approaches are flexible and have a long history [@Sklar] of use in fields such as economics [@MacKenzie_copula].
 More recently, copula approaches have been used for modelling or classification in multi-omics fields [@He_copula; @Ma_copula] and metagenomics [@Deek_microbiome].
@@
 No method exactly replicates the reference dataset's properties, indicating room for future research, but all improve upon the common approach of assuming independence.
-We implement this in an R package which supports normal, Poisson, DESeq2-based (negative binomial with sample-specific size factors), and empirical (for ordinal data) marginal distributions.
+We implement this in an R package which supports normal, Poisson, DESeq2-based (negative binomial with sample-specific size factors [@Love2014]), and empirical (for ordinal data) marginal distributions.
 
@@
 We start with a reference dataset $X$ given by a $p \times n$ data matrix of $p$ features measured in each of $n$ independent samples, which could be any real dataset of interest.
-Our goal is to be able to generate simulated datasets that look the same as this reference dataset, on average and when $n$ samples were simulated.
+Our goal is to be able to generate simulated datasets that look the same as this reference dataset, on average and when $n$ samples are simulated.
 To do this, we need to capture correlations between the $p$ features, which could represent gene expressions, protein abundances, or other measured values.
@@
 However, using $\widehat \Sigma_{\text{ref}}$ is a poor choice for $\Sigma_{\text{sim}}$.
-Indeed, we consider a choice of $\Sigma_{\text{sim}}$ to be successful if datasets of $n$ samples simulated using $\Sigma_{\text{sim}}$ have sample covaraince matrices $\widehat \Sigma_{\text{sim}}$ that are similar to $\widehat \Sigma_{\text{ref}}$.
+Indeed, we consider a choice of $\Sigma_{\text{sim}}$ to be successful if datasets of $n$ samples simulated using $\Sigma_{\text{sim}}$ have sample covariance matrices $\widehat \Sigma_{\text{sim}}$ that are similar to $\widehat \Sigma_{\text{ref}}$.
 Counter-intuitively, data simulated using $\Sigma_{\text{sim}} := \widehat \Sigma_{\text{ref}}$ have sample covariance matrices that systematically and substantially differ from $\widehat \Sigma_{\text{ref}}$, see @fig-intro b.
@@
 The goal is to pick $\\Sigma_{\\text{sim}}$ so that $\\widehat \\Sigma_{\\text{sim}} \\approx \\widehat \\Sigma_{\\text{ref}}$ while using a special, highly efficient form.
-(b) Eigenvalues of the sample covariance matrix $\\widehat \\Sigma_{\\text{ref}}$ of reference data compared to $\\widehat \\Sigma_{\\text{sim}}$ from simulated data using $\\Sigma_{\\text{sim}} = \\widehat \\Sigma_{\\text{ref}}$.
+(b) Eigenvalues of the sample covariance matrix $\\widehat \\Sigma_{\\text{ref}}$ of reference data compared to $\\widehat \\Sigma_{\\text{sim}}$ from simulated data using $\\Sigma_{\\text{sim}} := \\widehat \\Sigma_{\\text{ref}}$.
 100 genes and 12 samples were used for both reference and the 20 simulated datasets.
@@
 For example, RNA-seq involves a random sampling of fragments from the true sample of RNA molecules and other omics modes involve various background noise effects.
-We therefore need to "inject" some level of additional randomness that is independent between genes on top of the correlations captured by $\widehat \Sigma_{\text{ref}}$
+We therefore need to increase the amount of independence beyond what $\widehat \Sigma_{\text{ref}}$ has, a process we refer to as "injecting" independence.
 
@@
 where $D$ is a diagonal matrix capturing the injected independence and $UW^2U^T$ is a low-rank correlation matrix derived from $\widehat \Sigma_{\text{ref}}$.
-An additional advantage of this special form is that data may be generated very efficiently, @fig-intro b,c, compared to arbitrary $\Sigma_{\text{sim}}$ which typically require a Cholesky or eigenvalue decomposition.
+An additional advantage of this special form is that data may be generated very efficiently (@fig-intro b,c) compared to using arbitrary $\Sigma_{\text{sim}}$ which typically require a Cholesky or eigenvalue decomposition.
 Specifically, random samples are generated using the fact that if $u,v$ are independent vectors with iid standard normal entries, then $Dv + UWu$ has covariance matrix equal to $D^2 + UW^2U^T$, see Methods.
@@
 
-To compare the three simulation methods with a real dataset, we chose as a references dataset 12 mouse cortex RNA-seq samples from accession GSE151923 [@Wang2022].
+To compare the three simulation methods with a real dataset, we chose as a reference dataset 12 mouse cortex RNA-seq samples from accession GSE151923 [@Wang2022].
 We then simulated data mimicking this reference using all three simulation methods (PCA, spiked Wishart, and corpcor) as well as a simulation with independent genes.
@@
 The simulation with independence showed the least levels of gene-gene correlations (@fig-compare-to-real d).
-However, the PCA method overshot the reference dataset and the spiked Wishart and corpcor methods only slightly improved upon the independent simulation.
+However, the PCA method overshot the reference dataset and the spiked Wishart and corpcor methods modestly improved upon the independent simulation.
 The SPsimSeq simulator performed similarly to the PCA method.
@@
 
-We further demonstrated our methods on metabolomics data, where we use normal marginal distributions and with just 42 features measured, see Supplemental Figure 3.
+We further demonstrated our methods on metabolomics data, where we use normal marginal distributions and with just 42 features measured, see Supplemental Figure 1.
 
@@
 
-We benchmarked DESeq2 [@Love2014], a popular differential expression analysis tool, using datasets simulated with dependence and ones simulated without dependence to compare its performances on both.
+We benchmarked DESeq2 [@Love2014], a popular differential expression analysis tool, using datasets simulated with dependence and ones simulated without dependence to compare its performance on both.
 DESeq2 presents an interesting case because several aspects of it assume independence of genes and so may be adversely affected by gene-gene dependence.
 First, the independent filtering step [@Bourgon2010-uv] assumes independence but has been reported to be robust to typical gene-gene dependence.
-Relatedly, the false discovery rate (FDR) [@BH_FDR] allows only certain forms of dependence.
+Relatedly, the false discovery rate (FDR) [@BH_FDR] allows only certain forms of dependence [@Benjamini2001-na].
 Lastly, DESeq2's empirical Bayes steps could possibly be affected by gene dependence.
 
-We used a fly whole body RNA-Seq dataset GSE81142, see Supplemental Figure 1.
+We used a fly whole body RNA-seq dataset GSE81142, see Supplemental Figure 2.
 To select a "control" dataset samples were selected if they met the following criteria: male flies; without treatment; and after at least 2 hours of feeding.
@@
 We next used our simulation method to benchmark CYCLOPS [@Anafi2017], which infers relative times for a set of unlabeled samples using an autoencoder to identify circular structures.
-We chose a mouse cortex time series dataset GSE151565, see Supplemental Figure 2, which contains a total of 77 samples every 3 hours, for 36 hours.
+We chose a mouse cortex time series dataset GSE151565, see Supplemental Figure 3, which contains a total of 77 samples every 3 hours, for 36 hours.
 As before, unnormalized read counts were used as the reference dataset and all simulations used the DESeq2-based negative binomial model.
@@
 Therefore, given matrices $D, U, W$ and $u \sim N(0, I_k)$ and $v \sim N(0, I_p)$ (where $I_k$ and $I_p$ are the $k \times k$ and $p \times p$ identity matrices), then
-$$ D v + U W u \sim N(0, D^2 + UW^2 U^T) = N(0, \Sigma) $$
+$$ D v + U W u \sim N(0, D^2 + UW^2 U^T) = N(0, \Sigma_{\text{sim}}) $$
 as desired.
 
-In contrast, when given an arbitrary $\Sigma$, to generate a random value in $N(0, \Sigma)$, one must compute a matrix $V$ such that $VV^T = \Sigma$.
-Then, using $v \sim N(0,I)$ , we have $Vv \sim N(0, \Sigma)$ as desired.
-However, computing $V$ is done using a Cholesky decomposition or eigenvector decomposition, both of which are computationally expensive when $\Sigma$ is large.
+In contrast, when given an arbitrary $\Sigma_{\text{sim}}$, to generate a random value in $N(0, \Sigma_{\text{sim}})$, one must compute a matrix $V$ such that $VV^T = \Sigma_{\text{sim}}$.
+Then, using $v \sim N(0,I)$ , we have $Vv \sim N(0, \Sigma_{\text{sim}})$ as desired.
+However, computing $V$ is done using a Cholesky decomposition or eigenvector decomposition, both of which are computationally expensive when $\Sigma_{\text{sim}}$ is large.
 
@@
 Finally, those are fed to a standard normal distribution's inverse CDF to obtain values that are approximately normally distributed.
-These values are then used to compute a covariance matrix $\Sigma$ and the data is assumed to follow a multivariate normal distribution in $p$ dimensions with that covariance matrix.
+These values are then used to compute a covariance matrix $\Sigma_{\text{sim}}$ and the data is assumed to follow a multivariate normal distribution in $p$ dimensions with that covariance matrix.
 
-Here, we describe the approach using the form of covariance matrix $\Sigma = D^2 + UW^2U^T$ as above.
-Once data is obtained $Z \sim N(0, \Sigma)$, then one can undo the normalizing transformation to obtain data with the same marginal distributions as the fit marginal distributions but with dependence determined by $\Sigma$.
+Here, we describe the approach using the form of covariance matrix $\Sigma_{\text{sim}} = D^2 + UW^2U^T$ as above.
+Once data is obtained $Z \sim N(0, \Sigma_{\text{sim}})$, then one can undo the normalizing transformation to obtain data with the same marginal distributions as the fit marginal distributions but with dependence determined by $\Sigma_{\text{sim}}$.
 We describe this in detail:
@@
 2.  Apply normalizing transform to $X$ by setting $Z_{ij} = \Phi^{-1}(F_{i}(X_{ij}))$ where $\Phi$ is the CDF of the standard normal distribution.
-3.  Compute $D$, $U$, $W$ matrices from $X$ by one of three methods (see Methods).
+3.  Compute $D$, $U$, $W$ matrices from $X$ by one of three methods (described later).
 4.  Generate $k$ i.i.d. standard normally distributed values $u$ and $p$ i.i.d standard normally distributed values $v$.
@@
 
-Below, we describe the three methods for selecting the components of the covariance matrix $\Sigma_{\text{sim}} := D^2 + U W^2 U^T$.
+Below, we describe the three methods for selecting the components $D, U, W$ of the covariance matrix.
 
@@
 
-Here we have described the well-known Gaussian copula approach and argued that a specific form of covariance matrix which is well tailored to omics data simulation.
+Here we have described the well-known Gaussian copula approach and argued that a specific form of covariance matrix is well tailored to omics data simulation.
 We developed three methods using this form of covariance matrix which can be used to mimic a reference dataset for simulation.
@@
 This implementation is efficient and simple, requiring just two lines of code to fit a model to a reference dataset and then simulate data from it.
-We demonstrated this package on RNA-seq data, using the DESeq2 method to fit negative binomial marginal distributions as well as a metabolomics dataset using normal marginal distributions.
+We demonstrated this package on RNA-seq data, using the DESeq2 method to fit negative binomial marginal distributions, as well as a metabolomics dataset using normal marginal distributions.
 Our package is even more general and supports normal, Poisson, negative binomial, and arbitrary ordered discrete distributions using the empirical CDF.
@@
 
-We demonstrated that our methods are highly efficient and easily scale to the large feature counts typical of omics data sets.
+We demonstrated that our methods are highly efficient and easily scale to the large feature counts typical of omics datasets.
 While runtime of simulators is typically not critical, we posit that it introduces a sufficient inconvenience to have slowed the adoption of dependence in omics simulation and could be particularly cumbersome whenever large numbers of simulated datasets are needed.
-Moreover, high memory use of the alternatives could be an even larger barrier than run-time.
+Moreover, high memory use could be an even larger barrier than run-time, and is also dramatically improved by our methods.
 Of note is that alternative methods are typically run with only a subset of genes chosen, for example, 500-2000 genes for scDesign2 [@Sun2021], 1000 genes for scDesign3 [@scDesign3], and 621 for SeqNet [@SeqNet]; presumably these limitations are due to the high resource requirements.
@@
 
-In the second application, CYCLOPS performance in estimating circadian phases depends upon gene-gene dependence, was sensitive to dependence structure of the data.
+In the second application, CYCLOPS performance in estimating circadian phases was sensitive to dependence structure of the data under some configurations.
 Unlike DESeq2, CYCLOPS explicitly makes use of the correlations of genes and therefore this result is not surprising but still demonstrates the potential impact of the assumption of independence when benchmarking.
@@
 These observations demonstrate that there is room for future improvements over independent data in these techniques, possibly incorporating more recent developments in copulae [@copulae].
-Possibly, this could demonstrate the limitations of methods based on the multivariate normal distribution or of the low-rank approximation used by all three proposed methods. 
+This could demonstrate the limitations of methods based on the multivariate normal distribution or of the low-rank approximation used by all three proposed methods. 
 Nonetheless, these methods represent significant improvements by other metrics and we recommend the inclusion of some dependence in nearly every simulated omics dataset.
@@
 
-Other Gaussian copula-based R packages include `bindata`, `GenOrd`, and `SimMultiCorrData`, the last of these being the most comprehensive.
+Other Gaussian copula-based R packages include `copula`, `bindata`, `GenOrd`, and `SimMultiCorrData`, the last of these being the most comprehensive.
 The `bigsimr` package provides faster implementations of these methods which scale up to omics-level data.
 However, even this is computationally demanding; their paper references generating 20,000-dimensional vectors in “under an hour” using 16 threads.
-The `copula` package provides even more flexible dependence options through use of copulas.
 All of these packages provide more flexibility in specifying dependence and therefore the longer run-times may be unavoidable for use cases where researchers need to parameterize the dependence structure.

### 2025-06-17 manuscript/supplemental/supplemental.qmd
Finalizing revisions

--- 2025-05-23 manuscript/supplemental/supplemental.qmd
@@
 ```{r}
-#| label: fig-compare-to-real
+#| label: fig-compare-to-real-metabolomics
+#| fig-cap: "Comparison to plasma metabolomics data set, obtained from MetaboAnalyst (https://www.metaboanalyst.ca/MetaboAnalyst/upload/RocUploadView.xhtml).
+Methods were run using normal marginal distributions.
+Unlike other examples, this included more samples (59) than measured features (42 metabolites), demonstrating that these methods also work in that case.
+Data were log-scaled prior to simulation."
+#| fig-width: 10
+#| fig-height: 10
+readRDS("../../processed/compare_to_real_plot_metabolomics.RDS")
+
+```{r}
+#| label: fig-compare-to-real-fly
 #| fig-cap: "Comparison to real data run on a fly whole body data set from GSE81142.
@@
 ```{r}
-#| label: fig-compare-to-real2
+#| label: fig-compare-to-real-mouse
 #| fig-cap: "Comparison to real data run on a mouse cortex timeseries body data set from GSE151565.
@@
 readRDS("../../processed/compare_to_real_plot.GSE151565.RDS")
+```
 ```

### 2025-06-18 manuscript/paper.qmd
Add a clarification of normalization
Revert to original 2% cutoff
Mistakenly had left on 1% cutoff for CYCLOPS

@@
 We then simulated data mimicking this reference using all three simulation methods (PCA, spiked Wishart, and corpcor) as well as a simulation with independent genes.
-All data were simulated with negative binomial distributions fit by DESeq2.
+All data were simulated with negative binomial distributions fit by DESeq2 on unnormalized read counts.
 We repeated the simulations, each of 12 samples, a total of 8 times to estimate variance.
@@
 We found that CYCLOPS accuracy depended significantly on this parameter, with the default producing good accuracy across all simulations.
-However, when dropping CYCLOPS to require just 1% variance in each eigengene, we found that its accuracy depends significantly on the dependence structure of the simulated time series data (@fig-CYCLOPS a).
+However, when dropping CYCLOPS to require just 2% variance in each eigengene, we found that its accuracy depends significantly on the dependence structure of the simulated time series data (@fig-CYCLOPS a).
 At that setting, CYCLOPS accuracy is much higher in the PCA method and moderately improved in Wishart method, compared to the 'independent' method. 
@@
 #| label: fig-CYCLOPS
-#| fig-cap: "Performance of CYCLOPS on simulated time series datasets based on mouse cortex dataset (GSE151565), using eigengenes of at least 1% variance.
+#| fig-cap: "Performance of CYCLOPS on simulated time series datasets based on mouse cortex dataset (GSE151565), using eigengenes of at least 2% variance.
 (a) Absolute circular correlations between true phases and CYCLOPS estimated phases on the simulated datasets.

### 2025-06-18 manuscript/paper.qmd
Revise CYCLOPS comment

@@
 This difference is likely driven by the difference in the number of eigengenes used (@fig-CYCLOPS b), which is a measure of how much dependence is present in the dataset.
+Notably, the number of eigengenes and accuracy tracked both with each other and with the amount of gene-gene correlation (see Supplemental Figure 3 d) across the methods.
 This demonstrates that the correlation structure of the transcriptome can have a major impact on accuracy.
-In this application, the differences between corpcor and Wishart were such that, while Wishart improved upon the number of eigengenes used compared to corpcor, both undershot the level of correlation in the real data set (see Supplemental Figure 2 d) and the difference between them was too small to affect the actual accuracy of CYCLOPS as measured by the circular correlation.
 
@@
 #| label: fig-CYCLOPS
-#| fig-cap: "Performance of CYCLOPS on simulated time series datasets based on mouse cortex dataset (GSE151565), using eigengenes of at least 2% variance.
+#| fig-cap: "Accuracy of CYCLOPS on simulated time series datasets based on mouse cortex dataset (GSE151565), using eigengenes of at least 2% variance.
 (a) Absolute circular correlations between true phases and CYCLOPS estimated phases on the simulated datasets.

### 2025-06-18 manuscript/supplemental/supplemental.qmd
Adjust figure positions for PDF generation

@@
     toc-expand: true
+  pdf:
+    include-in-header:
+      - text: \usepackage{placeins}
 bibliography: ../references.bib
 ---
-
 ## Supplemental Material
@@
 ### Comparisons to real data
+Below, we give three comparisons of simulated data with their real reference data.
 
@@
 Data were log-scaled prior to simulation."
-#| fig-width: 10
-#| fig-height: 10
+#| fig-width: 8 
+#| fig-height: 8
 readRDS("../../processed/compare_to_real_plot_metabolomics.RDS")
+```
 
@@
 Unlike (d), these components were fit from each data set considered separately instead of reusing the weights from the real data."
-#| fig-width: 10
-#| fig-height: 10
+#| fig-width: 8
+#| fig-height: 8
 readRDS("../../processed/compare_to_real_plot.GSE81142.RDS")
@@
 Unlike (d), these components were fit from each data set considered separately instead of reusing the weights from the real data."
-#| fig-width: 10
-#| fig-height: 10
+#| fig-width: 8
+#| fig-height: 8
 readRDS("../../processed/compare_to_real_plot.GSE151565.RDS")
 ```
-```
-
+\FloatBarrier
 ## Supplemental Methods

### 2025-06-20 manuscript/paper.qmd
Revise according to editorial requests

@@
       affiliations:
-        - name: "Chronobiology and Sleep Institute, University of Pennsylvania"
+        - name: "Chronobiology and Sleep Institute, University of Pennsylvania, Philadelphia, Pennsylvania, United States of America"
     - name: "Gregory R. Grant"
@@
       affiliations:
-        - name: "Department of Genetics, University of Pennsylvania"
+        - name: "Department of Genetics, University of Pennsylvania, Philadelphia, Pennsylvania, United States of America"
     - name: "Thomas G. Brooks"
@@
       affiliations:
-        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania"
+        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania, Philadelphia, Pennsylvania, United States of America"
 editor: 
@@
 Points are colored according to the density of points in their region.
-Wishart and corpcor methods give similar results (not shown).
 (c) Quantile-quantile plot comparing correlation values of gene pairs from real data and simulated data (both with and without dependence).
@@
 All sequencing data used is available from the Gene Expression Omnibus (GEO) with accession numbers GSE151923, GSE81142, and GSE151565.
-Metabolomics data was obtained from MetaboAnalyst (https://www.metaboanalyst.ca/MetaboAnalyst/upload/RocUploadView.xhtml).
+Metabolomics data was obtained from MetaboAnalyst (https://api2.xialab.ca/api/download/metaboanalyst/plasma_nmr.csv).
 
@@
 
-The authors declare no competing interests.
+The authors have declared that no competing interests exist.
+
+## References
+
+::: {#refs}
+:::
+
+## Supporting Information
+
+### Supporting Information 1
+
+Supporting document containing all Supplemental Figures 1-3 and Supplemental Methods.

### 2025-08-05 manuscript/paper.qmd
Second review revisions

@@
 However, a major challenge when simulating datasets is choice of the copula, which is a mathematical object that encodes the dependence between two or more variables.
-For Guassian copulas, this is a choice of a covariance matrix.
+For Gaussian copulas, this is a choice of a covariance matrix.
 However, the sample covariance matrix of a reference dataset is a poor choice and we present three alternative choices.
@@
 Eigenvalues are numbered from largest to smallest.
-(c-d) Comparison of (c) run time and (d) memory usage for our methods (pca, wishart, coprcor), SPsimSeq, multivariate normal sampling (using the mvrnorm function from the MASS package), and vine copula (implemented by the rvinecopulib R package).
+(c-d) Comparison of (c) run time and (d) memory usage for our methods (pca, wishart, corpcor), SPsimSeq, mvrnorm (from the MASS package), the mvnfast package, and vine copula (implemented by the rvinecopulib R package).
 100 samples were generated with a varying number of genes.
 A mouse cortex dataset GSE151564 was randomly subsetted to have the required number of genes to be used as the reference data set.
-Vine copula method was only run up to 2000 genes due to the memory use required at higher gene counts."
+Vine copula method was only run up to 2000 genes due to the memory use required at higher gene counts.
+All methods used a single thread."
 #| fig-width: 8
@@
 # Plot benchmarking
-METHOD_ORDER <- c("real", "indep", "pca", "wishart", "corpcor", "SPsimSeq", "vinecopula", "mvrnorm")
+METHOD_ORDER <- c("real", "indep", "pca", "wishart", "corpcor", "SPsimSeq", "vinecopula", "mvrnorm", "mvnfast")
 METHOD_COLORS <- c("black", RColorBrewer::brewer.pal(length(METHOD_ORDER)-1, "Dark2"))
@@
 
-We note that, unlike many timeseries methods that consider correlation across time points, the dependence here is between genes within the same sample and no dependence between samples or timepoints is generated.
+We note that, unlike many timeseries methods that consider correlation across time points, the dependence here is between genes within the same sample and generated samples are independent, conditional upon the time point.
 This reflects scenarios where CYCLOPS is used, for example, post-mortem biopsy samples of unrelated individuals collected at different times [@CYCLOPS2].
@@
 Generating data for $\Sigma_{\text{sim}}$ of this form is highly efficient.
-We use two basic facts about the multivariate distribution.
+We use two basic facts about the multivariate normal distribution.
 First, if $A$ is a matrix and $u \sim N(0, \Sigma)$ then $Au \sim N(0, A \Sigma A^T)$.
-Second, if $u \sim N(0, \Sigma_1)$ and $v \sim N(0, \Sigma_2)$, then $u + v \sim N(0, \Sigma_1 + \Sigma_2)$.
-Therefore, given matrices $D, U, W$ and $u \sim N(0, I_k)$ and $v \sim N(0, I_p)$ (where $I_k$ and $I_p$ are the $k \times k$ and $p \times p$ identity matrices), then
+Second, if $u \sim N(0, \Sigma_1)$ and $v \sim N(0, \Sigma_2)$ are independent, then $u + v \sim N(0, \Sigma_1 + \Sigma_2)$.
+Therefore, given matrices $D, U, W$ and independent random vectors $u \sim N(0, I_k)$ and $v \sim N(0, I_p)$ (where $I_k$ and $I_p$ are the $k \times k$ and $p \times p$ identity matrices), then
 $$ D v + U W u \sim N(0, D^2 + UW^2 U^T) = N(0, \Sigma_{\text{sim}}) $$
@@
 
-We compared our simulated data to that of the R package SPsimSeq [@Assefa2020].
+We compared our simulated data to that of the R package SPsimSeq [@Assefa2020], due to its popularity, support for bulk RNA-Seq, and use of Gaussian copulas.
 In contrast to our package, SPsimSeq uses a two-step randomization.
@@
 All of these packages provide more flexibility in specifying dependence and therefore the longer run-times may be unavoidable for use cases where researchers need to parameterize the dependence structure.
+Another package that optimizes the speed of random multivariate normal vector generation is the `mvnfast` package for R [@mvnfast], however, we show that this does not provide a meaningful speedup for our use-cases due to the reliance on the Cholesky decomposition, see @fig-intro d.
 
@@
 All sequencing data used is available from the Gene Expression Omnibus (GEO) with accession numbers GSE151923, GSE81142, and GSE151565.
-Metabolomics data was obtained from MetaboAnalyst (https://api2.xialab.ca/api/download/metaboanalyst/plasma_nmr.csv).
+Metabolomics data was obtained from MetaboAnalyst ([https://api2.xialab.ca/api/download/metaboanalyst/plasma_nmr.csv](https://api2.xialab.ca/api/download/metaboanalyst/plasma_nmr.csv)).
 

### 2025-08-05 scripts/mvnfast.benchmark.R

@@ -1,0 +1,34 @@
+library(mvnfast)
+library(tidyverse)
+library(rlang)
+
+method <- "mvnfast"
+N_GENES <- snakemake@wildcards$n_genes
+N_SAMPLES <- 100
+
+# use ZT0 of raw time series data to get dependence structure
+male_read_counts <- read.csv("processed/Cortex_ZT0-counts.csv") |> 
+    column_to_rownames(var="raw_data.EnsemblID")
+
+selected_rows <- sample.int(nrow(male_read_counts), N_GENES)
+input <- male_read_counts[selected_rows,]
+
+## Generate Controls
+# With dependence
+time <- system.time({
+    # Note: add a small amount for positive definiteness
+    Sigma <- cov(input |> t()) + diag(N_GENES)*0.01
+    data <- rmvn(
+        n = N_SAMPLES,
+        mu = rowMeans(input),
+        sigma = Sigma,
+    )
+})
+
+message("Elapsed time:")
+message(time[['elapsed']])
+tibble(
+    method = method,
+    n_genes = N_GENES,
+    time = time[['elapsed']],
+) |> write_tsv(paste0("processed/benchmark/", method, "/", N_GENES, ".txt"))

### 2025-08-05 manuscript/paper.qmd
Tweak wording

@@
 All of these packages provide more flexibility in specifying dependence and therefore the longer run-times may be unavoidable for use cases where researchers need to parameterize the dependence structure.
-Another package that optimizes the speed of random multivariate normal vector generation is the `mvnfast` package for R [@mvnfast], however, we show that this does not provide a meaningful speedup for our use-cases due to the reliance on the Cholesky decomposition, see @fig-intro d.
+Another package that optimizes the speed of random multivariate normal vector generation is the `mvnfast` package for R [@mvnfast], however, we show that this does not provide a meaningful speedup for our use-cases, likely due to the reliance on the Cholesky decomposition, see @fig-intro d.
 

### 2025-08-05 manuscript/paper.qmd
Clarify future directions section

@@
 
-While we describe one special form of $\Sigma_{\text{sim}}$ where generation is highly efficient, there exist many other possible covariance matrix forms.
+While we describe one special form of $\Sigma_{\text{sim}}$ where generation is highly efficient, there exist many other possible covariance matrix forms which are promising directions for future work.
 Block diagonal covariance matrices are one such example which can efficiently represent even high-dimensional data [@Devijver02012018] and lend themselves to efficient simulation.
-Sparse matrix methods [@Bien2011-nx; @Bickel_sparse; @Cai_adaptive] yield exactly zero correlation between some variables, but it is less clear how to simulate data from these efficiently.
+Other approaches include sparse matrix methods [@Bien2011-nx; @Bickel_sparse; @Cai_adaptive] which yield exactly zero correlation between some variables, but it is less clear how to simulate data from these efficiently.
 Iterative sampling methods [@Krylov_subspaces] are one approach that could exploit the sparse nature and may even extend to sparse inverse covariance methods [@sparse_inverse_covariance].

### 2025-08-07 manuscript/paper.qmd
revisions for editorial and reviewer comments

@@
   warning: false
+crossref:
+    fig-prefix: Fig
 format:
@@
 Our methods are based off the well-known Gaussian copula method, which uses a multivariate normal distribution to capture aspects of the gene-gene dependence, without necessarily assuming that the data is normally distributed.
-See @fig-intro a for a high-level overview.
+See [@fig-intro]A for a high-level overview.
 This multivariate normal distribution is parameterized by a $p \times p$ covariance matrix, $\Sigma$.
@@
 Indeed, we consider a choice of $\Sigma_{\text{sim}}$ to be successful if datasets of $n$ samples simulated using $\Sigma_{\text{sim}}$ have sample covariance matrices $\widehat \Sigma_{\text{sim}}$ that are similar to $\widehat \Sigma_{\text{ref}}$.
-Counter-intuitively, data simulated using $\Sigma_{\text{sim}} := \widehat \Sigma_{\text{ref}}$ have sample covariance matrices that systematically and substantially differ from $\widehat \Sigma_{\text{ref}}$, see @fig-intro b.
+Counter-intuitively, data simulated using $\Sigma_{\text{sim}} := \widehat \Sigma_{\text{ref}}$ have sample covariance matrices that systematically and substantially differ from $\widehat \Sigma_{\text{ref}}$, see [@fig-intro]B.
 
@@
 where $D$ is a diagonal matrix capturing the injected independence and $UW^2U^T$ is a low-rank correlation matrix derived from $\widehat \Sigma_{\text{ref}}$.
-An additional advantage of this special form is that data may be generated very efficiently (@fig-intro b,c) compared to using arbitrary $\Sigma_{\text{sim}}$ which typically require a Cholesky or eigenvalue decomposition.
+An additional advantage of this special form is that data may be generated very efficiently ([Figs @fig-intro]B and [-@fig-intro]C) compared to using arbitrary $\Sigma_{\text{sim}}$ which typically require a Cholesky or eigenvalue decomposition.
 Specifically, random samples are generated using the fact that if $u,v$ are independent vectors with iid standard normal entries, then $Dv + UWu$ has covariance matrix equal to $D^2 + UW^2U^T$, see Methods.
@@
 
-For comparison we also ran SPsimSeq, another RNA-seq simulator for both bulk and single-cell that uses a Gaussian copula-based model of dependence, see Supplemental Methods for details.
+For comparison we also ran SPsimSeq, another RNA-seq simulator for both bulk and single-cell that uses a Gaussian copula-based model of dependence, see S1 Text for details.
 
-Simulated data captures the genes' mean and variance accurately (@fig-compare-to-real a-b).
-Next, we compared to the real dataset when projected onto the top two principal components of the real dataset (@fig-compare-to-real c).
+Simulated data captures the genes' mean and variance accurately ([Figs @fig-compare-to-real]A and [-@fig-compare-to-real]B).
+Next, we compared to the real dataset when projected onto the top two principal components of the real dataset ([@fig-compare-to-real]C).
 The simulations with dependence are distributed around the entire space like the real data, but the independent simulations have unrealistically low variance in these components, clustering tightly around the origin.
@@
 Then, we computed the gene-gene correlation on pairs of high-expressed genes (at least 100 mean reads in the real dataset).
-The simulation with independence showed the least levels of gene-gene correlations (@fig-compare-to-real d).
+The simulation with independence showed the least levels of gene-gene correlations ([@fig-compare-to-real]D).
 However, the PCA method overshot the reference dataset and the spiked Wishart and corpcor methods modestly improved upon the independent simulation.
@@
 
-Lastly, we compared the variances of principal components on each dataset (@fig-compare-to-real e).
-These were computed separately for each dataset, unlike (@fig-compare-to-real c) which used the reference dataset's PCA weights for all datasets.
+Lastly, we compared the variances of principal components on each dataset ([@fig-compare-to-real]E).
+These were computed separately for each dataset, unlike ([@fig-compare-to-real]C) which used the reference dataset's PCA weights for all datasets.
 The independent data has much lower variance than the real dataset in the top four principal components.
@@
 
-We further demonstrated our methods on metabolomics data, where we use normal marginal distributions and with just 42 features measured, see Supplemental Figure 1.
+We further demonstrated our methods on metabolomics data, where we use normal marginal distributions and with just 42 features measured, see S1 Fig.
 
@@
 
-We used a fly whole body RNA-seq dataset GSE81142, see Supplemental Figure 2.
+We used a fly whole body RNA-seq dataset GSE81142, see S2 Fig.
 To select a "control" dataset samples were selected if they met the following criteria: male flies; without treatment; and after at least 2 hours of feeding.
@@
 
-Finally, we ran DESeq2 on each simulated 5 vs 5 experiment and compared the output FDR with the true percentages of genes that are differential expressed (@fig-DESeq2 a-d).
+Finally, we ran DESeq2 on each simulated 5 vs 5 experiment and compared the output FDR with the true percentages of genes that are differential expressed ([Figs @fig-DESeq2]A-[-@fig-DESeq2]D).
 We observed that DESeq2 is anti-conservative on all datasets, with similar mean true FDRs for each estimated FDR cutoff.
@@
 To demonstrate the application of our simulation method for another organism, we also simulated datasets using the mouse cortex dataset GSE151923 [@Wang2022] and selected samples from male mice, as in @fig-compare-to-real.
-We then simulated differential expression experiments as above and observed a similar result (@fig-DESeq2 e,f) as for the fly whole body datasets.
+We then simulated differential expression experiments as above and observed a similar result ([Figs @fig-DESeq2]E and [-@fig-DESeq2]F) as for the fly whole body datasets.
 
@@
 We next used our simulation method to benchmark CYCLOPS [@Anafi2017], which infers relative times for a set of unlabeled samples using an autoencoder to identify circular structures.
-We chose a mouse cortex time series dataset GSE151565, see Supplemental Figure 3, which contains a total of 77 samples every 3 hours, for 36 hours.
+We chose a mouse cortex time series dataset GSE151565, see S3 Fig, which contains a total of 77 samples every 3 hours, for 36 hours.
 As before, unnormalized read counts were used as the reference dataset and all simulations used the DESeq2-based negative binomial model.
@@
 We found that CYCLOPS accuracy depended significantly on this parameter, with the default producing good accuracy across all simulations.
-However, when dropping CYCLOPS to require just 2% variance in each eigengene, we found that its accuracy depends significantly on the dependence structure of the simulated time series data (@fig-CYCLOPS a).
+However, when dropping CYCLOPS to require just 2% variance in each eigengene, we found that its accuracy depends significantly on the dependence structure of the simulated time series data ([@fig-CYCLOPS]A).
 At that setting, CYCLOPS accuracy is much higher in the PCA method and moderately improved in Wishart method, compared to the 'independent' method. 
-This difference is likely driven by the difference in the number of eigengenes used (@fig-CYCLOPS b), which is a measure of how much dependence is present in the dataset.
-Notably, the number of eigengenes and accuracy tracked both with each other and with the amount of gene-gene correlation (see Supplemental Figure 3 d) across the methods.
+This difference is likely driven by the difference in the number of eigengenes used ([@fig-CYCLOPS]B), which is a measure of how much dependence is present in the dataset.
+Notably, the number of eigengenes and accuracy tracked both with each other and with the amount of gene-gene correlation (see S3D Fig) across the methods.
 This demonstrates that the correlation structure of the transcriptome can have a major impact on accuracy.
@@
 All of these packages provide more flexibility in specifying dependence and therefore the longer run-times may be unavoidable for use cases where researchers need to parameterize the dependence structure.
-Another package that optimizes the speed of random multivariate normal vector generation is the `mvnfast` package for R [@mvnfast], however, we show that this does not provide a meaningful speedup for our use-cases, likely due to the reliance on the Cholesky decomposition, see @fig-intro d.
+Another package that optimizes the speed of random multivariate normal vector generation is the `mvnfast` package for R [@mvnfast], however, we show that this does not provide a meaningful speedup for our use-cases, likely due to the reliance on the Cholesky decomposition, see [@fig-intro]C.
 
@@
 
-### Supporting Information 1
+### S1 Fig
 
-Supporting document containing all Supplemental Figures 1-3 and Supplemental Methods.
+Evaluation of our simulation on a metabolomics dataset.
+
+### S2 Fig
+
+Evaluation of our simulation on a fly whole-body RNA-seq dataset.
+
+### S3 Fig
+
+Evaluation of our simulation on a mouse cortex RNA-seq dataset.
+
+### S1 Text
+
+Supplemental methods.

### 2025-08-07 manuscript/supplemental/supplemental.qmd
Revise according to editorial requests
revisions for editorial and reviewer comments

@@
 We set means to be the mean expression of each gene and used the sample covariance matrix as the `Sigma` parameter.
+
+### Running mvnfast
+
+The `mvnfast` R package provides a faster implementation of the multivariate normal distribution random vector generator function, called `rmvn`. 
+As for `mvrnorm`, we set means to be the mean expression of each gene and used the sample covariance matrix as the `Sigma` parameter.
+One core was used for a fair comparison to the other methods, though `mvnfast` does support multithreaded execution.
+Note that `mvnfast` has an option for using a provided pre-computed Cholesky decompositoin of the covariance matrix instead of producing it itself.
+Since Cholesky decomposition is the primary time-sink, using this option could likely considerably speed up the execution time if such a decomposition is known.
+We did not use this option since the comparison is for a generic covariance matrix, where the Cholesky decomposition is not known a priori.
+This contrasts to our implementation which avoids Cholesky decomposition by using only covariance matrices of a specific form.
